---
title: "CODIS Marker Allele Frequencies"
author: "Tina Lasisi"
date: "`r format(Sys.time(), '%Y-%m-%d %H:%M:%S')`"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## CODIS marker allele frequencies

Frequencies and raw genotypes for different populations were found [here](https://www.promega.com/products/pm/genetic-identity/population-statistics/allele-frequencies/) and refer to Steffen, C.R., Coble, M.D., Gettings, K.B., Vallone, P.M. (2017) Corrigendum to 'U.S. Population Data for 29 Autosomal STR Loci' [Forensic Sci. Int. Genet. 7 (2013) e82-e83]. Forensic Sci. Int. Genet. 31, e36–e40. The US core CODIS markers are a subset of the 29 described here. 

```{r setup, include=FALSE}
# Load required libraries
library(readxl)
library(tidyverse)

knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```


```{r convertxl, eval=FALSE, include=FALSE}

# This chunk converts the Excel file that was downloaded into separate CSVs as there is a bunch of information that is spread across different sheets within the original Excel file.

# Read all sheets
file_path <- "data/1036_allelefreqs.xlsx"
sheet_names <- excel_sheets(file_path)

# Iterate over each sheet
for (sheet in sheet_names) {
  # Read the marker names and sample sizes
  markers <- read_excel(file_path, sheet = sheet, range = cell_rows(2:2), col_names = FALSE)
  sample_sizes <- read_excel(file_path, sheet = sheet, range = cell_rows(3:3), col_names = FALSE)

  # Read the main data (starting from the fourth row)
  data <- read_excel(file_path, sheet = sheet, skip = 3, col_names = FALSE)

  # Set the column names
  colnames(data) <- c("allele", as.character(unlist(markers)))

  # Transform the data to long format
  tidy_data <- data %>%
    pivot_longer(cols = -allele, names_to = "marker", values_to = "frequency") %>%
    rename_with(~tolower(.), everything()) # Convert all variable names to lowercase

  # Remove the rows that contain 'Hobs', 'PI', 'PPE' or 'HWE'
  tidy_data <- tidy_data[!tidy_data$allele %in% c("Hobs", "PI", "PPE", "HWE", "Allele"),] %>% 
    drop_na() %>% 
    select(marker, everything())

  # Extract population from sheet name
  tidy_data$population <- str_extract(sheet, "(?<=revised\\s)[^,]*") # Extract string between "revised " and the first comma
  
  # Clean the name for the CSV file
  filename <- gsub(" ", "", tidy_data$population[1]) # Remove spaces
  
  # Write to CSV
  write_csv(tidy_data, paste0("data/1036_", filename, ".csv")) # Write CSV to the "data" folder
}

```

## Load CODIS allele frequencies

CODIS allele frequencies were found through NIST STR base and specifically downloaded from the supplementary materials of Steffen et al 2017. These are 1036 unrelated individuals from the U.S. population. 

```{r allelefreqs}
# Define the file paths
file_paths <- list.files(path = "data", pattern = "1036_.*\\.csv", full.names = TRUE)

# Create a list of data frames
df_list <- lapply(file_paths, function(path) {
  read_csv(path, col_types = cols(
    marker = col_character(),
    allele = col_double(),
    frequency = col_double(),
    population = col_character()
  ))
})

# Bind all data frames into one
df <- bind_rows(df_list)
```

### Box plot of allele frequencies
Below we show a box plot of the distribution of allele frequencies for each marker, grouped by population. 

```{r plt-boxplot-all-group-allelefreqs, fig.width=12}
df %>%
  group_by(population) %>%
  filter(population != "all") %>%
  ggplot(aes(x = marker, y = allele, fill = population)) +
  geom_boxplot() +
  labs(x = "Marker", y = "Allele",
       title = "Range of Alleles Across Different Markers (Grouped by Population)") +
  theme_classic() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### Histogram of allele frequencies
Below we show a histogram of the distribution of allele frequencies for each marker, grouped by population. 

```{r plt-histo-all-group-allelefreqs, fig.width=12}
# Find unique markers and split them into chunks of 5
unique_markers <- unique(df$marker)
marker_chunks <- split(unique_markers, ceiling(seq_along(unique_markers)/5))

# Loop through the chunks and create a plot for each chunk
for(i in seq_along(marker_chunks)) {
  df_chunk <- df %>% filter(marker %in% marker_chunks[[i]])
  
  p <- ggplot(df_chunk, aes(x = allele, y = frequency, fill = population)) +
    geom_col(position = "dodge", width = 0.7) +
    facet_grid(population ~ marker) +
    # scale_x_continuous(breaks = seq(2.2, 43.2, by = 1)) +
    labs(x = "Allele", y = "Frequency",
         title = paste("Allele Frequencies for Markers", i*5-4, "to", min(i*5, length(unique_markers))),
         caption = "Source: CSV data") +
    theme_bw()

  print(p)
}



```

### Summary of allele frequencies
```{r df-summary}
summary_df <- df %>%
  group_by(marker, population) %>%
  summarise(num_alleles = n_distinct(allele))

```

```{r df-summary-wide}
library(knitr)
library(kableExtra)

summary_df <- df %>%
  group_by(marker, population) %>%
  summarise(num_alleles = n_distinct(allele), .groups = 'drop')

summary_pivot <- summary_df %>%
  pivot_wider(names_from = population, values_from = num_alleles)

# Print the table using kable and kableExtra for styling
summary_pivot %>%
  kable("markdown", caption = "Summary of Distinct Alleles by Marker and Population") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))

```

## Simulating genotypes

### One marker at a time



```{r func-sim-geno-single}
# Function to simulate genotypes for a pair of individuals
simulate_genotypes <- function(allele_frequencies, relationship_type) {
  # Draw two alleles for the first individual from the population allele frequencies
  individual1 <- sample(names(allele_frequencies), size = 2, replace = TRUE, prob = allele_frequencies)
  
  # Define the probabilities of sharing 0, 1, or 2 IBD alleles for each relationship type
  relationship_probs <- list(
    'parent_child' = c(0, 1, 0), # In a parent-child relationship, always 1 allele is shared
    'full_siblings' = c(1/4, 1/2, 1/4), # For full siblings, the probabilities are 1/4 for sharing 0, 1/2 for sharing 1, and 1/4 for sharing 2 alleles
    'half_siblings' = c(1/2, 1/2, 0), # For half siblings, the probabilities are 1/2 for sharing 0 and 1/2 for sharing 1 allele
    'cousins' = c(7/8, 1/8, 0), # For cousins, the probabilities are 7/8 for sharing 0 and 1/8 for sharing 1 allele
    'second_cousins' = c(15/16, 1/16, 0), # For second cousins, the probabilities are 15/16 for sharing 0 and 1/16 for sharing 1 allele
    'unrelated' = c(1, 0, 0) # For unrelated individuals, they always share 0 alleles
  )
  
  # Get the probabilities of sharing alleles for the specific relationship type
  prob_shared_alleles <- relationship_probs[[relationship_type]]
  
  # Draw the number of shared alleles based on these probabilities
  num_shared_alleles <- sample(c(0, 1, 2), size = 1, prob = prob_shared_alleles)
  
  # Construct the genotype of the second individual by sampling the shared alleles from the first individual 
  # and the rest from the population allele frequencies
  individual2 <- c(sample(individual1, size = num_shared_alleles), sample(names(allele_frequencies), size = 2 - num_shared_alleles, replace = TRUE, prob = allele_frequencies))
  
  # Return the genotypes of the two individuals and the number of shared alleles
  return(list(individual1 = individual1, individual2 = individual2, num_shared_alleles = num_shared_alleles))
}

# Function to calculate the index of relatedness
calculate_relatedness <- function(simulated_genotypes, allele_frequencies) {
  # Calculate the number of alleles that the two individuals share
  num_shared_alleles <- sum(simulated_genotypes$individual1 %in% simulated_genotypes$individual2)
  
  # Calculate the index of relatedness as the number of shared alleles divided by the sum of the inverse of allele frequencies 
  # of the alleles in the genotypes of both individuals. This gives a higher weight to rare alleles.
  relatedness <- num_shared_alleles / (sum(1 / allele_frequencies[simulated_genotypes$individual1]) + sum(1 / allele_frequencies[simulated_genotypes$individual2]))
  
    # Return the index of relatedness and the number of shared alleles
  return(list(relatedness = relatedness, num_shared_alleles = simulated_genotypes$num_shared_alleles))
}

# Function to simulate genotypes and calculate relatedness for different relationships
simulate_relatedness <- function(df, marker, population, relationship_type) {
  # Filter the data frame to get the allele frequencies for the specific marker and population
  allele_frequencies <- df %>%
    filter(marker == marker, population == population) %>%
    pull(frequency) %>%
    setNames(df$allele)
  
  # Simulate the genotypes for the pair of individuals using these allele frequencies and the specific relationship type
  simulated_genotypes <- simulate_genotypes(allele_frequencies, relationship_type)
  
  # Calculate the relatedness index based on these simulated genotypes and the allele frequencies
  relatedness_data <- calculate_relatedness(simulated_genotypes, allele_frequencies)
  
  # Return the marker, population, relationship type, and the calculated relatedness data
  return(c(list(marker = marker, population = population, relationship_type = relationship_type), relatedness_data))
}

# Example usage
simulate_relatedness(df, marker = "F13A01", population = "Asian", relationship_type = "full_siblings")


```


```{r func-iter-rel-single}
simulate_all_relationships <- function(df, num_simulations) {
  # Define the list of relationship types
  relationship_types <- c('parent_child', 'full_siblings', 'half_siblings', 'cousins', 'second_cousins', 'unrelated')
  
  # Initialize an empty list to store results
  results <- list()
  
  # Iterate over all combinations of markers, populations, and relationship types
  for (marker in unique(df$marker)) {
    for (population in unique(df$population)) {
      for (relationship_type in relationship_types) {
        for (i in 1:num_simulations) {
          # Simulate relatedness and add the result to the list
          results[[length(results) + 1]] <- simulate_relatedness(df, marker, population, relationship_type)
        }
      }
    }
  }
  
  # Convert the list of results to a dataframe
  results_df <- do.call(rbind, lapply(results, function(x) as.data.frame(t(unlist(x)))))
  
  return(results_df)
}
```


```{r df-sim-single}
# Usage:
# df <- # Your dataframe here
results_df <- simulate_all_relationships(df, num_simulations = 10)

```

#### Visualization

```{r func-plt-single}
# Function to capitalize the first letter of a string
ucfirst <- function(s) {
  paste(toupper(substring(s, 1,1)), substring(s, 2), sep = "")
}

create_plot <- function(df, variable_to_plot) {
  # Create the plot
  p <- ggplot(df, aes(x = relationship_type, y = .data[[variable_to_plot]], color = population, shape = population, fill = population)) +
    geom_boxplot(alpha = 0.4) + # Change the order of geom_boxplot() and geom_point() and adjust alpha
    geom_point(position = position_jitterdodge(jitter.width = 0.1, dodge.width = 0.75), size = 1, alpha = 0.6) +
    facet_grid(. ~ marker) +
    theme_classic() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    scale_x_discrete(limits = c('parent_child', 'full_siblings', 'half_siblings', 'cousins', 'second_cousins', 'unrelated')) +
    labs(title = paste(ucfirst(variable_to_plot), "by Relationship Type, Population, and STR Marker"),
         x = "Relationship Type", 
         y = ucfirst(variable_to_plot), 
         color = "Population",
         shape = "Population")
  
  return(p)
}
```

```{r plt-single, fig.width=8}
# Filter your data for the first 3 unique STR markers
factor_vars <- c("marker", "population", "relationship_type")
numeric_vars <- c("relatedness", "num_shared_alleles")

# Filter your data for unique STR markers and remove the "all" population
results_df_filtered <- results_df %>%
  filter(population != "all") %>%
  mutate(
    across(all_of(factor_vars), as.factor),
    across(all_of(numeric_vars), as.numeric)
  )

# Find the total number of unique markers
total_markers <- length(unique(results_df_filtered$marker))

# Iterate through the unique markers in groups of 3
for (i in seq(1, total_markers, by = 3)) {
  # Filter the data for the current set of 3 or fewer markers
  filtered_results_df <- results_df_filtered %>%
    filter(marker %in% unique(marker)[i:min(i+2, total_markers)])
  
  # Create and display the plots for the current set of markers
  for (variable_to_plot in c("relatedness", "num_shared_alleles")) {
    plot <- create_plot(df = filtered_results_df, variable_to_plot = variable_to_plot)
    print(plot) # Print the plot to display it in the RMarkdown document
  }
}
```

### All Markers together


```{r}
# Function to simulate genotypes for a pair of individuals for all markers
simulate_genotypes_all_markers <- function(df, relationship_type, population) {
  markers <- unique(df$marker)
  
  # Simulate the first individual's alleles by drawing from the population frequency for each marker
  individual1 <- setNames(lapply(markers, function(marker) {
    allele_frequencies <- df %>%
      filter(marker == marker, population == population) %>%
      pull(frequency) %>%
      setNames(df$allele)
      
    sample(names(allele_frequencies), size = 2, replace = TRUE, prob = allele_frequencies)
  }), markers)
  
  # Relationship probabilities
  relationship_probs <- list(
    'parent_child' = c(0, 1, 0),
    'full_siblings' = c(1/4, 1/2, 1/4),
    'half_siblings' = c(1/2, 1/2, 0),
    'cousins' = c(7/8, 1/8, 0),
    'second_cousins' = c(15/16, 1/16, 0),
    'unrelated' = c(1, 0, 0)
  )
  
  prob_shared_alleles <- relationship_probs[[relationship_type]]
  num_shared_alleles <- sample(c(0, 1, 2), size = 1, prob = prob_shared_alleles)
  
  individual2 <- setNames(lapply(markers, function(marker) {
  allele_frequencies <- df %>%
    filter(marker == marker, population == population) %>%
    arrange(marker, allele) %>%
    pull(frequency) %>%
    setNames(df$allele)
  
  prob_shared_alleles <- relationship_probs[[relationship_type]]
  non_zero_indices <- which(prob_shared_alleles != 0)
  num_shared_alleles <- sample(non_zero_indices - 1, size = 1, prob = prob_shared_alleles[non_zero_indices])

  alleles_from_individual1 <- sample(individual1[[marker]], size = num_shared_alleles)
  alleles_from_population <- sample(names(allele_frequencies), size = 2 - num_shared_alleles, replace = TRUE, prob = allele_frequencies)
  
  return(c(alleles_from_individual1, alleles_from_population))
}), markers)

  # Return the simulated genotypes
  return(list(individual1 = individual1, individual2 = individual2))
}



# Function to calculate the index of relatedness for all markers
calculate_relatedness_all_markers <- function(simulated_genotypes, df, population) {
  markers <- names(simulated_genotypes$individual1)
  
  # Calculate the number of shared alleles for each marker
  num_shared_alleles <- sapply(markers, function(marker) {
    sum(simulated_genotypes$individual1[[marker]] %in% simulated_genotypes$individual2[[marker]])
  })
  
  # Calculate the index of relatedness as the number of shared alleles weighted inversely to their frequencies
  # Now considering both individuals' alleles for the inverse frequency weighting
  relatedness <- sapply(markers, function(marker) {
    allele_frequencies <- df %>%
      filter(marker == marker, population == population) %>%
      pull(frequency) %>%
      setNames(df$allele)
      
    num_shared_alleles[marker] / (sum(1 / allele_frequencies[simulated_genotypes$individual1[[marker]]]) + sum(1 / allele_frequencies[simulated_genotypes$individual2[[marker]]]))
  })
  
  # Return the index of relatedness
  return(list(relatedness = relatedness, num_shared_alleles = num_shared_alleles))
  }

# Function to simulate genotypes and calculate relatedness for different relationships for all markers
simulate_relatedness_all_markers <- function(df, relationship_type, population) {
  # Simulate genotypes for all markers
  simulated_genotypes <- simulate_genotypes_all_markers(df, relationship_type, population)
  
  # Calculate and return the relatedness for all markers
  relatedness_data <- calculate_relatedness_all_markers(simulated_genotypes, df, population)
  
   # Before returning the results_df, add marker and population information
  results_df$marker <- rownames(results_df)
  results_df$population <- population
  
  return(relatedness_data)
}


```

```{r eval=FALSE, include=FALSE}
df <- df %>%
  arrange(marker, allele)

results <- simulate_relatedness_all_markers(df, 'parent_child', 'Hispanic')

# print the results
results
```


#### Visualizations
```{r}
# Define the list of relationship types
relationship_types <- c('parent_child', 'full_siblings', 'half_siblings', 'cousins', 'second_cousins', 'unrelated')

# Create a dataframe of all combinations of populations and relationship types
combinations <- expand.grid(population = unique(df$population), relationship_type = relationship_types)

# Apply the function to each combination
results <- combinations %>%
  split(seq(nrow(.))) %>%
  map_dfr(function(combination) {
    population <- combination$population
    relationship_type <- combination$relationship_type
    # cat("Processing:", "population =", population, "; relationship_type =", relationship_type, "\n")
    sim_results <- simulate_relatedness_all_markers(df, relationship_type[[1]], population[[1]])
    
    # Bind resulting data frames
    tibble(
      population = population,
      relationship_type = relationship_type,
      sim_results = list(sim_results)
    )
  })

multi_results <- results %>% 
  mutate(
    sum_relatedness = map_dbl(sim_results, function(x) {
      sum(x[["relatedness"]], na.rm = TRUE)
    }),
    sum_alleles = map_dbl(sim_results, function(x) {
      sum(x[["num_shared_alleles"]], na.rm = TRUE)
    })
  ) %>% 
  select(-sim_results)

multi_results
```


```{r v2-multi}
# Define the number of simulations
n_sims <- 10

# Define the list of relationship types
relationship_types <- c('parent_child', 'full_siblings', 'half_siblings', 'cousins', 'second_cousins', 'unrelated')

# Create a dataframe of all combinations of populations, relationship types, and simulations
combinations <- expand.grid(population = unique(df$population), relationship_type = relationship_types, simulation = 1:n_sims)

# Apply the function to each combination
results <- combinations %>%
  split(seq(nrow(.))) %>%
  map_dfr(function(combination) {
    population <- combination$population
    relationship_type <- combination$relationship_type
    sim <- combination$simulation
    # cat("Processing:", "population =", population, "; relationship_type =", relationship_type, "; simulation =", sim, "\n")
    sim_results <- simulate_relatedness_all_markers(df, relationship_type[[1]], population[[1]])
    
    # Bind resulting data frames
    tibble(
      population = population,
      relationship_type = relationship_type,
      simulation = sim,
      sim_results = list(sim_results)
    )
  })

multi_results <- results %>% 
  mutate(
    sum_relatedness = map_dbl(sim_results, function(x) {
      sum(x[["relatedness"]], na.rm = TRUE)
    }),
    sum_alleles = map_dbl(sim_results, function(x) {
      sum(x[["num_shared_alleles"]], na.rm = TRUE)
    })
  ) %>% 
  select(-sim_results)

multi_results


```

```{r}
create_plot <- function(df, variable_to_plot) {
  # Create the plot
  p <- ggplot(df, aes(x = relationship_type, y = .data[[variable_to_plot]], color = population, shape = population, fill = population)) +
    geom_boxplot(alpha = 0.4, position = position_dodge(width = 0.75)) +
    geom_point(position = position_jitterdodge(jitter.width = 0.1, dodge.width = 0.75), size = 1, alpha = 0.6) +
    theme_classic() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    scale_x_discrete(limits = c('parent_child', 'full_siblings', 'half_siblings', 'cousins', 'second_cousins', 'unrelated')) +
    labs(title = paste(ucfirst(variable_to_plot), "by Relationship Type and Population"),
         x = "Relationship Type", 
         y = ucfirst(variable_to_plot), 
         color = "Population",
         shape = "Population")
  
  return(p)
}

```

```{r fig.width=8}

# Filter your data for unique STR markers and remove the "all" population
df_plt_multi_results <- multi_results %>%
  filter(population != "all") %>%
  select(-simulation)

p <- create_plot(df_plt_multi_results, "sum_relatedness")
p

plt_allele <- create_plot(df_plt_multi_results, "sum_alleles")

plt_allele
```



`

## Updated LR

### From Weight-of-evidence for forensic DNA profiles book

Likelihood ratio for a single locus is:

$$
R=\kappa_0+\kappa_1 / R_X^p+\kappa_2 / R_X^u
$$
Where $\kappa$ is the probability of having 0, 1 or 2 alleles IBD for a given relationship. 

The $R_X$ terms are quantifying the "surprisingness" of a particular pattern of allele sharing.

The $R_X^p$ terms attached to the $kappa_1$ are defined in the following table:

$$
\begin{aligned}
&\text { Table 7.2 Single-locus LRs for paternity when } \mathcal{C}_M \text { is unavailable. }\\
&\begin{array}{llc}
\hline c & Q & R_X \times\left(1+2 F_{S T}\right) \\
\hline \mathrm{AA} & \mathrm{AA} & 3 F_{S T}+\left(1-F_{S T}\right) p_A \\
\mathrm{AA} & \mathrm{AB} & 2\left(2 F_{S T}+\left(1-F_{S T}\right) p_A\right) \\
\mathrm{AB} & \mathrm{AA} & 2\left(2 F_{S T}+\left(1-F_{S T}\right) p_A\right) \\
\mathrm{AB} & \mathrm{AC} & 4\left(F_{S T}+\left(1-F_{S T}\right) p_A\right) \\
\mathrm{AB} & \mathrm{AB} & 4\left(F_{S T}+\left(1-F_{S T}\right) p_A\right)\left(F_{S T}+\left(1-F_{S T}\right) p_B\right) /\left(2 F_{S T}+\left(1-F_{S T}\right)\left(p_A+p_B\right)\right) \\
\hline
\end{array}
\end{aligned}
$$

For our purposes we will take out the $F_{S T}$ values. So the table will be as follows:

$$
\begin{aligned}
&\begin{array}{llc}
\hline c & Q & R_X \\
\hline \mathrm{AA} & \mathrm{AA} & p_A \\
\mathrm{AA} & \mathrm{AB} & 2 p_A \\
\mathrm{AB} & \mathrm{AA} & 2p_A \\
\mathrm{AB} & \mathrm{AC} & 4p_A \\
\mathrm{AB} & \mathrm{AB} & 4 p_A p_B/(p_A+p_B) \\
\hline
\end{array}
\end{aligned}
$$


If none of the alleles match, then the $\kappa_1 / R_X^p = 0$.

The $R_X^u$ terms attached to the $kappa_2$ are defined as:

If both alleles match and are homozygous the equation is 6.4 (pg 85).
Single locus match probability: $\mathrm{CSP}=\mathcal{G}_Q=\mathrm{AA}$
$$
\frac{\left(2 F_{S T}+\left(1-F_{S T}\right) p_A\right)\left(3 F_{S T}+\left(1-F_{S T}\right) p_A\right)}{\left(1+F_{S T}\right)\left(1+2 F_{S T}\right)}
$$
Simplified to:
$$
p_A{ }^2
$$

If both alleles match and are heterozygous, the equation is 6.5 (pg 85)
Single locus match probability: $\mathrm{CSP}=\mathcal{G}_Q=\mathrm{AB}$
$$
2 \frac{\left(F_{S T}+\left(1-F_{S T}\right) p_A\right)\left(F_{S T}+\left(1-F_{S T}\right) p_B\right)}{\left(1+F_{S T}\right)\left(1+2 F_{S T}\right)}
$$
Simplified to:

$$
2 p_A p_B
$$
If both alleles do not match then $\kappa_2 / R_X^u = 0$. 


### Calculating LR
We need 100,000 unrelated pairs. 


