---
title: "Regression to predict CODIS database sizes and racial proportions"
author: "Junhui He"
output: html_document
date: "`r format(Sys.time(), '%Y-%m-%d %H:%M:%S')`"
site: workflowr::wflow_site
editor_options:
  chunk_output_type: console
---

# 1 Introduction

The Combined DNA Index System (CODIS) is a national database of DNA profiles maintained by the FBI. It includes profiles from convicted offenders, arrestees, and forensic samples collected from crime scenes. CODIS consists of two indexing systems: the State DNA Index System (SDIS) and the National DNA Index System (NDIS). The racial composition of the forensic database has raised concerns, particularly due to the overrepresentation of Black individuals relative to their proportion in the U.S. population.

To analyze the racial demographics of the forensic database at the state level, we require two key pieces of information: (1) the racial breakdown of forensic samples in each state, and (2) the total size of the forensic database in each state. However, since racial composition data is not publicly available for all states, we must estimate the racial proportions using the data that is available.

The available data sources include:

*   Direct data on the racial composition in SDIS, from a Freedom of Information Act (FOIA) request conducted by Muprhy and Tong (2020) for the following seven states: California, Florida, Indiana, Maine, Nevada, South Dakota, and Texas. [1]
*   The total number and the racial composition of incarecated people in prison at each state. Klein et al. (2023) [2] provided the prison data for each US state except for Michigan state. For the state of Michigan, this data was pulled from  a separate source [5].
*   The number of people in the State DNA Indexing System (SDIS) and the National DNA Indexing System (NDIS). NDIS data is publicly available for all 50 states [3], whereas SDIS data is only available on a state-by-state basis and is obtained from internet searches.
*   The U.S. Census data for each state [4], which provides the total population and the racial composition of the general population. This data is available through the U.S. Census Bureau's API.

Based on the available data, the total size of the forensic database in each state is determined using the number of individuals in SDIS when such data is available. For states without SDIS data, the number of individuals in NDIS is used instead. To estimate the racial composition of each state's forensic database, we apply a regression model that incorporates two factors: the racial composition of the general population and the racial composition of the prison population within the state.

The plot below illustrates which data sources are available for each state:


```{r setup, include=FALSE, echo=FALSE}
library(here)
setwd(here::here())
knitr::opts_chunk$set(echo = TRUE)

# Load necessary libraries
library(readr)        # read csv files
library(tidyverse)    # census API
library(knitr)        
library(jtools) 
library(ggpubr)
library(sandwich)     # robust covariance estimator
library(ggplot2)      # For plotting
library(stats)        # For statistical functions
library(tidycensus)   # For accessing US census data
# Load additional libraries for combining plots
library(cowplot)
library(ggpubr) 
library(viridis)
library(sf)

# Load prison data
prison_data = read.csv("data/populations_states.csv")

codis_data = read.csv("data/codis_data.csv") # CODIS data

data.sources = data.frame(
  state = c("Alabama","Alaska","Arizona","Arkansas","California","Colorado" ,"Delaware" ,"Florida","Georgia","Hawaii" ,       
"Idaho","Illinois","Indiana","Iowa", "Kansas", "Kentucky","Louisiana","Maine","Maryland","Massachusetts" ,
"Michigan", "Minnesota","Mississippi","Missouri","Montana","Nebraska", "Nevada" , "New Hampshire","New Jersey","New Mexico" ,   
"New York","North Carolina", "North Dakota" ,"Ohio" , "Oklahoma","Oregon","Pennsylvania","Rhode Island", "South Carolina", "South Dakota"  ,
"Tennessee","Texas" ,"Utah" ,"Vermont","Virginia","Washington","West Virginia","Wisconsin","Wyoming"), 
  data = c("SDIS+NDIS","SDIS+NDIS", "NDIS only", "SDIS+NDIS", "Murphy & Tong","SDIS+NDIS", "NDIS only", "Murphy & Tong" ,"SDIS+NDIS", "NDIS only",
"SDIS+NDIS", "SDIS+NDIS", "Murphy & Tong", "NDIS only", "NDIS only", "NDIS only", "SDIS+NDIS", "Murphy & Tong" , "SDIS+NDIS", "NDIS only",
"NDIS only", "SDIS+NDIS" ,"SDIS+NDIS", "SDIS+NDIS" ,"SDIS+NDIS", "NDIS only", "Murphy & Tong","NDIS only", "SDIS+NDIS" ,"NDIS only",
"NDIS only", "SDIS+NDIS", "NDIS only" ,"NDIS only" ,"NDIS only" ,"NDIS only", "NDIS only", "SDIS+NDIS", "SDIS+NDIS" ,"Murphy & Tong", "SDIS+NDIS", "Murphy & Tong", "NDIS only", "NDIS only" ,"SDIS+NDIS", "SDIS+NDIS" ,"SDIS+NDIS", "NDIS only", "NDIS only")
)
```

```{r data-wrangling, results = FALSE,warning=FALSE, message=FALSE, echo = FALSE}
# Load necessary libraries

# Calculate the number of White and Black profiles in Murphy and Tong data
# murphy.tong$n = murphy.tong$total * murphy.tong$mt.percent

# Extract year from the date and filter data for 2022
prison_data$year = substring(prison_data$date, 1, 4)
prison_data_2022 = prison_data[which(prison_data$year == "2022"),]
prison_data_2022 = prison_data_2022[!duplicated(prison_data_2022[, 'state']),]

# Michigan source: Vera https://www.vera.org/downloads/pdfdownloads/state-incarceration-trends-michigan.pdf 
prison_data_2022$incarcerated_black[which(prison_data_2022$state == "Michigan")] <- 0.53 * as.numeric(prison_data_2022$incarcerated_total[which(prison_data_2022$state == "Michigan")])

# Load census data for each state
# P1_002N is the total population, P1_003N is the total White population, and P1_004N is the total Black population
us_state_density <- get_decennial(
  geography = "state",
  variables = c(all = "P1_002N", census.num.white = "P1_003N", census.num.black = "P1_004N"),
  year = 2020,
  geometry = TRUE,
  keep_geo_vars = TRUE
)


# Spread the data into a wider format
us_state_density = spread(us_state_density, variable, value)

# Calculate the proportion of Black and White populations
us_state_density$census.percent.black = us_state_density$census.num.black / us_state_density$all
us_state_density$census.percent.white = us_state_density$census.num.white / us_state_density$all

# Rename column for merging
us_state_density$state = us_state_density$NAME.x

data.source.sf = merge(us_state_density,data.sources, by = "state") 

data.source.sf = data.source.sf %>%
  tigris::shift_geometry()

ggplot() +
  geom_sf(data = data.source.sf$geometry, aes(fill = data.source.sf$data),color = NA) + 
  geom_sf(data = data.source.sf$geometry, color = "black",lwd = 0.3,alpha = 0) +
  theme_void(base_size = 11) + 
  labs(fill = "Data availability") + 
  #theme(plot.margin = unit(rep(0.5, 4), "cm")) +
  theme(legend.key.height= unit(1, 'cm'), legend.key.width= unit(1, 'cm'),
        legend.title=element_text(size=20),  legend.text=element_text(size=16)) 

# Merge census data with prison data
us_state_density = merge(us_state_density, prison_data_2022, by = "state")

# Calculate the proportion of Black and White incarcerated individuals
us_state_density$percent.black.incarc = us_state_density$incarcerated_black / us_state_density$incarcerated_total
us_state_density$percent.white.incarc = us_state_density$incarcerated_white / us_state_density$incarcerated_total

# Merge with CODIS data
us_state_density = merge(us_state_density, codis_data, by = "state")

# Calculate the number of Black and White profiles in CODIS
us_state_density$black_profiles = us_state_density$percent.black.incarc * (us_state_density$offender_profiles)
us_state_density$white_profiles = us_state_density$percent.white.incarc * (us_state_density$offender_profiles)

us_state_density_black = as.data.frame(us_state_density[,c("state","all","census.num.black","census.percent.black","incarcerated_total","incarcerated_black","percent.black.incarc","arrestee_profiles","offender_profiles","black_profiles")])

us_state_density_black$race = "Black"
colnames(us_state_density_black) = c("state","all","population","census.percent","incarcerated_total",
                                     "incarcerated_race","percent.incar","arrestee_profiles","offender_profiles","geometry","race_profiles","race")

us_state_density_white = as.data.frame(us_state_density[,c("state","all","census.num.white","census.percent.white","incarcerated_total",
                                             "incarcerated_white","percent.white.incarc", "arrestee_profiles","offender_profiles","white_profiles")])
us_state_density_white$race = "White"
colnames(us_state_density_white) = c("state","all","population","census.percent","incarcerated_total",
                                     "incarcerated_race","percent.incar","arrestee_profiles","offender_profiles","geometry","race_profiles","race")

inferred_data = rbind(us_state_density_black, us_state_density_white)

# Combine Murphy and Tong data with the merged dataset
# combined = merge(murphy.tong, inferred_data, by = c("state","race"), all.x = TRUE)
# combined$race_bin = ifelse(combined$race == "White",0,1)

```

# 2 Models to predict the racial composition of the forensic database

For states without SDIS racial composition data, we construct a multinomial regression model using stick-breaking techniques to estimate the racial composition of SDIS profiles. In the stick-breaking technique, we divide the US races into three types, black + white + other, and separately estimate the percent of white people in all people $p_{1}$ and the percent of black people in non-white people $p_{2}$. Then the racial breakdown is given by 
$$CODIS\_Prop_{white}=p_1, \quad CODIS\_Prop_{black}=(1-p_1)p_2,\quad CODIS\_Prop_{other}=(1-p_1)(1-p_2).$$
We use the following binomial regression models to estimate the racial composition of SDIS profiles:
$$P(\text{a person in CODIS is white}) = g(\beta_{10} + \beta_{11}*Census\_Prop_{white}+ \beta_{12}*Prison\_Prop_{white}),$$
and 
$$P(\text{a non-white person in CODIS is black}) = g(\beta_{20} + \beta_{21}*Census\_Prop_{black\mid nonwhite}+ \beta_{22}*Prison\_Prop_{black \mid nonwhite}),$$
where $g(x)=\frac{\exp(x)}{1+\exp(x)}$ is the logit link function. Consequently, the number of the corresponding racial groups in SDIS follows a binomial distribution with the trial number being the total number of people or non-white people in SDIS, and the success probability being the above estimated racial proportions. The independent variables are the racial proportions of population from the census database and the racial proportions of incarcerated people from the prison database.


# 3 Regression to estimate the racial composition of the SDIS database.

```{r data-racial-composition, results = FALSE,warning=FALSE,message=FALSE, echo = FALSE}
# read and preprocess data
# read data
data_path = './data'
response_data_path = file.path('./data', 'Murphy_combined_data.csv')
predictor_data_path = file.path('./data/regression_data', 'CODIS_regression_predict_data.csv')

response_data = read.csv(response_data_path) # Murphy & Tong data
predictor_data = read.csv(predictor_data_path) # Census and prison data
predictor_data[is.na(predictor_data)] = 0

idx = c(5, 9, 14, 19, 28, 41, 43) # obtain the index of seven CODIS states

# calculate the response values
y = array(dim = c(7, 5))
colnames(y) = c('white', 'black', 'total', 'nonwhite', 'other')
rownames(y) = response_data[,2]
y[,3] = response_data[,12]
y[,2] = round(y[,3] * response_data[,3])
y[,1] = round(y[,3] * response_data[,6])
y[,4] = y[,3]-y[,1]
y[,5] = y[,3]-y[,2]-y[,1]
mt_data = as.data.frame(y) 

# calculate the predictor values
x = matrix(nrow = 50, ncol = 6)
rownames(x) = predictor_data[,2]
colnames(x) = c('census.percent.white', 'census.percent.black', 'incarc.percent.white', 'incarc.percent.black', 'census.percent.black.remain', 'incarc.percent.black.remain')
x[,c(2,1)] = as.matrix(predictor_data[,c(12,13)])
x[,4] = predictor_data[,5]/(predictor_data[,3]-predictor_data[,10])
x[,3] = predictor_data[,4]/(predictor_data[,3]-predictor_data[,10])

x[,5] = x[,2] / (1 - x[,1])
x[,6] = x[,4] / (1 - x[,3])

# create training data and test data
train_data = as.data.frame(x[idx,])
test_data = as.data.frame(x[-idx,])
```

```{r binomial-regression, results = FALSE,warning=FALSE,message=FALSE, echo = FALSE}
logit <- function(y) {
  return(log(y/(1-y)))
}

ilogit <- function(x) {
  return(exp(x)/(1+exp(x)))
}

# White Americans
binomial_regression.white = glm(formula = cbind(mt_data$white, # the number of white people in CODIS
                                                mt_data$nonwhite) ~ # the number of nonwhite people in CODIS
                                  census.percent.white+ # the census proportion of white people
                                  incarc.percent.white, # the prison proportion of white people
                                family = binomial, data = train_data)



# confidence intervals
result.white = predict(binomial_regression.white, as.data.frame(x), se.fit = TRUE)
alpha = 0.05
z_hat.white = cbind(result.white$fit, result.white$fit-qnorm(1-alpha/2)*result.white$se.fit, result.white$fit+qnorm(1-alpha/2)*result.white$se.fit)
pi_hat.white = ilogit(z_hat.white)
rownames(pi_hat.white) = rownames(x)
colnames(pi_hat.white) = c('probability', 'lcl', 'ucl')

# Black Americans
binomial_regression.black = glm(formula = cbind(mt_data$black,  # the number of black people in CODIS
                                            mt_data$other) ~ # the number of other people in CODIS
                                  census.percent.black.remain + # the census proportion of black people in nonwhite people
                                  incarc.percent.black.remain, # the prison proportion of black people in nonwhite people
                                family = binomial, data = train_data)

# confidence intervals with Bonferroni methods
result.white = predict(binomial_regression.white, as.data.frame(x), se.fit = TRUE)
alpha = 0.05
z_hat.white = cbind(result.white$fit, result.white$fit+qnorm(1-alpha/4)*result.white$se.fit, result.white$fit-qnorm(1-alpha/4)*result.white$se.fit)
p_hat.white = ilogit(z_hat.white)

result.black = predict(binomial_regression.black, as.data.frame(x), se.fit = TRUE)
z_hat.black = cbind(result.black$fit, result.black$fit-qnorm(1-alpha/4)*result.black$se.fit, result.black$fit+qnorm(1-alpha/4)*result.black$se.fit)
p_hat.black = ilogit(z_hat.black)

pi_hat.black = (1-p_hat.white)*p_hat.black
rownames(pi_hat.black) = rownames(x)
colnames(pi_hat.black) = c('probability', 'lcl', 'ucl')

# Calculate CODIS black and white proportions using previous regression
p_hat = cbind(predict(binomial_regression.white, test_data, type='response'), predict(binomial_regression.black, test_data, type='response'))
pi_hat = array(dim=c(43,3))
pi_hat[,1] = p_hat[,1]
pi_hat[,2] = (1-p_hat[,1])*p_hat[,2]
pi_hat[,3] = 1 - pi_hat[,1] - pi_hat[,2]
# combine the proportions in forensic database
pis_mt = array(dim=c(50,3))
colnames(pis_mt) = c('mt.percent.white', 'mt.percent.black', 'mt.percent.other')
rownames(pis_mt) = rownames(x)
pis_mt[idx,] = cbind(response_data[,6], response_data[,3], 1 - response_data[,3] - response_data[,6])
pis_mt[-idx,] = pi_hat
pis_mt = pis_mt[,c(2, 1, 3)]
# extract the proportions in census database
pis_census = as.array(as.matrix(predictor_data[,12:14]))
colnames(pis_census) = colnames(predictor_data)[12:14]
rownames(pis_census) = rownames(pis_mt)
```

We estimate the linear coefficients $\beta$ of the binomial logistic regression for black Americans and white Americans using `glm()` function and do a Wald test on the beta coefficients. The $p$-values show that these coefficients are all statistically significant.

```{r summary-statistics-of-binomial-regression,warning=FALSE,message=FALSE, echo = FALSE}
summary(binomial_regression.white)
summary(binomial_regression.black)
```



The following plot showing the estimated racial composition using the regression model vs. the true values for the states with available data. We also plot a difference plot (also known as a Bland-Altman plot).

```{r Goodness-of-Fit, results = FALSE,warning=FALSE,message=FALSE, echo = FALSE}
# goodness of fit
pis_train = data.frame(States=rep(response_data[,2], 2), ground_truth=c(response_data[,3], response_data[,6]), fitted_value=c(pi_hat.black[idx,1], pi_hat.white[idx,1]), Races=rep(c('Black/African American', 'White American'), each=7))

ggplot(pis_train) + geom_point(aes(x=ground_truth, y=fitted_value, shape=Races, colour=States), size=4) + xlab('Murphy & Tong profiles (FOIA)') + ylab('Estimated CODIS profiles') + xlim(0,1) + ylim(0,1) + geom_abline(linetype='longdash', linewidth=1) + theme(plot.title = element_text(hjust = 0.5)) + theme_classic()

pis_train$Dif <- pis_train$fitted_value - pis_train$ground_truth
pis_train$Avg <- (pis_train$fitted_value + pis_train$ground_truth) / 2

ggplot(pis_train) + geom_point(aes(x=Avg, y=Dif, shape=Races, colour=States), size=4) + xlab('Average Measure') + ylab('Difference Between Measure') + geom_hline(yintercept = mean(pis_train$Dif), colour = "black", size = 1) +
  geom_hline(yintercept = mean(pis_train$Dif) - (1.96 * sd(pis_train$Dif)), colour = "grey30", size = 0.5) +
  geom_hline(yintercept = mean(pis_train$Dif) + (1.96 * sd(pis_train$Dif)), colour = "grey30", size = 0.5) + theme(plot.title = element_text(hjust = 0.5)) + theme_classic()
```


```{r SDIS-regression, results = FALSE,warning=FALSE,message=FALSE,out.width="100%", echo = FALSE}
# Read in CSV files
NDIS = read.csv("./data/NDIS.csv")       # National DNA Index System data
SDIS = read.csv("./data/SDIS.csv")       # State DNA Indexing System data
SDIS$N_total = ifelse(is.na(SDIS$N_total), SDIS$N_arrestees + SDIS$N_offenders, SDIS$N_total)

# Merge NDIS and SDIS datasets by state
NDIS_SDIS = merge(NDIS, SDIS, by = "State")

# Rename columns to distinguish between NDIS and SDIS data
colnames(NDIS_SDIS) <- gsub('.x','.NDIS', names(NDIS_SDIS))
colnames(NDIS_SDIS) <- gsub('.y','.SDIS', names(NDIS_SDIS))

# Merge NDIS_SDIS data with prison data
# colnames(prison_data_2022)[12] = "State"
# SDIS_states = merge(NDIS_SDIS, prison_data_2022, by = "State")

# Merge census data with NDIS_SDIS_prison data
colnames(us_state_density)[1] = "State"
us_state_population= us_state_density[, c("State", "census.num.black", "census.num.white", "census.percent.black","census.percent.white","percent.black.incarc","percent.white.incarc")]
CODIS_regression = merge(NDIS_SDIS, us_state_population, by = "State")
CODIS_regression$N_total.Census = round(CODIS_regression$census.num.black / CODIS_regression$census.percent.black)

# Calculate SDIS black and white proportions using previous regression
CODIS_regression$SDIS.black = pis_mt[,1]

CODIS_regression$SDIS.white = pis_mt[,2]

CODIS_regression$SDIS.other = pis_mt[,3]
CODIS_regression$source = "Regression"

```

# 4 Final results

Using the above regression models, we make predictions of the racial composition in SDIS by state and determine the total number of people by SDIS or NDIS. Then the number of people for each racial group in the forensic database is calculated by $N_{race} = N_{total} * Proportion_{race}$. The plot below shows our estimates for each state, colored by the data source used for each state. The number of Black people in the database are indicated with circles and the number of White people is indicated by triangles.  

```{r final-data, warning=FALSE,message=FALSE,fig.width = 20, fig.height = 10, echo = FALSE}
Murphy = read.csv("./data/Murphy_FOIA.csv") # FOIA data from Murphy

# Reshape Murphy data to wide format
Murphy_wide = reshape(Murphy, idvar = "State", timevar = "pop", direction = "wide")
Murphy_wide = Murphy_wide[, c("State", "value.Total", "value.White", "value.Black")]

# Prepare final dataset 
final.data = merge(CODIS_regression, Murphy_wide, by = "State", all.x = TRUE)
final.data$N_total.CODIS = ifelse(!is.na(final.data$N_total.SDIS), final.data$N_total.SDIS, final.data$N_total.NDIS)
final.data$final.Black = round(final.data$N_total.CODIS * final.data$SDIS.black)
final.data$final.White = round(final.data$N_total.CODIS * final.data$SDIS.white)

# Calculate final estimates for Black and White populations in CODIS based on different sources
for (i in seq_len(nrow(final.data))) {
  if (!is.na(final.data$value.Total[i])) {
    # final.data$final.Black[i] = final.data$value.Total[i] * (final.data$value.Black[i] / 100)
    # final.data$final.White[i] = final.data$value.Total[i] * (final.data$value.White[i] / 100)
    final.data$source[i] = "Murphy and Tong"
  } else if (!is.na(final.data$N_total.SDIS[i])) {
    # final.data$final.Black[i] = (final.data$N_total.SDIS[i]) * final.data$CODIS.black[i]
    # final.data$final.White[i] = (final.data$N_total.SDIS[i]) * final.data$CODIS.white[i]
    final.data$source[i] = "SDIS+regression"
  } else {
    # final.data$final.Black[i] = (final.data$SDIS.arrestee.prediction[i] + final.data$SDIS.offender.prediction[i]) * final.data$CODIS.black[i]
    # final.data$final.White[i] = (final.data$SDIS.arrestee.prediction[i] + final.data$SDIS.offender.prediction[i]) * final.data$CODIS.white[i]
    final.data$source[i] = "NDIS+Regression"
  }
}

# Plot the final estimates for Black and White populations in CODIS
ggplot(final.data) +
  geom_point(aes(x = State, y = final.Black, fill = source, shape = "Black"), size = 7, col = "black") +
  geom_point(aes(x = State, y = final.White, fill = source, shape = "White"), size = 7, col = "black") +
  scale_fill_manual(name = "Source", labels = c("Murphy", "Regression only", "SDIS+regression"), values = c("#9A77CF", "#EC4176", "#FFA45E")) +
  scale_shape_manual(name = "Race", labels = c("Black", "White"), values = c(21, 24)) +
  theme_bw() +
  guides(fill = guide_legend(override.aes = list(shape = 21))) +
  theme(axis.text.x = element_text(size = 18, angle = 90, vjust = 0.5, hjust = 1),
        axis.text.y = element_text(size = 18),
        plot.title = element_text(size = 28, hjust = 0.5),
        axis.title=element_text(size=20),
        legend.text=element_text(size=16),
        legend.title = element_text(size=20)) +
  ylab("Number of People in CODIS") + ggtitle("The Number of Black and White American in CODIS by State")
```

The table below reports the estimates for the racial composition and the number of people for each racial group in SDIS by state, along with the source of the data is below:
```{r df, warning=FALSE,message=FALSE,fig.width = 20, echo = FALSE}
final.df = final.data[,c("State","SDIS.black","SDIS.white","final.Black","final.White","source")]
colnames(final.df) = c("State","Black Proportion","White Proportion","Black Number", "White Number", "Source")
kable(final.df, caption = "Estimated racial composition of SDIS by state")
# write.csv(final.data, "../data/final_CODIS_data.csv")
```

we generate side-by-side pie charts showing the comparison between the  racial composition of the census (left) and the estimated racial composition of SDIS (right) for each state.

```{r pie-charts, warning=FALSE,message=FALSE,fig.width = 20,fig.height = 10, echo = FALSE}
final.data$census.percent.other = 1 - (final.data$census.percent.white + final.data$census.percent.black)
final.data.long = final.data %>%
  pivot_longer(
    cols = c("census.percent.white","census.percent.black", "census.percent.other"),
    names_to = "race",
    values_to = "census.proportion"
  )


final.data.long.CODIS = final.data %>%
  pivot_longer(
    cols = c("SDIS.black","SDIS.white", "SDIS.other"),
    names_to = "race",
    values_to = "CODIS.proportion"
  )

plot_list = list()
i=1
for(state in final.data$State){
  long_data = final.data.long[which(final.data.long$State == state),]
  long_data_CODIS = final.data.long.CODIS[which(final.data.long.CODIS$State == state),]

  p1 = ggplot(data=long_data, aes(x="", y=census.proportion, fill=race)) +
  geom_bar(stat="identity", width=1,color = "black") +
  scale_fill_manual(name = "",values = c("#648FFF","grey","#DC267F"),labels = c("Black/African American","Other American","White American")) +
  coord_polar("y", start=0) + theme_classic() +
  xlab("") + ylab("") + ggtitle("  Census") +
  theme(legend.position = "none", axis.line=element_blank(),axis.ticks=element_blank()) +
    theme(panel.background = element_blank(), plot.background = element_blank(),
        axis.text = element_blank()) +
    theme(plot.margin = unit(c(-0.1, -0.1, -0.1, -0.1), 
                                "inches")) 
  
  p2 = ggplot(data=long_data_CODIS, aes(x="", y=CODIS.proportion, fill=race)) +
  geom_bar(stat="identity", width=1,color = "black") +
  scale_fill_manual(name = "",values = c("#648FFF","grey","#DC267F"),labels = c("Black/African American", "Other American","White American")) +
  coord_polar("y", start=0) + theme_classic() +
  xlab("") + ylab("") + ggtitle("   SDIS") +
  theme(legend.position = "none", axis.line=element_blank(),axis.ticks=element_blank()) +
    theme(panel.background = element_blank(), plot.background = element_blank(),
        axis.text = element_blank())  +
    theme(plot.margin = unit(c(-0.1, -0.1, -0.1, -0.1), 
                                "inches")) 
    p4 = ggplot() + 
  annotate("text", x = 4, y = 25, size=6, label = paste0(state)) + 
  theme_void()

  p3 = plot_grid(NULL,NULL,p1,p2,NULL,NULL,ncol = 2, nrow = 3, rel_heights = c(-0.1,1,-0.1)) 
  p4 = plot_grid(p4,NULL,p3,nrow=3, rel_heights = c(0.2,-0.2,1))+ 
    theme(plot.background = element_rect(color="black"),plot.margin = unit(c(0.1, 0.1, -0.5, 0.1), "inches")) 
  plot_list[[i]] = p4
  i=i+1
}

plot1 = plot_grid(plotlist = plot_list, ncol = 10)
if(FALSE) {
plot1 = 
  plot_grid(plot_list[[1]],plot_list[[2]], plot_list[[3]], plot_list[[4]], plot_list[[5]], plot_list[[6]], plot_list[[7]], plot_list[[8]], plot_list[[9]], plot_list[[10]],
          plot_list[[11]], plot_list[[12]], plot_list[[13]], plot_list[[14]], plot_list[[15]], plot_list[[16]], plot_list[[17]], plot_list[[18]], plot_list[[19]], 
          plot_list[[20]], plot_list[[21]], plot_list[[22]], plot_list[[23]], plot_list[[24]], plot_list[[25]], plot_list[[26]], plot_list[[27]], plot_list[[28]],                     plot_list[[29]], plot_list[[30]], plot_list[[31]], plot_list[[32]], plot_list[[33]], plot_list[[34]], plot_list[[35]], plot_list[[36]], plot_list[[37]],                     plot_list[[38]], plot_list[[39]], plot_list[[40]], plot_list[[41]], plot_list[[42]], plot_list[[43]], plot_list[[44]], plot_list[[45]], plot_list[[46]],                     plot_list[[47]], plot_list[[48]], plot_list[[49]], ncol = 10)
}
```


```{r pie-charts-final, echo = FALSE, warning=FALSE,message=FALSE,fig.width = 20,fig.height = 12}
g_legend <- function(a.gplot){ 
    tmp <- ggplot_gtable(ggplot_build(a.gplot)) 
    leg <- which(sapply(tmp$grobs, function(x) x$name) == "guide-box") 
    legend <- tmp$grobs[[leg]] 
    legend
} 

leg = g_legend(ggplot(data=long_data_CODIS, aes(x="", y=CODIS.proportion, fill=race)) +
  geom_bar(stat="identity", width=1,color = "black") +
  scale_fill_manual(name = "",values = c("#648FFF","grey","#DC267F"),labels = c("Black/African American","Other American","White American")) +
  coord_polar("y", start=0) + theme_classic() +
  xlab("") + ylab("") + ggtitle(paste0(state)) +
  theme(legend.position = "bottom", axis.line=element_blank(),axis.ticks=element_blank()) +
    theme(panel.background = element_blank(), plot.background = element_blank(),
        axis.text = element_blank(),legend.text=element_text(size=20)) )

title <- ggdraw() +
  draw_label(
    "Racial composition of Census (left) and SDIS (right) for each state", 
    fontface = "bold", 
    size = 25,       # adjust size as you like
    hjust = 0.5
  )

plot_grid(title, plot1, leg, nrow = 3, rel_heights = c(0.1,0.9,0.1))

```

Furthermore, we calculate the weighted averages of the racial proportions in Census and SDIS across seven states provided in Murphy and Tong [1] and all states, weighted by the corresponding total number of people in each state. The following table shows the weighted racial proportions in Census and SDIS across states.


```{r weighted-averages, warning=FALSE,message=FALSE, echo = FALSE}
# Calculate the weighted averages of the racial proportions in Census and SDIS across seven states provided in Murphy and Tong [1] and all states
weighted_props_all_census = data.frame(race = c("Black/African American", "White American", "Other American"), proportion = c(sum(final.data$census.percent.black * final.data$N_total.Census) / sum(final.data$N_total.Census), sum(final.data$census.percent.white * final.data$N_total.Census) / sum(final.data$N_total.Census), sum(final.data$census.percent.other * final.data$N_total.Census) / sum(final.data$N_total.Census)))

weighted_props_all_sdis = data.frame(race = c("Black/African American", "White American", "Other American"), proportion = c(sum(final.data$SDIS.black * final.data$N_total.CODIS) / sum(final.data$N_total.CODIS), sum(final.data$SDIS.white * final.data$N_total.CODIS) / sum(final.data$N_total.CODIS), sum(final.data$SDIS.other * final.data$N_total.CODIS) / sum(final.data$N_total.CODIS)))

final.mt = final.data[which(final.data$source == "Murphy and Tong"), ]
weighted_props_seven_census = data.frame(race = c("Black/African American", "White American", "Other American"), proportion = c(sum(final.mt$census.percent.black * final.mt$N_total.Census) / sum(final.mt$N_total.Census), sum(final.mt$census.percent.white * final.mt$N_total.Census) / sum(final.mt$N_total.Census), sum(final.mt$census.percent.other * final.mt$N_total.Census) / sum(final.mt$N_total.Census)))

weighted_props_seven_sdis = data.frame(race = c("Black/African American", "White American", "Other American"), proportion = c(sum(final.mt$SDIS.black * final.mt$N_total.CODIS) / sum(final.mt$N_total.CODIS), sum(final.mt$SDIS.white * final.mt$N_total.CODIS) / sum(final.mt$N_total.CODIS), sum(final.mt$SDIS.other * final.mt$N_total.CODIS) / sum(final.mt$N_total.CODIS)))
```

```{r plot-weighted-averages, warning=FALSE,message=FALSE, echo = FALSE}
p1 = ggplot(data=weighted_props_seven_census, aes(x="", y=proportion, fill=race)) +
  geom_bar(stat="identity", width=1,color = "black") +
  scale_fill_manual(name = "",values = c("#648FFF","grey","#DC267F"),labels = c("Black/African American","Other American","White American")) +
  coord_polar("y", start=0) + theme_classic() +
  xlab("") + ylab("") + ggtitle("Census across seven states") +
  theme(plot.title = element_text(size=10, face="bold", hjust = 0.5)) + 
  theme(legend.position = "none", axis.line=element_blank(),axis.ticks=element_blank()) +
    theme(panel.background = element_blank(), plot.background = element_blank(),
        axis.text = element_blank()) +
    theme(plot.margin = unit(c(-0.1, -0.1, -0.1, -0.1), 
                                "inches")) 

p2 = ggplot(data=weighted_props_seven_sdis, aes(x="", y=proportion, fill=race)) +
  geom_bar(stat="identity", width=1,color = "black") +
  scale_fill_manual(name = "",values = c("#648FFF","grey","#DC267F"),labels = c("Black/African American","Other American","White American")) +
  coord_polar("y", start=0) + theme_classic() +
  xlab("") + ylab("") + ggtitle("SDIS across seven states") +
  theme(plot.title = element_text(size=10, face="bold", hjust = 0.5)) + 
  theme(legend.position = "none", axis.line=element_blank(),axis.ticks=element_blank()) +
    theme(panel.background = element_blank(), plot.background = element_blank(),
        axis.text = element_blank()) +
    theme(plot.margin = unit(c(-0.1, -0.1, -0.1, -0.1), 
                                "inches")) 

p3 = ggplot(data=weighted_props_all_census, aes(x="", y=proportion, fill=race)) +
  geom_bar(stat="identity", width=1,color = "black") +
  scale_fill_manual(name = "",values = c("#648FFF","grey","#DC267F"),labels = c("Black/African American","Other American","White American")) +
  coord_polar("y", start=0) + theme_classic() +
  xlab("") + ylab("") + ggtitle("Census across all states") +
  theme(plot.title = element_text(size=10, face="bold", hjust = 0.5)) + 
  theme(legend.position = "none", axis.line=element_blank(),axis.ticks=element_blank()) +
    theme(panel.background = element_blank(), plot.background = element_blank(),
        axis.text = element_blank()) +
    theme(plot.margin = unit(c(-0.1, -0.1, -0.1, -0.1), 
                                "inches")) 

p4 = ggplot(data=weighted_props_all_sdis, aes(x="", y=proportion, fill=race)) +
  geom_bar(stat="identity", width=1,color = "black") +
  scale_fill_manual(name = "",values = c("#648FFF","grey","#DC267F"),labels = c("Black/African American","Other American","White American")) +
  coord_polar("y", start=0) + theme_classic() +
  xlab("") + ylab("") + ggtitle("SDIS across all states") +
  theme(plot.title = element_text(size=10, face="bold", hjust = 0.5)) + 
  theme(legend.position = "none", axis.line=element_blank(),axis.ticks=element_blank()) +
    theme(panel.background = element_blank(), plot.background = element_blank(),
        axis.text = element_blank()) +
    theme(plot.margin = unit(c(-0.1, -0.1, -0.1, -0.1), 
                                "inches")) 

plot1 = plot_grid(p1, p2, p3, p4, ncol = 2)

g_legend <- function(a.gplot){ 
    tmp <- ggplot_gtable(ggplot_build(a.gplot)) 
    leg <- which(sapply(tmp$grobs, function(x) x$name) == "guide-box") 
    legend <- tmp$grobs[[leg]] 
    legend
} 

leg = g_legend(ggplot(data=weighted_props_seven_census, aes(x="", y=proportion, fill=race)) +
  geom_bar(stat="identity", width=1,color = "black") +
  scale_fill_manual(name = "",values = c("#648FFF","grey","#DC267F"),labels = c("Black/African American","Other American","White American")) +
  theme(legend.position = "bottom", axis.line=element_blank(),axis.ticks=element_blank()) +
    theme(panel.background = element_blank(), plot.background = element_blank(),
        axis.text = element_blank(),legend.text=element_text(size=10)))

title <- ggdraw() +
  draw_label(
    "Weighted racial proportions in Census and SDIS", 
    fontface = "bold", 
    size = 15,       # adjust size as you like
    hjust = 0.5
  )

plot_grid(title, plot1, leg, nrow = 3, rel_heights = c(0.1,0.5,0.1))
```



We additionally generate plots showing the difference in the racial proportions between SDIS and Census in each state. 

```{r maps, warning=FALSE,message=FALSE,fig.width = 20,fig.height = 15, echo = FALSE}
final.data.sf = st_as_sf(final.data)
final.data.sf = final.data.sf %>%
  tigris::shift_geometry()

ggplot() +
  geom_sf(data = final.data.sf$geometry, aes(fill = final.data.sf$SDIS.black-final.data.sf$census.percent.black),color = NA) + 
  geom_sf(data = final.data.sf$geometry, color = "black",lwd = 0.3,alpha = 0) +
  theme_void(base_size = 11) + 
  scale_fill_viridis(option = "magma", direction = -1)+
  labs(fill = "Difference between proportions in SDIS and Census for Black Americans") + 
  #theme(plot.margin = unit(rep(0.5, 4), "cm")) +
  theme(legend.key.height= unit(1, 'cm'), legend.key.width= unit(1, 'cm'),
        legend.title=element_text(size=20),  legend.text=element_text(size=20)) 

ggplot() +
  geom_sf(data = final.data.sf$geometry, aes(fill = final.data.sf$SDIS.white-final.data.sf$census.percent.white),color = NA) + 
  geom_sf(data = final.data.sf$geometry, color = "black",lwd = 0.3,alpha = 0) +
  theme_void(base_size = 11) + 
  scale_fill_viridis(option = "magma", direction = -1)+
  labs(fill = "Difference between proportions in SDIS and Census for White Americans") + 
  #theme(plot.margin = unit(rep(0.5, 4), "cm")) +
  theme(legend.key.height= unit(1, 'cm'), legend.key.width= unit(1, 'cm'),
        legend.title=element_text(size=20),  legend.text=element_text(size=20)) 
```

We conclude that Black/African Americans are overrepresented and white Americans are underrepresented in the forensic database compared to Census.

# References

[1] Murphy, Erin, and Jun H. Tong. "The racial composition of forensic DNA databases." Calif. L. Rev. 108 (2020): 1847.

[2] Klein, Brennan, et al. "COVID-19 amplified racial disparities in the US criminal legal system." Nature 617.7960 (2023): 344-350.

[3] https://le.fbi.gov/science-and-lab/biometrics-and-fingerprints/codis/codis-ndis-statistics

[4] U.S. Census Bureau. "Population." https://www.census.gov/topics/population.html \

[5] https://www.vera.org/downloads/pdfdownloads/state-incarceration-trends-michigan.pdf \
