---
title: "Estimates of DTC Demographics"
author: "Tina Lasisi, Stella BooydeGraaff"
date: "`r format(Sys.time(), '%Y-%m-%d %H:%M:%S')`"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE, include=FALSE}
# Load required packages
library(tidyverse)
library(patchwork)

knitr::opts_knit$set(root.dir = "..")
knitr::opts_chunk$set(eval = TRUE, echo = FALSE, fig.width = 7, fig.height = 6)

```


# Direct-to-Consumer Database Statistics

## Accessibility of Direct-to-Consumer Databases

```{r}
dtc_access <- read_csv("data/DTC_access_export.csv", show_col_types = FALSE) 
```

```{r}
dtc_access_cleaned <- dtc_access |> 
  select(-Source_Link, -Comments) |> 
  drop_na() |> 
  mutate(summary = case_when(
    pub_access == "yes" ~ "Public Access",
    law_access == "yes" ~ "Law Enforcement Access",
    pub_access == "no" & law_access == "no" ~ "Neither"
  ))

```

Each Direct-to-Consumer company has different policies on whether their databases are accessible to law enforcement or to the public. The following chart shows the ratio of which of the six main companies allow for which group to access them.

```{r}

ggplot(dtc_access_cleaned, aes(x = "", fill = summary)) +
  geom_bar(width = 1, stat = "count") +  # Use 'count' to plot the frequencies
  coord_polar(theta = "y") +  # Convert bar chart into a pie chart
  labs(title = "Access Summary Distribution", fill = "Access Summary") +
  theme_void() +  # Remove background and axis
  scale_fill_manual(values = c("Public Access" = "#DE3A8A", 
                               "Law Enforcement Access" = "#FFB000",
                               "Neither" = "#7253FF"))

```


The companies that law enforcement can access are FamilyTreeDNA and GEDmatch. FamilyTreeDNA allows users to opt out but have a history of violating their Terms of Service. GEDmatch also sells access to law enforcement and has a history of violating their Terms of Service, but since the law enforcement used this database to find the Golden State Killer in 2018, they have created an opt-in option for sharing data with law enforcement.

The company whose database is open to the public is MyHeritage. Genetic data is not publicly available, but records such as birth, marriage, death, burial, census, military, immigration, yearbooks are. Family sites and certain settings can allow for publicly accessible genetic data as well.

The other Direct-to-Consumer companies are characterized as allowing access to neither law enforcement or the public. Ancestry DNA requires valid legal reasoning to share data with law enforcement. 23andMe has an opt-in option for scientific researcher that 81% of users do. Living DNA shares anonymous data with third parties, and has an opt-in option to share data with researchers, however, they do not sell data to law enforcement unless feel they are legally compelled.

## Direct-to-Consumer Database Total Size

```{r}
dtc_race <- read_csv("data/DTC_race_export.csv", show_col_types = FALSE)
```

The following graph demonstrates the total number of people in each Direct-to-Consumer Database. AncestryDNA has the largest DNA database at 25,000,000. The second-largest database is 23andMe, which has 10,000,000 DNA samples. Living DNA has the smallest database, with 300,000 DNA samples.

```{r}
dtc_race_cleaned <- dtc_race |> 
  select(-Source_link, -Comments) |> 
  drop_na()

dtc_race_totals <- dtc_race_cleaned |> 
  filter(pop == "Total")

ggplot(dtc_race_totals, aes(x = Database, y = value)) +
  geom_col(fill = "#F77A2E") +
  geom_text(aes(label = scales::comma(value)),
            vjust = -0.5,  # above the bar
            size = 2.5) +  # text size
  scale_y_continuous(labels = scales::comma, breaks = seq(0, 25000000, by = 5000000)) +
  labs(title = "Database Size Totals",
       x = "Database",
       y = "Number of People",
       caption = "*These values were measured in varying years between 2021 and 2024") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))
```


Another option is to show this on a log scale:

```{r}
ggplot(dtc_race_totals, aes(x = Database, y = value)) +
  geom_col(fill = "#F77A2E") +
  geom_text(aes(label = scales::comma(value)),
            vjust = -0.5,  # above the bar
            size = 2.5) +  # text size
  scale_y_continuous(trans = 'log10', labels = scales::comma) +
  labs(title = "Database Size Totals (log scale)",
       x = "Database",
       y = "Number of People (log scale)") +
  theme_minimal()

```

## Racial Composition of Direct-to-Consumer Databases

Most Direct-to-Consumer companies do not publish the racial or gender breakdown of their DNA samples. We have data from the 23andMe Research Innovation Collaborations and Replications Program, which is a portion of their database used for scientific research. The total number of DNA samples in this cohort is 8,000,000. The racial breakdown of this sample of the 23andMe database is shown below.

```{r}
dtc_23andMe <- dtc_race_cleaned |> 
  filter(Database == "23andMe", trimws(pop) != "Total") |> 
  mutate(proportion = value / sum(value)) |> 
  mutate(pop = factor(pop, levels = c('Other', 'West Asian and North African', 'Hispanic', 'Asian', 'Black', 'White')))

ggplot(dtc_23andMe, aes(x = Database, y = proportion, fill = pop)) +
  geom_col() +
  geom_text(aes(label = scales::comma(value)), position = position_stack(vjust = 0.5), color = "black", size = 1.5) +
  labs(title = "Racial Composition of 23andMe Database",
       x = "Database",
       y = "Proportion",
       fill = "Race Category") +
  theme_minimal() +
  scale_fill_manual(values = c("#5E8BFF1A", "#5E8BFF52", "#5E8BFF85","#5E8BFFC4", "#5E8BFF", "#DE3A8A")) +
  theme(plot.title = element_text(hjust = 0.5))

```

From this data, we see that the majority of the 23andMe database is White (67.5%), then the order of biggest proportion to smallest proportion is Hispanic (16.4%), Black (5.5%), Asian (5%), Other (4.6%), and West Asian and North African (1%).

