---
title: "Family size findability"
author: "Junhui He and Tina Lasisi"
date: "`r format(Sys.time(), '%Y-%m-%d %H:%M:%S')`"
site: workflowr::wflow_site
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  root.dir = normalizePath("../")
)


# load package
library(ggplot2)
library(ggpubr)

source("code/family_size_simulator.R")
```

## 1. Introduction

We aim to clarify the probability of finding at least one cousin in a database, given the family structure in the PODFRIDGE project. This probability depends on two key factors: the number of relatives—determined by the family size (number of children per family) distribution—and the database coverage, which is the proportion of individuals in the database relative to the total population.

Erlich’s earlier paper modeled the family size distribution using a Poisson distribution, a common approach in many genetic studies. However, this assumption doesn’t accurately capture real-world family sizes, which often show overdispersion and an excess of zeros (i.e., families with no children).

In this report, we simulate the family size distribution using a Zero-Inflated Negative Binomial (ZINB) distribution, which better reflects real-world data than the Poisson model. We then calculate the probability of finding at least one cousin in the database based on the simulated family sizes and different levels of database coverage.

## 2. Methods

### 2.1. Zero-Inflated Negative Binomial (ZINB) Distribution

Based on the empirical data from the US, we find that the distribution of family size is better modeled as a Zero-Inflated Negative Binomial (ZINB) distribution. The ZINB distribution is a more realistic family size distribution, as it accounts for the excess zeros (families with no children) and the overdispersion often observed in family size data.

The Zero-Inflated Negative Binomial (ZINB) distribution is a statistical model that combines a negative binomial distribution with an additional process that generates excess zeros. It is characterized by three parameters: the zero-inflated probability $p$, the mean number $\mu$ and the size $s$ of the negative binomial distribution, denoted by $\text{ZINB}(p, s, \mu) = p\delta_0 + (1-p)\text{NB}(s,\mu)$. The average family size is $(1-p)\mu$. Based on empirical data, we set the parameters to $p=0.1, ~ s=3, ~ \mu = 3$.

### 2.2. Calculating the Number of Relatives

In this report, we only consider the relatives whose generation difference between the individual and the relative is less than or equal to 1. Therefore, the first-degree relatives include parents, siblings, and children; the second-degree relatives include aunts, uncles, nephews, nieces, and half-siblings; and the third-degree relatives include first cousins. The number of relatives is calculated based on the family size distribution simulated by the ZINB distribution. For now, the number of half-siblings is set to 0. 

Specifically, if one couple is known as the ancestor couple of the individual, this couple has at least one child, and the number of children is simulated from the ZINB distribution conditional on at least one child. For simplicity, we assume that $X$ follows a ZINB distribution and $Y$ follows a ZINB distribution conditional on at least one child minus one.

The number of first-degree relatives is given by $2+Y+X$. The number of second-degree relatives is given by $4 + \sum_{i=1}^2Y_i + \sum_{i=1}^{Y}X_i$. The number of third-degree relatives is given by $\sum_{i=1}^{\sum_{j=1}^2 Y_j} X_i$.

**Remark:** For simplicity, sometimes we approximate the number of second-degree relatives as $4 + 2Y + YX$, and the number of third-degree relatives as $2YX$.

### 2.3 The Probability of Finding at Least One Cousin in the Database

The number of cousins is simulated from the ZINB distribution as above. The simulated data is denoted as $\{r_k\}_{k=1}^n$, where $n$ is the number of simulations. For each simulated number of cousins $r_k$, we assume that the number of cousins found in the database follows a binomial distribution with parameters $r_k$ and the coverage of the database $c$, denoted as $\text{Binomial}(r_k, c)$. The probability of finding at least one cousin in the database is therefore $1-(1-c)^{r_k}$. The expected probability across all simulations is given by $\frac{1}{n}\sum_{k=1}^n (1-(1-c)^{r_k})$.

## 3. Results

### 3.1. Family Size Simulation

We explore the distribution of first-, second-, and third-degree relatives as family size increases in the simulation. Specifically, we fix the zero-inflated probability $p=0.1$, and the size $s=3$ of the ZINB distribution, and vary the mean number $\mu$ from 1 to 8. The simulation is run for 1000 times, and the distribution of relatives is calculated for each mean family size. The boxplots show that as the mean family size increases, the number of first-degree, second-degree, and third-degree relatives also increases. This is expected, as larger families tend to have more relatives.

```{r family_size_simulation, message=FALSE, warning=FALSE, fig.width=12, fig.height=4}
library(tidyr)
library(dplyr)

family_size <- family_size_simulator(n = 1000, inflation = 0.1, size = 3, mu = 3)
write.csv(family_size, "output/simulated_relatives.csv")

n <- 1000
mus <- c(1:8)
family_size_simulation <- data.frame()
for(i in 1:length(mus)) {
  family_size_simulation <- rbind(family_size_simulation, family_size_simulator(n, inflation = 0.1, size = 3, mu = mus[i]))
}
family_size_simulation$mean_family_size <- factor(rep(mus, each = n))

# Define a nice color palette
library(RColorBrewer)
# Option 1: Using a gradient from light to dark blue
blue_palette <- colorRampPalette(brewer.pal(9, "Blues")[3:9])(8)

# Option 2: Using a diverging color palette (uncomment to use)
# blue_palette <- brewer.pal(8, "BuPu")

# Option 3: Using viridis colors (uncomment to use)
# library(viridis)
# blue_palette <- viridis(8, option = "D")

p1 <- ggplot(family_size_simulation, aes(x = mean_family_size, y = n_first_degree)) +
  geom_boxplot(aes(fill = mean_family_size), alpha = 0.7, outlier.alpha = 0.3) +
  scale_fill_manual(values = blue_palette) +
  labs(x = "Mean family size (children per family)", 
       y = "Number of first-degree relatives") +
  theme_minimal() +
  theme(legend.position = "none",
        panel.grid.major.x = element_blank(),
        panel.grid.minor = element_blank(),
        axis.title = element_text(size = 12),
        axis.text = element_text(size = 10))

p2 <- ggplot(family_size_simulation, aes(x = mean_family_size, y = n_second_degree)) +
  geom_boxplot(aes(fill = mean_family_size), alpha = 0.7, outlier.alpha = 0.3) +
  scale_fill_manual(values = blue_palette) +
  labs(x = "Mean family size (children per family)", 
       y = "Number of second-degree relatives") +
  theme_minimal() +
  theme(legend.position = "none",
        panel.grid.major.x = element_blank(),
        panel.grid.minor = element_blank(),
        axis.title = element_text(size = 12),
        axis.text = element_text(size = 10))

p3 <- ggplot(family_size_simulation, aes(x = mean_family_size, y = n_third_degree)) +
  geom_boxplot(aes(fill = mean_family_size), alpha = 0.7, outlier.alpha = 0.3) +
  scale_fill_manual(values = blue_palette) +
  labs(x = "Mean family size (children per family)", 
       y = "Number of third-degree relatives") +
  theme_minimal() +
  theme(legend.position = "none",
        panel.grid.major.x = element_blank(),
        panel.grid.minor = element_blank(),
        axis.title = element_text(size = 12),
        axis.text = element_text(size = 10))

# Create combined data for p4
family_size_long <- family_size_simulation %>%
  select(mean_family_size, n_first_degree, n_second_degree, n_third_degree) %>%
  pivot_longer(cols = c(n_first_degree, n_second_degree, n_third_degree),
               names_to = "degree_type",
               values_to = "count") %>%
  mutate(degree_type = factor(degree_type, 
                              levels = c("n_first_degree", "n_second_degree", "n_third_degree"),
                              labels = c("First-degree", "Second-degree", "Third-degree")))

# Define colors for degree types
degree_colors <- c("First-degree" = "#E41A1C",    # Red
                   "Second-degree" = "#377EB8",   # Blue
                   "Third-degree" = "#4DAF4A")    # Green

# Alternative color scheme (uncomment to use)
# degree_colors <- c("First-degree" = "#1B9E77",    # Teal
#                    "Second-degree" = "#D95F02",   # Orange
#                    "Third-degree" = "#7570B3")    # Purple

p4 <- ggplot(family_size_long, aes(x = mean_family_size, y = count, fill = degree_type)) +
  geom_boxplot(alpha = 0.7, outlier.alpha = 0.3, position = position_dodge(0.8)) +
  scale_fill_manual(values = degree_colors, name = "Relative Type") +
  labs(x = "Mean family size (children per family)", 
       y = "Number of relatives",
       title = "Distribution of Relatives by Degree and Family Size") +
  theme_minimal() +
  theme(panel.grid.major.x = element_blank(),
        panel.grid.minor = element_blank(),
        axis.title = element_text(size = 12),
        axis.text = element_text(size = 10),
        legend.position = "bottom",
        plot.title = element_text(size = 14, face = "bold", hjust = 0.5))

# Arrange the first three plots in a single row
ggarrange(p1, p2, p3, ncol = 3, nrow = 1, common.legend = FALSE)
```

```{r combined_plot, message=FALSE, warning=FALSE, fig.width=10, fig.height=6}
# Display p4 separately as a larger plot
p4
```


```{r family_size_simulation_incremental, message=FALSE, warning=FALSE, fig.width=12, fig.height=5}
library(tidyr)
library(dplyr)
library(ggplot2)
library(ggpubr)

# Incremental simulation from 1 to 4 with 0.1 steps
n <- 10000  # Increased to 10k for more robust estimates
mus <- seq(1, 4, by = 0.1)  # Create sequence from 1 to 4 with 0.1 increments
family_size_simulation_incremental <- data.frame()

for(i in 1:length(mus)) {
  family_size_simulation_incremental <- rbind(family_size_simulation_incremental, 
                                              family_size_simulator(n, inflation = 0.1, size = 3, mu = mus[i]))
}
family_size_simulation_incremental$mean_family_size <- rep(mus, each = n)

# Calculate summary statistics for smooth visualization
summary_stats <- family_size_simulation_incremental %>%
  group_by(mean_family_size) %>%
  summarise(
    first_median = median(n_first_degree),
    first_q25 = quantile(n_first_degree, 0.25),
    first_q75 = quantile(n_first_degree, 0.75),
    second_median = median(n_second_degree),
    second_q25 = quantile(n_second_degree, 0.25),
    second_q75 = quantile(n_second_degree, 0.75),
    third_median = median(n_third_degree),
    third_q25 = quantile(n_third_degree, 0.25),
    third_q75 = quantile(n_third_degree, 0.75),
    # Calculate totals
    total_median = median(n_first_degree + n_second_degree + n_third_degree),
    total_q25 = quantile(n_first_degree + n_second_degree + n_third_degree, 0.25),
    total_q75 = quantile(n_first_degree + n_second_degree + n_third_degree, 0.75)
  )

# Create individual smooth plots
p1_smooth <- ggplot(summary_stats, aes(x = mean_family_size)) +
  geom_ribbon(aes(ymin = first_q25, ymax = first_q75), alpha = 0.3, fill = "#E41A1C") +
  geom_line(aes(y = first_median), color = "#E41A1C", size = 1.5) +
  labs(x = "Mean family size", y = "First-degree relatives") +
  theme_minimal() +
  theme(axis.title = element_text(size = 11),
        axis.text = element_text(size = 9))

p2_smooth <- ggplot(summary_stats, aes(x = mean_family_size)) +
  geom_ribbon(aes(ymin = second_q25, ymax = second_q75), alpha = 0.3, fill = "#377EB8") +
  geom_line(aes(y = second_median), color = "#377EB8", size = 1.5) +
  labs(x = "Mean family size", y = "Second-degree relatives") +
  theme_minimal() +
  theme(axis.title = element_text(size = 11),
        axis.text = element_text(size = 9))

p3_smooth <- ggplot(summary_stats, aes(x = mean_family_size)) +
  geom_ribbon(aes(ymin = third_q25, ymax = third_q75), alpha = 0.3, fill = "#4DAF4A") +
  geom_line(aes(y = third_median), color = "#4DAF4A", size = 1.5) +
  labs(x = "Mean family size", y = "Third-degree relatives") +
  theme_minimal() +
  theme(axis.title = element_text(size = 11),
        axis.text = element_text(size = 9))

# Display three separate smooth plots in a row
ggarrange(p1_smooth, p2_smooth, p3_smooth, ncol = 3, nrow = 1)
```

```{r combined_smooth_plots, message=FALSE, warning=FALSE, fig.width=10, fig.height=7}
# Reshape data for combined plot
summary_long <- summary_stats %>%
  select(mean_family_size, first_median, second_median, third_median) %>%
  pivot_longer(cols = c(first_median, second_median, third_median),
               names_to = "degree_type",
               values_to = "median") %>%
  mutate(degree_type = factor(degree_type,
                              levels = c("first_median", "second_median", "third_median"),
                              labels = c("First-degree", "Second-degree", "Third-degree")))

# Add IQR data
summary_long_iqr <- summary_stats %>%
  select(mean_family_size, 
         first_q25, first_q75, 
         second_q25, second_q75, 
         third_q25, third_q75) %>%
  pivot_longer(cols = -mean_family_size,
               names_to = c("degree", "stat"),
               names_sep = "_",
               values_to = "value") %>%
  pivot_wider(names_from = stat, values_from = value) %>%
  mutate(degree_type = factor(paste0(degree, "_median"),
                              levels = c("first_median", "second_median", "third_median"),
                              labels = c("First-degree", "Second-degree", "Third-degree")))

# Merge median and IQR data
summary_combined <- left_join(summary_long, summary_long_iqr, 
                              by = c("mean_family_size", "degree_type"))

# Define colors
degree_colors <- c("First-degree" = "#E41A1C",    # Red
                   "Second-degree" = "#377EB8",   # Blue
                   "Third-degree" = "#4DAF4A")    # Green

# Plot: Combined curves for all degree types
p_combined_curves <- ggplot(summary_combined, aes(x = mean_family_size)) +
  geom_ribbon(aes(ymin = q25, ymax = q75, fill = degree_type), alpha = 0.3) +
  geom_line(aes(y = median, color = degree_type), size = 1.5) +
  scale_fill_manual(values = degree_colors, name = "Relative Type") +
  scale_color_manual(values = degree_colors, name = "Relative Type") +
  labs(x = "Mean family size (children per family)", 
       y = "Number of relatives",
       title = "Distribution of Relatives by Degree and Family Size") +
  theme_minimal() +
  theme(legend.position = "bottom",
        plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
        axis.title = element_text(size = 12),
        axis.text = element_text(size = 10))

# Plot: Total relatives (sum of all degrees)
p_total <- ggplot(summary_stats, aes(x = mean_family_size)) +
  geom_ribbon(aes(ymin = total_q25, ymax = total_q75), alpha = 0.3, fill = "#7570B3") +
  geom_line(aes(y = total_median), color = "#7570B3", size = 2) +
  labs(x = "Mean family size (children per family)", 
       y = "Total number of relatives",
       title = "Total Number of Relatives (All Degrees Combined)") +
  theme_minimal() +
  theme(plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
        axis.title = element_text(size = 12),
        axis.text = element_text(size = 10))

# Display both plots
p_combined_curves
p_total
```

```{r comparison, message=FALSE, warning=FALSE, fig.width=10, fig.height=10}
# Comparison of 2.5 vs 3.5 children mean using boxplots
comparison_data <- family_size_simulation_incremental %>%
  filter(mean_family_size %in% c(2.5, 3.5)) %>%
  mutate(mean_family_size = factor(mean_family_size, labels = c("2.5 children", "3.5 children")))

# Calculate total relatives for comparison
comparison_data$n_total <- comparison_data$n_first_degree + 
                           comparison_data$n_second_degree + 
                           comparison_data$n_third_degree

# Reshape for plotting
comparison_long <- comparison_data %>%
  select(mean_family_size, n_first_degree, n_second_degree, n_third_degree, n_total) %>%
  pivot_longer(cols = -mean_family_size,
               names_to = "degree_type",
               values_to = "count") %>%
  mutate(degree_type = factor(degree_type,
                              levels = c("n_first_degree", "n_second_degree", 
                                        "n_third_degree", "n_total"),
                              labels = c("First-degree", "Second-degree", 
                                        "Third-degree", "Total")))

# Define colors for the comparison
comparison_colors <- c("2.5 children" = "#FF7F00", "3.5 children" = "#1F77B4")

# Create comparison boxplot
p_comparison_box <- ggplot(comparison_long, aes(x = degree_type, y = count, fill = mean_family_size)) +
  geom_boxplot(alpha = 0.7, outlier.alpha = 0.3) +
  scale_fill_manual(values = comparison_colors, name = "Mean Family Size") +
  labs(x = "Relative Type", 
       y = "Number of Relatives",
       title = "Distribution of Relative Counts: 2.5 vs 3.5 Children per Family") +
  theme_minimal() +
  theme(legend.position = "top",
        plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
        axis.title = element_text(size = 12),
        axis.text = element_text(size = 10))

# Calculate percentage increases based on medians
pct_changes <- comparison_long %>%
  group_by(degree_type, mean_family_size) %>%
  summarise(median_count = median(count), .groups = "drop") %>%
  pivot_wider(names_from = mean_family_size, values_from = median_count) %>%
  mutate(pct_increase = round(((`3.5 children` - `2.5 children`) / `2.5 children`) * 100, 1))

# Create percentage change plot
p_pct_change <- ggplot(pct_changes, aes(x = degree_type, y = pct_increase)) +
  geom_bar(stat = "identity", fill = "#2CA02C", alpha = 0.8) +
  labs(x = "Relative Type", 
       y = "Percentage Increase (%)",
       title = "Percentage Increase in Relatives from 2.5 to 3.5 Children per Family") +
  theme_minimal() +
  theme(plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
        axis.title = element_text(size = 12),
        axis.text = element_text(size = 10)) +
  geom_text(aes(label = paste0(pct_increase, "%")), 
            vjust = -0.5, size = 4) +
  # Expand y-axis to accommodate labels
  scale_y_continuous(expand = expansion(mult = c(0, 0.15)))

# Display comparison plots
ggarrange(p_comparison_box, p_pct_change, ncol = 1, nrow = 2)
```

### 3.2. Relationship between cousin-in-database probability and average family size

In this section, we explore the relationship between family size and the probability of finding at least one cousin in a database. We simulate family sizes using the ZINB distribution and calculate the probability of finding at least one cousin in a database with varying coverages (e.g., 10%, 30%, 50%). For the ZINB distribution, we set the parameters as follows: zero-inflated probability $p=0.1$, size $s=3$, and mean number $\mu$ varying from 1 to 8. The simulation is run for 1000 times.

We draw a plot for visualizing the relationship between family size and cousin-in-database probability across the chosen database coverages. This plot illustrates that as the mean family size increases, the probability of finding at least one cousin in the database also increases. This is because larger families tend to have more cousins, which increases the chances of finding at least one cousin in a database with a given coverage. In addition, we observe that the probability of finding at least one cousin in the database increases with higher database coverage. This is because a higher coverage means that more individuals are included in the database.


```{r cousin_in_database, message=FALSE, warning=FALSE, fig.width=8, fig.height=6}
# Set up parameters for cousin probability analysis
# Using 1 to 4 with 0.5 increments
mus_cousin <- seq(1, 4, by = 0.5)
n_cousin <- 10000  # Increase simulation size for more robust estimates

# Generate simulations
cousin_simulation_data <- data.frame()
for(i in 1:length(mus_cousin)) {
  cousin_simulation_data <- rbind(cousin_simulation_data, 
                                  family_size_simulator(n_cousin, inflation = 0.1, size = 3, mu = mus_cousin[i]))
}

# Create array of cousin counts
cousin_simulation <- array(cousin_simulation_data$n_third_degree, c(n_cousin, length(mus_cousin)))

# More realistic coverage levels: 1% to 15%
coverages = c(0.01, 0.02, 0.05, 0.10, 0.15)

cousin_match_probs <- array(NA, c(length(mus_cousin), length(coverages)))

cousin_in_database_probability <- function(n_cousins, coverage = 0.1) {
  # Calculate the probability of finding at least one cousin in the database
  p_match = genetic_match(g = 2)
  return(1 - (1 - p_match * coverage) ^ n_cousins)
}

for(i in 1:length(mus_cousin)) {
  for(j in 1:length(coverages)) {
    cousin_match_probs[i, j] <- mean(cousin_in_database_probability(cousin_simulation[, i], coverage = coverages[j]))
  }
}

# Create data frame for better plotting
cousin_prob_df <- expand.grid(mean_family_size = mus_cousin, 
                              coverage = coverages) %>%
  mutate(probability = as.vector(cousin_match_probs),
         coverage_label = factor(paste0(coverage * 100, "% Coverage"),
                                levels = c("15% Coverage", "10% Coverage", 
                                          "5% Coverage", "2% Coverage", "1% Coverage")))

# Plot with all coverage levels
ggplot(cousin_prob_df, aes(x = mean_family_size, y = probability, color = coverage_label)) +
  geom_line(size = 1.2) +
  labs(x = "Mean family size (children per family)", 
       y = "Probability of finding at least one cousin", 
       title = "Cousin Database Match Probability by Family Size and Database Coverage") +
  scale_y_continuous(limits = c(0, 1), labels = scales::percent) +
  scale_color_manual(name = "Database Coverage", 
                     values = c("15% Coverage" = "#FDE725",
                                "10% Coverage" = "#5DC863",
                                "5% Coverage" = "#21908C",
                                "2% Coverage" = "#3B528B", 
                                "1% Coverage" = "#440154")) +
  theme_minimal() +
  theme(legend.position = "right",
        plot.title = element_text(size = 14, face = "bold", hjust = 0.5))

ggsave("output/cousin_probability_plot.png", width = 8, height = 6, dpi = 300)
```

## 4. Comparison: Erlich Original vs ZINB Family Size Model

To understand how realistic family size distributions affect long-range familial search probabilities, we compare Erlich et al.'s original model with our Zero-Inflated Negative Binomial (ZINB) approach. Both models aim to predict the probability of finding genetic relatives in databases of varying sizes.

### Direct Comparison: Original Erlich (r=2.5) vs ZINB (μ=2.5)

The figure below shows a side-by-side comparison of the two models, both using a mean family size of 2.5 children per couple:

```{r comparison_erlich_zinb, fig.width=12, fig.height=6}
# Function to run original Erlich-style calculations
run_erlich_original <- function(mu_value) {
  N <- 250000000
  num_K <- 1000
  m <- 6
  r <- mu_value
  
  Ks <- round(seq(from = N/num_K, to = N * 0.05, length.out = num_K))
  db_proportions <- Ks / N
  
  results <- data.frame()
  
  for (i in 1:length(db_proportions)) {
    K <- Ks[i]
    
    # Expected number of cousins in Erlich's model
    n_1c <- 4^2 * (r/2)^2
    n_2c <- 4^3 * (r/2)^3
    n_3c <- 4^4 * (r/2)^4
    n_4c <- 4^5 * (r/2)^5
    
    # Genetic match probabilities
    p_1c <- genetic_match(g = 2, m = m)
    p_2c <- genetic_match(g = 3, m = m)
    p_3c <- genetic_match(g = 4, m = m)
    p_4c <- genetic_match(g = 5, m = m)
    
    # Probability of finding at least one
    prob_1c <- 1 - exp(-n_1c * p_1c * K/N)
    prob_2c <- 1 - exp(-n_2c * p_2c * K/N)
    prob_3c <- 1 - exp(-n_3c * p_3c * K/N)
    prob_4c <- 1 - exp(-n_4c * p_4c * K/N)
    
    results <- rbind(results,
      data.frame(
        db_proportion = rep(db_proportions[i], 4),
        probability = c(prob_1c, prob_2c, prob_3c, prob_4c),
        cousin_degree = factor(c("1C", "2C", "3C", "4C"))
      )
    )
  }
  
  return(results)
}

# Function to run ZINB calculations
run_zinb_model <- function(mu_value, n_sim = 10000) {
  N <- 250000000
  num_K <- 1000
  m <- 6
  
  Ks <- round(seq(from = N/num_K, to = N * 0.05, length.out = num_K))
  db_proportions <- Ks / N
  
  # Simulate families
  sim_data <- family_size_simulator(n_sim, inflation = 0.1, size = 3, mu = mu_value)
  
  # Calculate growth factor
  effective_mu <- (1 - 0.1) * mu_value
  growth_factor <- effective_mu^2
  
  results <- data.frame()
  
  for (i in 1:length(db_proportions)) {
    db_prop <- db_proportions[i]
    
    # Genetic match probabilities
    p_1c <- genetic_match(g = 2, m = m)
    p_2c <- genetic_match(g = 3, m = m)
    p_3c <- genetic_match(g = 4, m = m)
    p_4c <- genetic_match(g = 5, m = m)
    
    # Use simulated data for 1st cousins
    prob_1c <- 1 - (1 - p_1c * db_prop)^sim_data$n_third_degree
    
    # Project higher degree cousins
    n_2c <- sim_data$n_third_degree * growth_factor
    n_3c <- sim_data$n_third_degree * growth_factor^2
    n_4c <- sim_data$n_third_degree * growth_factor^3
    
    prob_2c <- 1 - (1 - p_2c * db_prop)^n_2c
    prob_3c <- 1 - (1 - p_3c * db_prop)^n_3c
    prob_4c <- 1 - (1 - p_4c * db_prop)^n_4c
    
    results <- rbind(results,
      data.frame(
        db_proportion = rep(db_prop, 4),
        probability = c(mean(prob_1c), mean(prob_2c), mean(prob_3c), mean(prob_4c)),
        cousin_degree = factor(c("1C", "2C", "3C", "4C"))
      )
    )
  }
  
  return(results)
}

# Create side-by-side comparison for μ=2.5
library(ggplot2)
library(ggpubr)

# Get results for both models
erlich_2.5 <- run_erlich_original(2.5)
erlich_2.5$model <- "Original Erlich Model"

zinb_2.5 <- run_zinb_model(2.5)
zinb_2.5$model <- "ZINB Model"

# Define colors
cousin_colors <- c("1C" = "#E41A1C", "2C" = "#377EB8", 
                   "3C" = "#4DAF4A", "4C" = "#984EA3")

# Create plots
p_erlich <- ggplot(erlich_2.5, aes(x = db_proportion, y = probability, 
                                    color = cousin_degree)) +
  geom_line(size = 1.2) +
  scale_x_continuous(breaks = seq(0, 0.05, 0.01), 
                     limits = c(0, 0.05),
                     labels = scales::percent) +
  scale_y_continuous(limits = c(0, 1), labels = scales::percent) +
  scale_color_manual(values = cousin_colors) +
  labs(title = "Original Erlich Model (r = 2.5)",
       x = "Database Size / Population Size",
       y = "Probability of Match",
       color = "Cousin\nDegree") +
  theme_minimal() +
  theme(legend.position = "right",
        plot.title = element_text(size = 14, face = "bold"))

p_zinb <- ggplot(zinb_2.5, aes(x = db_proportion, y = probability, 
                                color = cousin_degree)) +
  geom_line(size = 1.2) +
  scale_x_continuous(breaks = seq(0, 0.05, 0.01), 
                     limits = c(0, 0.05),
                     labels = scales::percent) +
  scale_y_continuous(limits = c(0, 1), labels = scales::percent) +
  scale_color_manual(values = cousin_colors) +
  labs(title = "ZINB Model (μ = 2.5)",
       x = "Database Size / Population Size",
       y = "Probability of Match",
       color = "Cousin\nDegree") +
  theme_minimal() +
  theme(legend.position = "right",
        plot.title = element_text(size = 14, face = "bold"))

# Arrange side by side
ggarrange(p_erlich, p_zinb, ncol = 2, common.legend = TRUE, legend = "right")
```

The comparison reveals several key differences:

1. **Lower probabilities across all cousin degrees**: The ZINB model consistently predicts lower probabilities of finding relatives compared to Erlich's model.

2. **Zero inflation impact**: With 10% of families having no children in the ZINB model, many genealogical lines terminate, reducing the overall number of cousins.

3. **Variance effects**: The ZINB distribution's higher variance means some families have many relatives while others have very few, creating a more realistic but less predictable pattern.

### Sensitivity Analysis: Mean Family Size from 1.5 to 4.0

To understand how mean family size affects these probabilities, we examined a range of values from 1.5 to 4.0 children per family:

```{r zinb_family_size_comparison, fig.width=12, fig.height=10}
# Create plots for different mean family sizes
mu_values <- c(1.5, 2.0, 2.5, 3.0, 3.5, 4.0)
plot_list <- list()

for (i in 1:length(mu_values)) {
  mu <- mu_values[i]
  
  # Run ZINB model
  zinb_results <- run_zinb_model(mu)
  
  # Create plot
  p <- ggplot(zinb_results, aes(x = db_proportion, y = probability, 
                                 color = cousin_degree)) +
    geom_line(size = 1.2) +
    scale_x_continuous(breaks = seq(0, 0.05, 0.01), 
                       limits = c(0, 0.05),
                       labels = scales::percent) +
    scale_y_continuous(limits = c(0, 1), labels = scales::percent) +
    scale_color_manual(values = cousin_colors) +
    labs(title = paste0("ZINB Model (μ = ", mu, ")"),
         x = "Database Size / Population Size",
         y = "Probability of Match") +
    theme_minimal() +
    theme(legend.position = "none",
          plot.title = element_text(size = 12, face = "bold"))
  
  # Add cousin labels at the right edge
  label_data <- zinb_results %>%
    group_by(cousin_degree) %>%
    filter(db_proportion == max(db_proportion)) %>%
    slice(1)
  
  p <- p + 
    geom_text(data = label_data,
              aes(x = 0.051, y = probability, label = cousin_degree),
              hjust = 0, size = 3.5)
  
  plot_list[[i]] <- p
}

# Arrange all plots
ggarrange(plotlist = plot_list, ncol = 2, nrow = 3)
```

### Key Findings

The ZINB model reveals several important insights about familial search probabilities:

1. **Non-linear scaling**: The probability of finding relatives doesn't scale linearly with mean family size. The difference between μ=1.5 and μ=2.0 is much smaller than between μ=3.5 and μ=4.0.

2. **Database coverage threshold**: For smaller family sizes (μ < 2), even large databases (5% of population) have limited success in finding distant cousins. 

3. **Fourth cousin saturation**: With larger family sizes (μ ≥ 3), the probability of finding fourth cousins approaches certainty even with modest database coverage (2-3%).

4. **Real-world implications**: Most developed countries have fertility rates between 1.5-2.5, suggesting that the lower probability curves are more representative of actual genealogical networks.

These results suggest that Erlich et al.'s original estimates may be overly optimistic for populations with realistic family size distributions, particularly when accounting for childless couples and the high variance in family sizes observed in real populations.


## 5. Conclusion

1. The family size distribution is better modeled as a Zero-Inflated Negative Binomial (ZINB) distribution than a Poisson distribution, which accounts for the excess zeros and overdispersion often observed in realistic family size data.

2. The number of relatives increases with the mean family size,.

3. The probability of finding at least one cousin in a database increases with both the mean family size and the coverage of the database. Larger families tend to have more cousins, and higher database coverage means more individuals are included.
