---
title: "How fertility parameters affect the family structure distribution"
author: "Junhui He"
date: "`r Sys.Date()`"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

# 1 Family structure model

## 1.1 Fertility distribution

Here we establish a zero inflated negative binomial distribution for the number of children per family---**fertility distribution** with parameters: **mean fertility** ($\mu$), **dispersion/size** ($\theta$), and **zero‑inflation probability** ($\pi_0$), denoted as $ZINB(\mu, \theta, \pi_0)$. According to the definition of the negative binomial distribution, the mean and variance of $NB(\mu,\theta)$ is given by $\mu$ and $\mu + \frac{\mu^2}{\theta}$, respectively. Therefore, the mean and variance of $ZINB(\mu, \theta, \pi_0)$ is defined by $\mu(1-\pi_0)$ and $\mu(1-\pi_0) + \frac{\mu^2(1-\pi_0)}{\theta}+\mu^2\pi_0(1-\pi_0)$, respectively.

## 1.2 Sibling distribution

The **sibling distribution** is the distribution of the number of siblings per person, which can be derived from the fertility distribution. For any $k \geq 0$, the probability of having $k$ siblings is given by: $$P(num_{sib}=k) \propto (k+1)P(num_{child}=k+1).$$ This is because a person with $k$ siblings must have been born into a family with $k+1$ children, and there are $k+1$ ways to choose which child they are among those children. If the fertility distribution is $ZINB(\mu, \theta, \pi_0)$, then the sibling distribution is a negative binomial distribution with parameters **mean fertility** ($\frac{\mu(\theta+1)}{\theta}$), **dispersion/size** ($\theta+1$), denoted as $NB(\frac{\mu(\theta+1)}{\theta}, \theta+1)$. We note that the average number of siblings per person $\frac{\mu(\theta+1)}{\theta}$ is higher than the average number of children $\mu(1-\pi_0)$ minus one, called the “sibship size paradox”.

## 1.3 Fertility and sibling parameters

Based on the empirical data on U.S. women, a reasonable range of parameters for the fertility distribution is $$\mu \in [2, 6], ~ \theta \in [1, 10], ~ and ~ \pi_0 \in [0, 0.2].$$

# 2 Effect of fertility parameters on relative counts

We explore the effect of the fertility parameters on the family structure distribution, which is defined as the distribution of the number of relatives one individual has.

```{r setup, message = FALSE, warning = FALSE, echo = FALSE}
knitr::opts_chunk$set(eval = TRUE, echo = TRUE, warning = FALSE, fig.width = 9, fig.height = 7)

library(dplyr)
library(tidyr)
library(readr)
library(MASS)     ## glm.nb
library(pscl)     ## zeroinfl
library(knitr)    ## kable
```

## 2.1 Simulate the relative counts for the focal individual

```{r simulate-relatives}
simulate_relatives_zinb <- function(mu, size, pi0,
                                    n_sim = 10000,
                                    max_kids = 20) {
  draw_fert <- function(n) {
    ifelse(runif(n) < pi0, 0L,
           as.integer(pmin(rnbinom(n, size, mu = mu), max_kids)))
  }
  
  draw_sibs <- function(n) {
    as.integer(pmin(rnbinom(n, size + 1, mu = mu * (size + 1) / size), max_kids - 1))
  }
  
  # 1. Siblings for the focal individual
  siblings <- draw_sibs(n_sim)

  # 2. Aunts & uncles (no subtraction needed)
  aunts_maternal <- draw_sibs(n_sim)   # mother’s siblings
  aunts_paternal <- draw_sibs(n_sim)   # father’s siblings
  aunts_uncles   <- aunts_maternal + aunts_paternal


  # 3. Cousins: for each aunt/uncle, simulate their fertility and sum
  cousins <- purrr::map_int(aunts_uncles, function(n_au) {
    if (n_au == 0L) 0L else as.integer(sum(draw_fert(n_au)))
  })

  tibble::tibble(
    Siblings    = siblings,
    AuntsUncles = aunts_uncles,
    Cousins     = cousins,
    Parents     = 2L,
    Grandparents= 4L,
    TotalKin = siblings + aunts_uncles + cousins + 2L + 4L
  )
}
```

## 2.2 Prepare parameter combinations

```{r parameter-combinations}
# Prepare fertility parameter tables
fert_params = expand.grid(
  mu = c(3, 4, 5, 6),
  size = c(1, 4, 10),
  pi0 = c(0, 0.1, 0.2)
)
fert_params = as.data.frame(fert_params)[, c("pi0", "size", "mu")]
n_sim = 10000  # Number of simulations per parameter combination
```

## 2.3 Simulate relative counts for each parameter combination

```{r simulate-relative-counts}
relative_counts = list()
for (i in seq_len(nrow(fert_params))) {
  params = fert_params[i, ]
  relative_counts[[i]] = simulate_relatives_zinb(
    mu = params$mu,
    size = params$size,
    pi0 = params$pi0,
    n_sim = n_sim
  )
}
```

## 2.4 Summarize mean relative counts

```{r mean-relative-counts}
mean_relative_counts = t(sapply(relative_counts, colMeans))[, c("Siblings", "AuntsUncles", "Cousins", "TotalKin")]
mean_relative_counts = as.data.frame(mean_relative_counts)
colnames(mean_relative_counts) = c("mean_siblings", "mean_aunts_uncles", "mean_cousins", "mean_total")
mean_relative_counts = cbind(fert_params, mean_relative_counts)

kable(mean_relative_counts, 
      digits = 2, 
      caption = "Expected number of first-, second-, and third-degree relatives for a focal individual (simulated, by each fertility paramater combination)")
```

## 2.5 Visualize the simulated relative distributions

```{r relative-distributions, fig.cap = "Simulated relative distributions under different fertility parameters"}
library(ggplot2)
# Combine all relative counts into a single data frame
relative_counts_df = bind_rows(relative_counts)
relative_counts_df = relative_counts_df %>%
  mutate(
    mu = rep(fert_params$mu, each = n_sim),
    size = rep(fert_params$size, each = n_sim),
    pi0 = rep(fert_params$pi0, each = n_sim)
  )
# Pivot to long format for plotting
relative_counts_long = relative_counts_df %>%
  pivot_longer(cols = c(Siblings, AuntsUncles, Cousins), 
               names_to = "RelativeType", values_to = "Count") %>% 
  mutate(
    mu = factor(mu, labels = c("3", "4", "5", "6")),
    RelativeType = factor(RelativeType, levels = c("Siblings", "AuntsUncles", "Cousins")),
    size = factor(size, labels = c("theta = 1", "theta = 4", "theta = 10")),
    pi0 = factor(pi0, labels = c("pi_0 = 0", "pi_0 = 0.1", "pi_0 = 0.2"))
  )

# Plot the distributions
ggplot(relative_counts_long,
       aes(x = mu, y = Count, fill = RelativeType)) +
  geom_boxplot(position = position_dodge(width = 0.8),
               width = 0.7, alpha = .8, outlier.size = .3) +
  stat_summary(fun = mean, geom = "point", shape = 21, size = 1.4,
               aes(group = RelativeType),
               position = position_dodge(width = 0.8),
               colour = "black", fill = "white") +
  facet_grid(pi0 ~ size) +
  scale_fill_viridis_d(option = "D") +
  labs(title   = "Simulated relative distributions under different fertility parameters",
       x       = "Mean fertility (mu)",
       y       = "Relative count",
       fill    = "Relative type") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "top")
```

Based on the simulated relative distributions, we observe the following patterns: the number of relatives increases with higher mean fertility ($\mu$), and decreases as the dispersion parameter ($\theta$) increases—--since greater dispersion corresponds to lower fertility variance. Similarly, the number of relatives declines as the zero-inflation probability ($\pi_0$) increases, though this effect is less pronounced. In general, the number of siblings is lower than that of aunts and uncles, which in turn is lower than the number of cousins—--consistent with the real family structure.

## 2.6 Regression Model to analyze the effects of fertility parameters


```{r regression-model, results='asis'}
# Prepare wide fertility parameter tables
fert_params = expand.grid(
  mu = c(2, 3, 4, 5, 6),
  size = c(1, 2, 4, 6, 10),
  pi0 = c(0, 0.05, 0.1, 0.15, 0.2)
)
fert_params = as.data.frame(fert_params)[, c("pi0", "size", "mu")]
n_sim = 10000  # Number of simulations per parameter combination

# Simulate relative counts for each parameter combination
relative_counts = list()
for (i in seq_len(nrow(fert_params))) {
  params = fert_params[i, ]
  relative_counts[[i]] = simulate_relatives_zinb(
    mu = params$mu,
    size = params$size,
    pi0 = params$pi0,
    n_sim = n_sim
  )
}

# Combine all relative counts into a single data frame
relative_counts_df = bind_rows(relative_counts)
relative_counts_df = relative_counts_df %>%
  mutate(
    mu = rep(fert_params$mu, each = n_sim),
    size = rep(fert_params$size, each = n_sim),
    pi0 = rep(fert_params$pi0, each = n_sim)
  )

library(broom)

relatives <- c("Siblings", "AuntsUncles", "Cousins")

for (rel in relatives) {
  cat("\n\n### Focal-individual model for", rel, "\n\n")
  form <- as.formula(paste0(rel, " ~ mu + size + pi0"))
  lm_fit <- lm(form, data = relative_counts_df)
  lm_tidy <- broom::tidy(lm_fit, conf.int = TRUE)
  print(kable(
    lm_tidy,
    digits = 3,
    caption = paste(
      "Regression coefficients for", rel, 
      "(focal-individual)"
    )
  ))
}
```

We can conclude that the mean fertility ($\mu$) has a positive effect on the number of relatives, while the dispersion parameter ($\theta$) have a negative effects. The zero-inflation probability ($\pi_0$) has a negative effect on the number of cousins, but the effect is not significant for siblings and aunts/uncles.
