---
title: "Forensic DNA database data overview"
author: "Junhui He"
date: "`r format(Sys.time(), '%Y-%m-%d %H:%M:%S')`"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE, echo=FALSE}

knitr::opts_chunk$set(eval = TRUE, echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 7, fig.height = 6)

library(tidyverse)
library(knitr)
library(readr)        # read csv files
library(knitr)        
library(jtools) 
library(ggpubr)
library(sandwich)     # robust covariance estimator
library(ggplot2)      # For plotting
library(stats)        # For statistical functions
library(tidycensus)   # For accessing US census data
# Load additional libraries for combining plots
library(cowplot)
library(ggpubr) 
library(viridis)
library(sf)
```

# 1 Raw datasets

This section provides an overview of the raw datasets used in this analysis. The **raw** datasets are obtained from the original sources without any modifications or transformations, including the publication and websites. The datasets are stored in the `data/raw` directory of this project.

## 1.1 Murphy and Tong FOIA

Murphy and Tong forensic DNA database is obtained from the **Freedom of Information Act (FOIA)**, provided in the Appendix of the paper by Murphy and Tong (2020) [1]. 

The file path is `../data/raw/murphy_appendix.csv`.

```{r murphy_tong}
# read Murphy and Tong forensic database
mt_raw_path <- "./data/raw/murphy_appendix.csv"
mt_raw <- read_csv(mt_raw_path, show_col_types = FALSE)

kable(head(mt_raw, n = 10), caption = "Murphy and Tong forensic DNA database (raw data from FOIA)")
```

## 1.2 SDIS

The **State DNA Index System (SDIS)** data contains DNA profiles collected at the state level from individuals, crime scenes, and other sources. In some states, SDIS data is publicly available through official government websites; however, many states do not make this information accessible to the public.

The file path is `data/raw/SDIS.csv`.

```{r sdis}
# read SDIS data
sdis_raw_path <- "./data/raw/SDIS.csv"
sdis_raw <- read_csv(sdis_raw_path, show_col_types = FALSE)

kable(head(sdis_raw, n = 10), caption = "State DNA Index System (SDIS) data (raw data from state websites)")
```

## 1.3 NDIS

The **National DNA Index System (NDIS)** data contains DNA profiles collected at the national level from individuals, crime scenes, and other sources. NDIS data is maintained by FBI, and is publicly available in the federal website [2]. 

The file path is `data/raw/NDIS.csv`.

```{r ndis}
# read NDIS data
ndis_raw_path <- "./data/raw/NDIS.csv"
ndis_raw <- read_csv(ndis_raw_path, show_col_types = FALSE)

kable(head(ndis_raw, n = 10), caption = "National DNA Index System (NDIS) data (raw data from FBI)")
```

## 1.4 Prison data

Klein et al. (2023) [3] provided the number of incarcerated people of racial groups for each US state except for Michigan state. For the state of Michigan, this **Prison population** data is pulled from a separate Vera website [4].

The file path is `data/raw/prison_populations.csv`.

```{r prison}
# read prison population data
prison_raw_path <- "./data/raw/prison_populations.csv"
prison_raw <- read_csv(prison_raw_path, show_col_types = FALSE)

kable(head(prison_raw, n = 10), caption = "Prison population data (raw data from Klein's paper and Vera)")
```

## 1.5 Census data

The **Census population** database is obtained through the `tidycensus` package in R.

```{r census, results=FALSE}
# download census data for each state
# P1_002N is the total population, P1_003N is the total White population, and P1_004N is the total Black population
us_state_density <- get_decennial(
  geography = "state",
  variables = c(all = "P1_002N", census.num.white = "P1_003N", census.num.black = "P1_004N"),
  year = 2020,
  geometry = TRUE,
  keep_geo_vars = TRUE
)
```


```{r census_data}
kable(head(us_state_density, n = 10), caption = "Census population data (raw data from US Census Bureau API)")
```

# 2 Internal datasets

This section provides an overview of the internal datasets used in this analysis. The **internal** datasets are derived from the raw datasets, and are stored in the `data/internal` directory of this project.

## 2.1 Murphy and Tong processed dataset

The **Murphy and Tong processed dataset** is derived from the Murphy and Tong FOIA database. It includes the racial composition of the forensic DNA database in seven states: **California, Florida, Indiana, Maine, Nevada, South Dakota, and Texas**. 

The file path is `data/internal/murphy_processed.csv`

```{r murphy_processed}
# read Murphy and Tong processed dataset
mt_processed_path <- "./data/internal/murphy_processed.csv"
mt_processed <- read_csv(mt_processed_path, show_col_types = FALSE)

kable(head(mt_processed, n = 10), caption = "Murphy and Tong processed dataset (internal data derived from FOIA)")
```

## 2.2 Prison and Census combined dataset

The **Prison and Census combined dataset** is derived from the prison population data and census population data. It includes the racial composition of the prison and census population in each state.

The file path is `data/internal/prison_census_combined.csv`.

```{r prison_census_combined}
# read prison and census combined dataset
prison_census_combined_path <- "./data/internal/prison_census_combined.csv"
prison_census_combined <- read_csv(prison_census_combined_path, show_col_types = FALSE)

kable(head(prison_census_combined, n = 10), caption = "Prison and Census combined dataset (internal data derived from prison and census data)")
```

# 3 Gender 

In the downstream analysis, the sex ratio is fixed at **80% male / 20% female**.


# References

[1] Murphy, Erin, and Jun H. Tong. "The racial composition of forensic DNA databases." Calif. L. Rev. 108 (2020): 1847.\
[2] https://le.fbi.gov/science-and-lab/biometrics-and-fingerprints/codis/codis-ndis-statistics \
[3] Klein, Brennan, et al. "COVID-19 amplified racial disparities in the US criminal legal system." Nature 617.7960 (2023): 344-350. \
[4] https://www.vera.org/downloads/pdfdownloads/state-incarceration-trends-michigan.pdf \


