---
title: "Likelihood Ratio Distribution"
output: html_document
author: "Hannah Van Wyk"
date: "`r format(Sys.time(), '%Y-%m-%d %H:%M:%S')`"
site: workflowr::wflow_site
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r data, results =FALSE,warning=FALSE,message=FALSE, echo = FALSE}
# Load Required Libraries
library(tidyverse)
library(furrr)
library(data.table)
library(future)
library(parallel)
library(dichromat)
library(MASS)

file_type<-"pdf" #option to toggle between pdf or png outputs, pdf being faster

output_dir <- file.path("output/example_plots")
#output_dir <- file.path("output", slurm_job_id)
#dir.create(output_dir, recursive = TRUE)

# load combined_lrs
#combined_lrs = fread(paste0("output/simulation_",slurm_job_id,"/sim_summary_genotypes.csv"))
combined_lrs = fread(paste0("data/example_data/sim_summary_genotypes.csv"))

combined_lrs$LR<-as.numeric(combined_lrs$LR)
combined_lrs$relationship_tested <- factor(combined_lrs$relationship_tested, levels = c("parent_child", "full_siblings", "half_siblings", "cousins", "second_cousins", "unrelated"))
combined_lrs$population<-as.factor(combined_lrs$population)

# load proportions_exceeding_cutoffs
#proportions_exceeding_cutoffs = fread(paste0("output/simulation_",slurm_job_id,"/sim_proportions_exceeding_cutoffs.csv"))
```

```{r cutoff_functions, results =FALSE,warning=FALSE,message=FALSE, echo = FALSE,fig.width = 20,fig.height = 15}
# Function to calculate cut-off values for 1%, 0.1%, and 0.01% FPR
calculate_cutoffs <- function(input_df, fp_rates, hypothesis) {
  input_df<-input_df[input_df$relationship_known == "unrelated" & input_df$relationship_tested == hypothesis,]
  cutoffs <- input_df[,
                                  list(
      fixed_cutoff = 1.00,
      cutoff_1 = quantile(LR, probs = 1 - fp_rates[1] / 100, na.rm = TRUE),
      cutoff_0_1 = quantile(LR, probs = 1 - fp_rates[2] / 100, na.rm = TRUE),
      cutoff_0_01 = quantile(LR, probs = 1 - fp_rates[3] / 100, na.rm = TRUE),
      n_unrelated = .N),
    by=c("population", "loci_set")]
  return(cutoffs)
}

calculate_proportions_exceeding_cutoffs <- function(input_df, cutoffs, hypothesis) {
  input_df<-input_df[input_df$relationship_tested == hypothesis,]
  df_with_cutoffs <- left_join(input_df, cutoffs, by = c("population", "loci_set"))
  df_with_cutoffs <- df_with_cutoffs [,list(
      population=population,
      relationship_tested = relationship_tested,
      relationship_known = relationship_known,
      loci_set = loci_set,
      exceeds_fixed_cutoff = LR > fixed_cutoff,
      exceeds_cutoff_1 = LR > cutoff_1,
      exceeds_cutoff_0_1 = LR > cutoff_0_1,
      exceeds_cutoff_0_01 = LR > cutoff_0_01
    )]

  df_no_cutoffs<-df_with_cutoffs[!df_with_cutoffs$relationship_tested == "unrelated",]
  proportions_exceeding <- df_no_cutoffs[,    list(
      proportion_exceeding_fixed = sum(exceeds_fixed_cutoff) / .N,
      proportion_exceeding_1 = sum(exceeds_cutoff_1, na.rm = TRUE) / .N,
      proportion_exceeding_0_1 = sum(exceeds_cutoff_0_1, na.rm = TRUE) / .N,
      proportion_exceeding_0_01 = sum(exceeds_cutoff_0_01, na.rm = TRUE) / .N,
      n_related = .N),
      by=c("population","relationship_tested","loci_set", "relationship_known")]
  return(proportions_exceeding)
}
```

Likelihood ratio summary plots

```{r summary_states, results =FALSE,warning=FALSE,message=FALSE, echo = FALSE,fig.width = 20,fig.height = 15}
# make sure loci set is in the order we want
combined_lrs$loci_set_factor <- factor(combined_lrs$loci_set, levels=c("core_13", "identifiler_15", "expanded_20", "supplementary", "autosomal_29"))

# make sure relationships are in the order we want
combined_lrs$relationship_known_factor <- factor(combined_lrs$relationship_known, levels=c("parent_child", "full_siblings", "half_siblings", "cousins", "second_cousins", "unrelated"))

# race labels (can be changed)
combined_lrs <- combined_lrs %>%
                     mutate(population_label = case_when(population == "AfAm" ~ "African-American",
                                                         population == "Asian" ~ "Asian",
                                                         population == "Cauc" ~ "Caucasian",
                                                         population == "Hispanic" ~ "Hispanic"))

summary_stats <- combined_lrs[,    list(
  mean_LR = mean(LR),
  lower_95 = quantile(LR, 0.025),
  upper_95 = quantile(LR, 0.975)#,
),
by=c("relationship_tested","population_label","loci_set_factor")]

color_palette_race = c("#AA4499", "#DDCC77", "#88CCEE", "#117733")

ggplot(summary_stats, aes(x = loci_set_factor, y = mean_LR, group = population_label, color = population_label)) +
      geom_line(size = 2) +
      # geom_errorbar(aes(x = loci_set_factor, ymin=lower_95, ymax=upper_95, group = population, color = population), width=.2,
      #            position=position_dodge(.1))  +
      facet_wrap(~ relationship_tested,  ncol = 2) +
      scale_y_log10() +
      labs(
        title = "Mean LR Across Populations and Relationship Types",
        x = "Loci Set",
        y = "Combined LR",
        color = "Population"
      ) +
      scale_color_manual(values = color_palette_race) +
      theme_minimal(base_size = 24) +
      theme(
        axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "bottom"
      )

#combined_lrs$LR = ifelse(combined_lrs$LR < 1e-32, 1e-32, combined_lrs$LR)

ggplot(combined_lrs, aes(x = relationship_tested, y = LR, fill = population_label, color = population_label)) +
    geom_boxplot(position = position_dodge(width = 0.9)) +
    facet_wrap(~ loci_set_factor, scales = "fixed") +
    labs(
      title = "LR Distributions Across Populations and Relationship Types",
      x = "Relationship Tested",
      y = "LR",
      fill = "Population",
      color = "Population"
    ) +
    scale_color_manual(values = color_palette_race) +
    scale_fill_manual(values = color_palette_race) +
    theme_minimal(base_size = 24) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1)
    ) +
    scale_y_log10()

gb_df = combined_lrs %>% 
    group_by(relationship_known, relationship_tested, loci_set) %>%
    summarize(mean = mean(LR))

gb_df$indicator = ifelse(gb_df$mean == 0, "*", "")

ann_text1 <- tribble(
  ~relationship_tested, ~LR, ~relationship_known, ~loci_set, ~population_label,
  "parent_child", 0, "cousins", "expanded_20", "African-American",
  "parent_child", 0, "cousins", "supplementary", "African-American",
  "parent_child", 0, "cousins", "autosomal_29", "African-American",
  "parent_child", 0, "second_cousins", "expanded_20", "African-American",
  "parent_child", 0, "second_cousins", "supplementary", "African-American",
  "parent_child", 0, "second_cousins", "autosomal_29", "African-American",
  "parent_child", 0, "unrelated", "expanded_20", "African-American",
  "parent_child", 0, "unrelated", "supplementary", "African-American",
  "parent_child", 0, "unrelated", "autosomal_29", "African-American",
  "parent_child", 0, "unrelated", "identifiler_15", "African-American"
)
ann_text1$loci_set_factor <- factor(ann_text1$loci_set, levels=c("core_13", "identifiler_15", "expanded_20", "supplementary", "autosomal_29"))
ann_text1$relationship_known_factor <- factor(ann_text1$relationship_known, levels=c("parent_child", "full_siblings", "half_siblings", "cousins", "second_cousins", "unrelated"))

combined_lrs$trunc <- ifelse(combined_lrs$LR == 0, 1e-16, combined_lrs$LR)
n = nrow(combined_lrs[which(relationship_known == "unrelated" & relationship_tested == "parent_child"),])
ggplot(combined_lrs, aes(x = relationship_tested, y = trunc, fill = population_label)) +
    geom_boxplot(position = position_dodge(width = 0.9)) +
    facet_grid(relationship_known_factor ~ loci_set_factor, scales = "free_y") +
    labs(
      title = "LR Distributions for Relationship Comparisons",
      subtitle = paste0("(n=", n, " related/unrelated pairs)"),
      x = "Tested Relationship Type",
      y = "LR",
      fill = "Population"
    ) +
    #geom_text(data = ann_text1, aes(y = LR), label = "*", size = 10) +
    theme_minimal(base_size = 24) +
    scale_fill_manual(values = color_palette_race) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1)
    ) +
    scale_y_log10()+
    coord_cartesian(ylim=c(1e-15,NA))

```
Plots showing the distribution of likelihood ratios relative to the cutoffs for each hypothesis type (cousins, full siblings, half siblings, parent_child, and second cousins).
Here, we use the following cutoffs: false positive rate (FPR) of 0.01%, 0.1%, 1%, and a fixed cutoff of 1.
Values that are not present in the plot correspond to a likelihood ratio of 0.
```{r lr_distributions, results =FALSE,warning=FALSE,message=FALSE, echo = FALSE,fig.width = 20,fig.height = 15}
# color palettte
my_colors = colorRampPalette(c("#FFB000","#F77A2E","#DE3A8A","#7253FF","#5E8BFF"))(6)

hypotheses = c("parent_child","full_siblings","cousins","half_siblings","second_cousins")
plot_list = c()
i=1
for(hypothesis in hypotheses){
  cutoffs <- calculate_cutoffs(combined_lrs, c(1, 0.1, 0.01), hypothesis = hypothesis)
  combined_lrs_hyp_cutoffs = merge(combined_lrs, cutoffs, by = c("population","loci_set"), all.x = TRUE)
  combined_lrs_hyp_cutoffs$relationship_known_factor <- factor(combined_lrs_hyp_cutoffs$relationship_known, levels=c("parent_child", "full_siblings", "half_siblings", "cousins", "second_cousins", "unrelated"))
  
  combined_lrs_hyp_cutoffs$loci_set_factor <- factor(combined_lrs_hyp_cutoffs$loci_set, levels=c("core_13", "identifiler_15", "expanded_20", "supplementary", "autosomal_29"))
  
  combined_lrs_hyp_cutoffs = combined_lrs_hyp_cutoffs[which(relationship_tested == hypothesis),]

  plot1 = ggplot(data = combined_lrs_hyp_cutoffs) +
      geom_density(aes(x = log(LR), fill =relationship_known_factor),alpha = 0.5) + 
      geom_vline(aes(xintercept = log(fixed_cutoff), col = "Fixed cutoff (1)"),linetype = "dashed") +
      geom_vline(aes(xintercept = log(cutoff_0_1), col = "0.1% FPR"),linetype = "dashed") +
      geom_vline(aes(xintercept = log(cutoff_0_01), col = "0.01% FPR"),linetype = "dashed") +
      geom_vline(aes(xintercept = log(cutoff_1), col = "1% FPR"),linetype = "dashed") +
      ggtitle(paste0("Log Likelihood Ratios for the ", hypothesis,  " hypothesis")) +
      scale_fill_manual(values = my_colors, labels = c("parent_child","full_siblings","half_siblings","cousins","second_cousins","unrelated")) +
      scale_color_manual(name = paste0("Hypothesis: ",hypothesis), labels = c("0.1% FPR", "0.01% FPR","1% FPR","Fixed cutoff (1)"), values = c("red","blue","forestgreen","black")) +
      facet_grid(loci_set_factor~population_label, scales = "free") +
      theme(axis.text = element_text(size = 28), axis.title = element_text(size = 28)) +
      guides(fill=guide_legend(title="Known Relationship")) +
      xlab("Log Likelihood Ratio (LLR)") + ylab("Frequency") +
      theme_minimal(base_size = 24) 

  plot2 = ggplot(data = combined_lrs_hyp_cutoffs) +
      geom_histogram(aes(x = log(LR), fill =relationship_known_factor),position="identity", alpha = 0.5) + 
      geom_vline(aes(xintercept = log(fixed_cutoff), col = "Fixed cutoff (1)"),linetype = "dashed") +
      geom_vline(aes(xintercept = log(cutoff_0_1), col = "0.1% FPR"),linetype = "dashed") +
      geom_vline(aes(xintercept = log(cutoff_0_01), col = "0.01% FPR"),linetype = "dashed") +
      geom_vline(aes(xintercept = log(cutoff_1), col = "1% FPR"),linetype = "dashed") +
      ggtitle(paste0("Log Likelihood Ratios for the ", hypothesis,  " hypothesis")) +
      scale_fill_manual(values = my_colors, labels = c("parent_child","full_siblings","half_siblings","cousins","second_cousins","unrelated")) +
      scale_color_manual(name = paste0("Hypothesis: ",hypothesis), labels = c("0.1% FPR", "0.01% FPR","1% FPR","Fixed cutoff (1)"), values = c("red","blue","forestgreen","black")) +
      facet_grid(loci_set_factor~population_label, scales = "free") +
      theme(axis.text = element_text(size = 28), axis.title = element_text(size = 28)) +
      guides(fill=guide_legend(title="Known Relationship")) +
      xlab("Log Likelihood Ratio (LLR)") + ylab("Frequency") +
      theme_minimal(base_size = 24)


  plot_list[[i]] = plot1
  plot_list[[i+1]] = plot2

  i = i+2

}

plot_list[[1]]
plot_list[[2]]

plot_list[[3]]
plot_list[[4]]

plot_list[[5]]
plot_list[[6]]

plot_list[[7]]
plot_list[[8]]

plot_list[[9]]
plot_list[[10]]
```

Below, for each hypothesis (parent_child, full siblings, half siblings, cousins, and second cousins), 
we plot the proportion of pairs with a likelihood ratio exceeding the cutoffs for a false positive rate (FPR) of
0.01%, 0.1%, 1% and fixed cutoff of 1. Results are faceted by loci set and known relationship type on the x-axis.

```{r cutoffs_results =FALSE,warning=FALSE,message=FALSE, echo = FALSE, fig.width = 20,fig.height = 15}
hypotheses = c("parent_child","full_siblings","cousins","half_siblings","second_cousins")
hypotheses_label = c("parent_child","Full Siblings","Cousins","Half Siblings","Second Cousins")

plot_list = c()
i=1

for(hypothesis in hypotheses){
  cutoffs <- calculate_cutoffs(combined_lrs, c(1, 0.1, 0.01), hypothesis = hypothesis)
  proportion_exceeding = calculate_proportions_exceeding_cutoffs(input_df = combined_lrs, cutoffs = cutoffs, hypothesis = hypothesis) 

  # race labels (can be changed)
  proportion_exceeding <- proportion_exceeding %>%
                     mutate(population_label = case_when(population == "AfAm" ~ "African-American",
                                                         population == "Asian" ~ "Asian",
                                                         population == "Cauc" ~ "Caucasian",
                                                         population == "Hispanic" ~ "Hispanic"))

  proportions_exceeding_cutoffs_long = proportion_exceeding %>%
    pivot_longer(
      cols = c(proportion_exceeding_1,proportion_exceeding_0_1,proportion_exceeding_0_01,proportion_exceeding_fixed),
     names_to = c("Cutoff"),
     values_to = "Value")

  proportions_exceeding_cutoffs_long$loci_set_factor <- factor(proportions_exceeding_cutoffs_long$loci_set, levels=c("core_13", "identifiler_15", "expanded_20", "supplementary", "autosomal_29"))

  proportions_exceeding_cutoffs_long = proportions_exceeding_cutoffs_long %>%
    mutate(relationship_tested_label = case_when(relationship_tested == "cousins" ~ "Cousins",
                                                 relationship_tested == "full_siblings" ~ "Full Siblings",
                                                 relationship_tested == "half_siblings" ~ "Half Siblings",
                                                relationship_tested == "parent_child" ~ "Parent-Child",
                                                relationship_tested == "second_cousins" ~ "Second Cousins"))

  proportions_exceeding_cutoffs_long = proportions_exceeding_cutoffs_long %>%
    mutate(cutoff_label = case_when(Cutoff == "proportion_exceeding_0_01" ~ "0.01% FPR",
                                   Cutoff == "proportion_exceeding_0_1" ~ "0.1% FPR",
                                   Cutoff == "proportion_exceeding_1" ~ "1% FPR",
                                   Cutoff == "proportion_exceeding_fixed" ~ "Fixed (1)"))

  proportions_exceeding_cutoffs_long$loci_set_factor <- factor(proportions_exceeding_cutoffs_long$loci_set, levels=c("core_13", "identifiler_15", "expanded_20", "supplementary", "autosomal_29"))

  # make sure relationships are in the order we want
  proportions_exceeding_cutoffs_long$relationship_known_factor <- factor(proportions_exceeding_cutoffs_long$relationship_known, levels=c("parent_child", "full_siblings", "half_siblings", "cousins", "second_cousins", "unrelated"))

  #proportions_exceeding_cutoffs_long_parent_child = proportions_exceeding_cutoffs_long[proportions_exceeding_cutoffs_long$relationship_tested == "parent_child",]

  plot1 = ggplot(proportions_exceeding_cutoffs_long, aes(x = relationship_known_factor, y = Value, fill = population_label, color = population_label)) +
    geom_bar(stat = "identity", position = position_dodge(width = 0.9)) +
    facet_grid(cutoff_label ~ loci_set_factor, scales = "fixed") +
    labs(
      title = "Proportions Exceeding Likelihood Cut-offs",
      subtitle = paste0(hypotheses_label[i], " hypothesis"),
      x = "Relationship Known",
      y = "Proportion Exceeding Cut-off",
      fill = "Population",
      color = "Population"
    ) +
    scale_color_manual(values = color_palette_race) +
    scale_fill_manual(values = color_palette_race) +
    theme_minimal(base_size = 24) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1)
    )

    plot_list[[i]] = plot1
    i = i+1
}

plot_list[[1]]
plot_list[[2]]
plot_list[[3]]
plot_list[[4]]
plot_list[[5]]

```
