---
title: "Likelihood Ratio Distributions Mar 25"
date: "10 March, 2025"
site: workflowr::wflow_site
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(dplyr)
library(data.table)
library(ggplot2)
library(tidyr)
#library(DBI)

# Define parameters
relationships <- c("parent_child", "full_siblings", "half_siblings", "cousins", "second_cousins", "unrelated")
populations <- c("AfAm", "Asian", "Cauc", "Hispanic")
loci_sets <- c("core_13", "identifiler_15", "expanded_20", "supplementary", "autosomal_29")


#Set up Bigquery connection and recover input table:
#bq_auth() #You will need to be signed into google and have permissions to access
#2 #Enter either 1 or 2 here in response to the prompt to authenticate your ID

#project_id<<-"Umbizo"
#con <- dbConnect(bigrquery::bigquery(), project = project_id)

#input<-paste0(project_id,".LasisiSTR.summary_genotypes")
#eval(parse(text=paste0("SQL<-\"SELECT * FROM `",input,"`")))

#Other SQL options can be used to filter here, e.g. WHERE x LIKE "Y" and WHERE x IS NOT NULL etc

#con <- dbConnect(
 # bigrquery::bigquery(),
 # project = project_id,
  #billing = project_id,
#)

#combined_lrs<-as.data.table(data.frame(dbGetQuery(con,SQL)))
combined_lrs<-readRDS("data/sim_summary_genotypes.rds")
combined_lrs$population_label<-NA
combined_lrs$population_label<-ifelse(combined_lrs$population=="AfAm","African-American",combined_lrs$population_label)
combined_lrs$population_label<-ifelse(combined_lrs$population=="Asian","Asian",combined_lrs$population_label)
combined_lrs$population_label<-ifelse(combined_lrs$population=="Cauc","Caucasian",combined_lrs$population_label)
combined_lrs$population_label<-ifelse(combined_lrs$population=="Hispanic","Hispanic",combined_lrs$population_label)
# Add population labels
#combined_lrs <- combined_lrs %>%
 # mutate(population_label = case_when( #This syntax seems to cause a problem within workflowr only
  #  population == "AfAm" ~ "African-American",
   # population == "Asian" ~ "Asian",
    #population == "Cauc" ~ "Caucasian",
#    population == "Hispanic" ~ "Hispanic"
 # ))

# Set factor levels
combined_lrs$loci_set_factor <- factor(combined_lrs$loci_set, 
                                    levels = c("core_13", "identifiler_15", "expanded_20", "supplementary", "autosomal_29"))
combined_lrs$relationship_known_factor <- factor(combined_lrs$relationship_known, 
                                             levels = relationships)
```

### Plot 0: Mean LR Across Populations and Relationship Types

Mean LR by population, loci set and relationship type, assuming no errors in assigning relationship or population.

```{r plot0, results=FALSE, warning=FALSE, message=FALSE, echo=FALSE, fig.width=20, fig.height=15}
color_palette_race = c("#AA4499", "#DDCC77", "#88CCEE", "#117733")

# Calculate summary statistics
summary_stats<-combined_lrs[combined_lrs$population_known==combined_lrs$population_tested & combined_lrs$relationship_tested==combined_lrs$relationship_known,]

#summary_stats <- summary_stats[,
#  list(
 #   mean_LR = mean(LR),
  #  lower_95 = quantile(LR, 0.025),
   # upper_95 = quantile(LR, 0.975)
#  ),
 # by = c("relationship_tested", "population_label", "loci_set_factor")
#]

summary_stats <- summary_stats %>%
  group_by("relationship_tested", "population_label", "loci_set_factor") %>%
  mutate(mean_LR = mean(LR),lower_95 = quantile(LR, 0.025), upper_95 = quantile(LR, 0.975)) %>%
  ungroup()


#  list(
 #   mean_LR = mean(LR),
  #  lower_95 = quantile(LR, 0.025),
   # upper_95 = quantile(LR, 0.975)
#  ),
 # by = c("relationship_tested", "population_label", "loci_set_factor")
#]
# Example data
summary_stats$relationship_tested <- factor(summary_stats$relationship_tested, levels = relationships)

plot0 <- ggplot(summary_stats, 
       aes(x = loci_set_factor, y = mean_LR, 
           group = population_label, color = population_label)) +
  geom_line(linewidth = 2) +
  facet_wrap(~ relationship_tested, ncol = 2) +
 # scale_y_log10() +
  labs(
    title = "Mean LR Across Populations and Relationship Types",
    x = "Loci Set",
    y = "Combined LR",
    color = "Population"
  ) +
  scale_color_manual(values = color_palette_race) +
  theme_minimal(base_size = 24) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "bottom"
  )

plot0
```

### Plot A: LR Distributions Across Populations

LR distributions shown by population, loci set and relationship type, assuming no errors in assigning relationship or population.

```{r plota, results=FALSE, warning=FALSE, message=FALSE, echo=FALSE, fig.width=20, fig.height=15}

summary_stats<-combined_lrs[combined_lrs$population_known==combined_lrs$population_tested & combined_lrs$relationship_tested==combined_lrs$relationship_known,]

summary_stats$relationship_tested <- factor(summary_stats$relationship_tested, levels = relationships)


plota <- ggplot(summary_stats, 
       aes(x = relationship_tested, y = LR, 
           fill = population_label, color = population_label)) +
  geom_boxplot(position = position_dodge(width = 0.9)) +
  facet_wrap(~ loci_set_factor, scales = "fixed") +
  labs(
    title = "LR Distributions Across Populations and Relationship Types",
    x = "Relationship Tested",
    y = "LR",
    fill = "Population",
    color = "Population"
  ) +
  scale_color_manual(values = color_palette_race) +
  scale_fill_manual(values = color_palette_race) +
  theme_minimal(base_size = 24) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) +
  scale_y_log10()

plota
```

### Plot B: LR Distributions for Relationship Comparisons

LR distributions, comparing known and tested relationship values, assuming no errors in assigning population.

```{r plotb, results=FALSE, warning=FALSE, message=FALSE, echo=FALSE, fig.width=20, fig.height=15}
n <- nrow(combined_lrs[which(relationship_known == "unrelated" & 
                            relationship_tested == "parent_child"),])

summary_stats$relationship_known_factor <- factor(summary_stats$relationship_known_factor, levels = relationships)
summary_stats$relationship_tested <- factor(summary_stats$relationship_tested, levels = relationships)

plotb <- ggplot(combined_lrs[combined_lrs$population_known==combined_lrs$population_tested,], 
       aes(x = relationship_tested, y = LR, fill = population_label)) +
  geom_boxplot(position = position_dodge(width = 0.9)) +
  facet_grid(relationship_known_factor ~ loci_set_factor, scales = "free_y") +
  labs(
    title = "LR Distributions for Relationship Comparisons",
    subtitle = paste0("(n=", n, " related/unrelated pairs)"),
    x = "Tested Relationship Type",
    y = "LR",
    fill = "Population"
  ) +
  theme_minimal(base_size = 24) +
  scale_fill_manual(values = color_palette_race) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) +
  scale_y_log10()

plotb
```

### Plots 1 & 2: Log Likelihood Ratio Distributions

Plots showing the distribution of likelihood ratios relative to the cutoffs for each hypothesis type (cousins, full siblings, half siblings, parent_child, and second cousins). Here, we use the following cutoffs: false positive rate (FPR) of 0.01%, 0.1%, 1%, and a fixed cutoff of 1. Values that are not present in the plot correspond to a likelihood ratio of 0.

```{r plot1_and_2, results=FALSE, warning=FALSE, message=FALSE, echo=FALSE, fig.width=20, fig.height=15}
# Function to calculate cut-off values
calculate_cutoffs <- function(input_df, fp_rates, hypothesis) {
  input_df <- input_df[input_df$relationship_known == "unrelated" & 
                      input_df$relationship_tested == hypothesis & input_df$population_tested== input_df$population_known,]
  cutoffs <- input_df[,
    list(
      fixed_cutoff = 1.00,
      cutoff_1 = quantile(LR, probs = 1 - fp_rates[1] / 100, na.rm = TRUE),
      cutoff_0_1 = quantile(LR, probs = 1 - fp_rates[2] / 100, na.rm = TRUE),
      cutoff_0_01 = quantile(LR, probs = 1 - fp_rates[3] / 100, na.rm = TRUE),
      n_unrelated = .N
    ),
    by = c("population", "loci_set")]
  return(cutoffs)
}

# Color palette for relationships
my_colors = colorRampPalette(c("#FFB000","#F77A2E","#DE3A8A","#7253FF","#5E8BFF"))(6)

# Example for parent-child hypothesis
hypothesis <- "parent_child"
cutoffs <- calculate_cutoffs(combined_lrs, c(1, 0.1, 0.01), hypothesis)
combined_lrs_hyp_cutoffs <- merge(combined_lrs, cutoffs, 
                                 by = c("population", "loci_set"), all.x = TRUE)

combined_lrs_hyp_cutoffs$relationship_known_factor<- factor(combined_lrs_hyp_cutoffs$relationship_known_factor, levels = relationships)


plot1 <- ggplot(data = combined_lrs_hyp_cutoffs[which(relationship_tested == hypothesis & population_tested== population_known),]) +
  geom_density(aes(x = log(LR), fill = relationship_known_factor), alpha = 0.5) +
  geom_vline(aes(xintercept = log(fixed_cutoff), col = "Fixed cutoff (1)"), 
             linetype = "dashed") +
  geom_vline(aes(xintercept = log(cutoff_0_1), col = "0.1% FPR"), 
             linetype = "dashed") +
  geom_vline(aes(xintercept = log(cutoff_0_01), col = "0.01% FPR"), 
             linetype = "dashed") +
  geom_vline(aes(xintercept = log(cutoff_1), col = "1% FPR"), 
             linetype = "dashed") +
  ggtitle(paste0("Log Likelihood Ratios for the ", hypothesis, " hypothesis")) +
  scale_fill_manual(values = my_colors) +
  scale_color_manual(name = paste0("Hypothesis: ", hypothesis),
                    values = c("red", "blue", "forestgreen", "black")) +
  facet_grid(loci_set_factor ~ population_label, scales = "free") +
  theme_minimal(base_size = 24) +
  guides(fill = guide_legend(title = "Known Relationship")) +
  xlab("Log Likelihood Ratio (LLR)") + 
  ylab("Frequency")

plot1
```

### Final Plot: Proportions Exceeding Cutoffs

Below, for each hypothesis (parent_child, full siblings, half siblings, cousins, and second cousins), we plot the proportion of pairs with a likelihood ratio exceeding the cutoffs for a false positive rate (FPR) of 0.01%, 0.1%, 1% and fixed cutoff of 1. Results are faceted by loci set and known relationship type on the x-axis.

```{r final_plot, results=FALSE, warning=FALSE, message=FALSE, echo=FALSE, fig.width=20, fig.height=15}
calculate_proportions_exceeding_cutoffs <- function(input_df, cutoffs, hypothesis) {
  input_df <- input_df[input_df$relationship_tested == hypothesis & input_df$population_tested==input_df$population_known,]
  df_with_cutoffs <- merge(input_df, cutoffs, by = c("population", "loci_set"))
  
  df_with_cutoffs <- df_with_cutoffs[,
    list(
      population = population,
      relationship_tested = relationship_tested,
      relationship_known = relationship_known,
      loci_set = loci_set,
      exceeds_fixed_cutoff = LR > fixed_cutoff,
      exceeds_cutoff_1 = LR > cutoff_1,
      exceeds_cutoff_0_1 = LR > cutoff_0_1,
      exceeds_cutoff_0_01 = LR > cutoff_0_01
    )]
  
  proportions_exceeding <- df_with_cutoffs[,
    list(
      proportion_exceeding_fixed = sum(exceeds_fixed_cutoff) / .N,
      proportion_exceeding_1 = sum(exceeds_cutoff_1, na.rm = TRUE) / .N,
      proportion_exceeding_0_1 = sum(exceeds_cutoff_0_1, na.rm = TRUE) / .N,
      proportion_exceeding_0_01 = sum(exceeds_cutoff_0_01, na.rm = TRUE) / .N,
      n_related = .N
    ),
    by = c("population", "relationship_tested", "loci_set", "relationship_known")]
  
  return(proportions_exceeding)
}

# Generate final plot for parent-child hypothesis
proportion_exceeding <- calculate_proportions_exceeding_cutoffs(
  input_df = combined_lrs, 
  cutoffs = cutoffs,
  hypothesis = "parent_child"
)

proportion_exceeding <- proportion_exceeding %>%
  mutate(population_label = case_when(
    population == "AfAm" ~ "African-American",
    population == "Asian" ~ "Asian",
    population == "Cauc" ~ "Caucasian",
    population == "Hispanic" ~ "Hispanic"
  ))

proportions_long <- proportion_exceeding %>%
  pivot_longer(
    cols = starts_with("proportion_exceeding"),
    names_to = "Cutoff",
    values_to = "Value"
  ) %>%
  mutate(
    cutoff_label = case_when(
      Cutoff == "proportion_exceeding_0_01" ~ "0.01% FPR",
      Cutoff == "proportion_exceeding_0_1" ~ "0.1% FPR",
      Cutoff == "proportion_exceeding_1" ~ "1% FPR",
      Cutoff == "proportion_exceeding_fixed" ~ "Fixed (1)"
    )
  )

final_plot <- ggplot(proportions_long, 
       aes(x = relationship_known, y = Value, 
           fill = population_label, color = population_label)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.9)) +
  facet_grid(cutoff_label ~ loci_set, scales = "fixed") +
  labs(
    title = "Proportions Exceeding Likelihood Cut-offs",
    subtitle = "Parent-Child hypothesis",
    x = "Relationship Known",
    y = "Proportion Exceeding Cut-off",
    fill = "Population",
    color = "Population"
  ) +
  scale_color_manual(values = color_palette_race) +
  scale_fill_manual(values = color_palette_race) +
  theme_minimal(base_size = 24) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )


final_plot
```
