---
title: "Murphy Data Cleaning"
author: "Stella BooydeGraaff"
date: "`r format(Sys.time(), '%Y-%m-%d %H:%M:%S')`"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE, echo=FALSE}

knitr::opts_chunk$set(eval = TRUE, echo = FALSE, warning = FALSE, fig.width = 7, fig.height = 6)

if (!requireNamespace("tidyverse", quietly = TRUE)) {
  install.packages("tidyverse")
}

if (!requireNamespace("knitr", quietly = TRUE)) {
  install.packages("knitr")
}

if (!requireNamespace("patchwork", quietly = TRUE)) {
  install.packages("patchwork")
}


library(tidyverse)
library(knitr)
library(patchwork)
```

```{r}
data <- read_csv("data/Murphy_Appendix_all_data.csv", show_col_types = FALSE) 

```

```{r} 
data <- data |>
  select(-file) 
```

```{r}
  data <- data |> 
      mutate(
        # Check if 'proportion' is in the variable_detailed
        is_proportion = grepl("proportion", variable_detailed, ignore.case = TRUE),
         
        # Create value_type based on the presence of 'proportion'
        value_type = ifelse(is_proportion, 
                          "proportion", 
                          "number"),
          
          # Remove 'proportion' text from variable_detailed
          variable_detailed = ifelse(is_proportion, gsub("proportion", "", variable_detailed, ignore.case = TRUE), variable_detailed),
          
          # Trim whitespace
          variable_detailed = trimws(variable_detailed),
          
          # cleaned variable_detailed to numeric
          variable_detailed_numeric = suppressWarnings(as.numeric(variable_detailed)),
          
          # Update value_type to "number" if variable_detailed_numeric is greater than 1
          value_type = ifelse(!is.na(variable_detailed_numeric) & variable_detailed_numeric > 1, 
                              "number", 
                              value_type)
      ) |> 
      # Drop temporary columns
      select(-is_proportion, -variable_detailed_numeric)

```

```{r}
data <- data |> 
    rename(variable_type = variatble_type)

```

```{r}
#race
data <- data |> 
  mutate(
    variable_detailed = case_when(
      variable_type == "race" & grepl("\\basian\\b", variable_detailed, ignore.case = TRUE) ~ "asian_total",
      variable_type == "race" & grepl("\\bblack\\b", variable_detailed, ignore.case = TRUE) ~ "black_total",
      variable_type == "race" & grepl("\\bafrican_american\\b", variable_detailed, ignore.case = TRUE) ~ "black_total",
      variable_type == "race" & grepl("\\bblack_\\b", variable_detailed, ignore.case = TRUE) ~ "black_total",
      variable_type == "race" & grepl("\\bhispanic\\b", variable_detailed, ignore.case = TRUE) ~ "hispanic_total",
      variable_type == "race" & grepl("\\bhispanic_\\b", variable_detailed, ignore.case = TRUE) ~ "hispanic_total",
      variable_type == "race" & grepl("\\bnative\\b", variable_detailed, ignore.case = TRUE) ~ "native_american_total",
      variable_type == "race" & grepl("\\bamerican_indian\\b", variable_detailed, ignore.case = TRUE) ~ "native_american_total",
       variable_type == "race" & grepl("\\bnative_american\\b", variable_detailed, ignore.case = TRUE) ~ "native_american_total",
      variable_type == "race" & grepl("\\bwhite\\b", variable_detailed, ignore.case = TRUE) ~ "white_total",
      variable_type == "race" & grepl("\\bcausasian\\b", variable_detailed, ignore.case = TRUE) ~ "white_total",
      variable_type == "race" & grepl("\\bcaucasian\\b", variable_detailed, ignore.case = TRUE) ~ "white_total",
      variable_type == "race" & grepl("\\bcausasian_\\b", variable_detailed, ignore.case = TRUE) ~ "white_total",
      variable_type == "race" & grepl("\\bother\\b", variable_detailed, ignore.case = TRUE) ~ "other_total",
      variable_type == "race" & grepl("\\basian_offender\\b", variable_detailed, ignore.case = TRUE) ~ "asian_convicted_offender",
      variable_type == "race" & grepl("\\bblack_offender\\b", variable_detailed, ignore.case = TRUE) ~ "black_convicted_offender",
      variable_type == "race" & grepl("\\bhispanic_offender\\b", variable_detailed, ignore.case = TRUE) ~ "hispanic_convicted_offender",
      variable_type == "race" & grepl("\\bnative_american_offender\\b", variable_detailed, ignore.case = TRUE) ~ "native_american_convicted_offender",
      variable_type == "race" & grepl("\\bwhite_offender\\b", variable_detailed, ignore.case = TRUE) ~ "white_convicted_offender",
      variable_type == "race" & grepl("\\bcaucasian_offender\\b", variable_detailed, ignore.case = TRUE) ~ "white_convicted_offender",
      variable_type == "race" & grepl("\\bcaucasian_convicted_offender\\b", variable_detailed, ignore.case = TRUE) ~ "white_convicted_offender",
      variable_type == "race" & grepl("\\bother_offender\\b", variable_detailed, ignore.case = TRUE) ~ "other_convicted_offender",
      variable_type == "race" & grepl("\\bafrican_american__convicted_offender\\b", variable_detailed, ignore.case = TRUE) ~ "black_convicted_offender",
      variable_type == "race" & grepl("\\bafrican_american_convicted_offender\\b", variable_detailed, ignore.case = TRUE) ~ "black_convicted_offender",
      variable_type == "race" & grepl("\\bafrican_american_offender\\b", variable_detailed, ignore.case = TRUE) ~ "black_convicted_offender",
      variable_type == "race" & grepl("\\bafrican_american_arrestee\\b", variable_detailed, ignore.case = TRUE) ~ "black_arrestee",
      variable_type == "race" & grepl("\\bcaucasian_arrestee\\b", variable_detailed, ignore.case = TRUE) ~ "white_arrestee",
      TRUE ~ variable_detailed
    ))
    
#DNA collection breakdown
data <- data |> 
  mutate(
    variable_detailed = case_when(
      variable_type == "DNA_collection_breakdown" & grepl("asian", variable_detailed, ignore.case = TRUE) ~ "asian_annual",
      variable_type == "DNA_collection_breakdown" & grepl("black", variable_detailed, ignore.case = TRUE) ~ "black_annual",
      variable_type == "DNA_collection_breakdown" & grepl("hispanic", variable_detailed, ignore.case = TRUE) ~ "hispanic_annual",
      variable_type == "DNA_collection_breakdown" & grepl("native", variable_detailed, ignore.case = TRUE) ~ "native_american_annual",
      variable_type == "DNA_collection_breakdown" & grepl("white", variable_detailed, ignore.case = TRUE) ~ "white_annual",
      variable_type == "DNA_collection_breakdown" & grepl("other", variable_detailed, ignore.case = TRUE) ~ "other_annual",
      TRUE ~ variable_detailed
    )
  )
    
```

```{r}
# Update 'variable_type' to 'race' where it is 'DNA_collection_breakdown'

data <- data |> 
  mutate(
    variable_type = ifelse(variable_type == "DNA_collection_breakdown", "race", variable_type))
```

```{r}
#state demographic
data <- data |> 
  mutate(
    variable_detailed = case_when(
      variable_type == "state_demographic" & grepl("asian", variable_detailed, ignore.case = TRUE) ~ "asian_state",
      variable_type == "state_demographic" & grepl("black", variable_detailed, ignore.case = TRUE) ~ "black_state",
      variable_type == "state_demographic" & grepl("hispanic", variable_detailed, ignore.case = TRUE) ~ "hispanic_state",
      variable_type == "state_demographic" & grepl("native", variable_detailed, ignore.case = TRUE) ~ "native_american_state",
      variable_type == "state_demographic" & grepl("white", variable_detailed, ignore.case = TRUE) ~ "white_state",
      variable_type == "state_demographic" & grepl("other", variable_detailed, ignore.case = TRUE) ~ "other_state",
      TRUE ~ variable_detailed
    )
  )

```

```{r}
# Update 'variable_type' to 'race' where it is 'state_demographic'

data <- data |> 
  mutate(
    variable_type = ifelse(variable_type == "state_demographic", "race", variable_type))
```

```{r}
#population data collected variables
data <- data |> 
  mutate(
    variable_detailed = case_when(
      variable_type == "population_data_collected" & grepl("asian", variable_detailed, ignore.case = TRUE) ~ "asian_pop_collected",
      variable_type == "population_data_collected" & grepl("black", variable_detailed, ignore.case = TRUE) ~ "black_pop_collected",
      variable_type == "population_data_collected" & grepl("hispanic", variable_detailed, ignore.case = TRUE) ~ "hispanic_pop_collected",
      variable_type == "population_data_collected" & grepl("native", variable_detailed, ignore.case = TRUE) ~ "native_american_pop_collected",
      variable_type == "population_data_collected" & grepl("white", variable_detailed, ignore.case = TRUE) ~ "white_pop_collected",
      variable_type == "population_data_collected" & grepl("other", variable_detailed, ignore.case = TRUE) ~ "other_pop_collected",
      TRUE ~ variable_detailed
    )
  )
```

```{r}
# Update 'variable_type' to 'race' where it is 'population_data_collected'

data <- data |> 
  mutate(
    variable_type = ifelse(variable_type == "population_data_collected", "race", variable_type))
```

```{r}
#sex variable type
data <- data |> 
  mutate(
    variable_detailed = case_when(
      variable_type == "sex" & grepl("\\bfemale\\b", variable_detailed, ignore.case = TRUE) ~ "female_total",
      variable_type == "sex" & grepl("\\bfemale_\\b", variable_detailed, ignore.case = TRUE) ~ "female_total",
      variable_type == "sex" & grepl("\\bfemales\\b", variable_detailed, ignore.case = TRUE) ~ "female_total",
      variable_type == "sex" & grepl("\\bfemale_offender\\b", variable_detailed, ignore.case = TRUE) ~ "female_convicted_offender",
      variable_type == "sex" & grepl("\\bmale\\b", variable_detailed, ignore.case = TRUE) ~ "male_total",
      variable_type == "sex" & grepl("\\bmale_\\b", variable_detailed, ignore.case = TRUE) ~ "male_total",
      variable_type == "sex" & grepl("\\bmales\\b", variable_detailed, ignore.case = TRUE) ~ "male_total",
      TRUE ~ variable_detailed 
    )
  )
```

```{r}
data <- data |> 
  mutate(
    variable_type = ifelse(variable_type == "gender", "sex", variable_type))
```

```{r}
data <- data |> 
  mutate(
    variable_detailed = case_when(
      variable_type == "offender_type" & grepl("\\barrestee\\b", variable_detailed, ignore.case = TRUE) ~ "arrestee_total",
      variable_type == "offender_type" & grepl("\\barrested_offender\\b", variable_detailed, ignore.case = TRUE) ~ "arrestee_total",
      variable_type == "offender_type" & grepl("\\bconvicted_offender\\b", variable_detailed, ignore.case = TRUE) ~ "convicted_offender_total",
      variable_type == "offender_type" & grepl("\\bconviceted_offender\\b", variable_detailed, ignore.case = TRUE) ~ "convicted_offender_total",
      variable_type == "offender_type" & grepl("\\boffender\\b", variable_detailed, ignore.case = TRUE) ~ "convicted_offender_total",
      TRUE ~ variable_detailed 
    )
  )
```

```{r}
write.csv(data, "data/murphy_foia_cleaned.csv")
```

# Visualizations of State DNA Databases
Below is a small sample of the data set used to create these visualizations.

```{r}
data_sample <- data |> sample_n(5)
kable(data_sample, caption = "Sample of the Full Data")
```

## Data on Sex

```{r}
#creating gender database
data_gender <- data |> 
  filter(variable_type == "sex")
```

```{r}

#California totals
california_totals <- data_gender |> 
  filter(State == "California") |> 
  summarise(
    male_total = sum(value[grepl("male", variable_detailed, ignore.case = TRUE)]),
    female_total = sum(value[grepl("female", variable_detailed, ignore.case = TRUE)]),
    unknown_total = sum(value[grepl("unknown", variable_detailed, ignore.case = TRUE)])
  ) 

california_totals <- california_totals |>  
  pivot_longer(cols = everything(), names_to = "variable_detailed", values_to = "value")|> 
  mutate(
    State = "California",
    variable_type = "sex",
    value_type = "proportion"
  )
```

The following bar chart shows the gender breakdown of the state DNA databases.

```{r}
#create dataset with gender values as proportions only
data_gender_proportions <- data_gender  |> 
  group_by(State, variable_type, value_type) |>   # group by columns
  mutate(proportion = value / sum(value)) |>      # calculate proportion
  ungroup() |>                                    # Ungroup
  select(-value, -value_type)|>                   # remove value and value_type
  rename(proportion_value = proportion)           # Rename new column

data_gender_simplified <- data_gender_proportions |> 
  mutate(
 simplified_sex = case_when(
      str_detect(variable_detailed, "\\bmale\\b") ~ "male",
      str_detect(variable_detailed, "\\bmale_total\\b") ~ "male",
      str_detect(variable_detailed, "\\bmale_convicted_offender\\b") ~ "male",
      str_detect(variable_detailed, "\\bmale_arrestee\\b") ~ "male",
      str_detect(variable_detailed, "\\bfemale_total\\b") ~ "female",
      str_detect(variable_detailed, "\\bfemale_convicted_offender\\b") ~ "female",
      str_detect(variable_detailed, "\\bfemale_offender\\b") ~ "female",
      str_detect(variable_detailed, "\\bfemale_arrestee\\b") ~ "female",
      str_detect(variable_detailed, "\\bfemale\\b") ~ "female",
      TRUE ~ "unknown"
    )
  ) |> 
  group_by(State, simplified_sex) |> 
  summarize(total_value = sum(proportion_value), .groups = 'drop')

#stacked bar chart simplified
ggplot(data_gender_simplified, aes(x = State, y = total_value, fill = simplified_sex)) +
  geom_bar(stat = "identity") +  # Stacked bar chart by default
  scale_fill_manual(values = c("#489C92", "#F38C79", "#D8DCDA")) +  # color palette
  labs(y = "Proportion", x = "State", title = "Gender Distribution of DNA Database by State", fill = "Sex") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

## Data on Offender Type
The following pie chart shows a summary of the proportions of arrestees and convicted offenders in the state DNA databases.

```{r}
data_offender_type <- data |> 
  filter(variable_type == "offender_type")

#pie chart of offender type

# Rename the offender types
data_offender_type_renamed <- data_offender_type |> 
  mutate(variable_detailed = case_when(
    variable_detailed == "arrestee_total" ~ "Arrestee",
    variable_detailed == "convicted_offender_total" ~ "Convicted Offender",
    TRUE ~ variable_detailed
  ))

# Filter and group by offender type, then summarize
data_offender_type_summarized <- data_offender_type_renamed |> 
  filter(variable_detailed %in% c("Arrestee", "Convicted Offender")) |> 
  group_by(variable_detailed) |> 
  summarize(total_value = sum(value, na.rm = TRUE))

# Calculate total sum of the summarized data
full_value <- sum(data_offender_type_summarized$total_value, na.rm = TRUE)

# Calculate percentages and create labels
data_offender_type_summarized <- data_offender_type_summarized |> 
  mutate(percentage = total_value / full_value * 100,
         label = paste0(round(percentage, 1), "%"))

# Create the pie chart
ggplot(data_offender_type_summarized, aes(x = "", y = total_value, fill = variable_detailed)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar("y") +  # Converts bar chart to pie chart
  labs(title = "Distribution of Offender Types across State DNA Databases", caption = "Data from California, Indiana, Texas, and Nevada", fill = "Offender Type") +
  geom_text(aes(label = label), position = position_stack(vjust = 0.5)) +
  theme_minimal() +
  theme(axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_blank()) +
  scale_fill_manual(values = c("#59B8B2", "#F38C79"))

```

## Data on Race

### State Demographics
Below are some summary statistics of the racial demographics of each state's total population.

```{r}
#create data set with variable_type "race"
data_race <- data |> 
  filter(variable_type == "race")

#Summary stats
data_state <- data_race |> 
  filter(str_detect(variable_detailed, "_state"))

black_state <- data_state |> 
  filter(variable_detailed == "black_state") |>
  mutate(percentage = round(value * 100, 2))

# Top 3 states with the highest value for "black_state"
top_3_highest <- black_state |> 
  arrange(desc(percentage)) |> 
  slice_head(n = 3) |> 
  select(State, percentage)

# Top 3 states with the lowest value for "black_state"
top_3_lowest <- black_state |> 
  arrange(percentage) |> 
  slice_head(n = 3) |> 
  select(State, percentage)

# Average value for "black_state" across all states
average_black <- black_state |> 
  summarise(average_proportion = mean(value, na.rm = TRUE) * 100)

# Output the results

kable(top_3_highest, caption = "States with Highest Black Population", format = "markdown")

# Display the top 3 lowest states table
kable(top_3_lowest, caption = "States with Lowest Black Population", format = "markdown")
```

And below is the average proportion of Black people in a state and the average proportion of Black people in a state's DNA database.

```{r}
# Display the average value table
kable(average_black, caption = "Average Black Population Proportion Across All States", format = "markdown")


```

```{r}
#create a data set with totals combining "convicted_offender" and "arrestee" data into totals by race
convicted_arrestee_totals <- data_race |> 
  filter(str_detect(variable_detailed, "_convicted_offender|_arrestee")) |> 
  mutate(variable_detailed = str_replace(variable_detailed, "_convicted_offender|_arrestee", "_total")) |> 
  group_by(State, variable_detailed) |> 
  summarize(value = sum(value), .groups = 'drop') |>
  mutate(value_type = "number")

# Combine and process race_totals
race_totals <- data_race |> 
  mutate(variable_detailed = case_when(
    str_detect(variable_detailed, "other_total|unknown") ~ "other_or_unknown",
    TRUE ~ variable_detailed
  )) |> 
  filter(str_detect(variable_detailed, "_total") | str_detect(variable_detailed, "other_or_unknown"))

# Add convicted_arrestee_totals to race_totals
race_totals <- race_totals |> 
  bind_rows(convicted_arrestee_totals) |> 
  group_by(State, variable_detailed) |> 
  summarize(value = sum(value), value_type = first(value_type), .groups = 'drop') |> 
  ungroup()

# Calculate the proportion value within each state
race_totals <- race_totals |> 
  group_by(State) |> 
  mutate(proportion_value = ifelse(value_type == "number", value / sum(value), NA)) |> 
  ungroup()

#keep already calculated proportions and remove value_type column
race_totals <- race_totals |> 
  mutate(proportion_value = ifelse(is.na(proportion_value), value, proportion_value)) |> 
  select(-value_type)

race_totals <- race_totals |> 
  mutate(percentage = round(proportion_value * 100, 2))

# Replace race category names for clarity
race_totals <- race_totals |> 
  mutate(variable_detailed = case_when(
    variable_detailed == "white_total" ~ "White",
    variable_detailed == "black_total" ~ "Black",
    variable_detailed == "asian_total" ~ "Asian",
    variable_detailed == "hispanic_total" ~ "Hispanic",
    variable_detailed == "native_american_total" ~ "Native American",
    variable_detailed == "other_total" ~ "Other or Unknown",
    variable_detailed == "other_or_unknown" ~ "Other or Unknown",
    TRUE ~ variable_detailed
  ))

average_black_database <- race_totals |> 
  filter(variable_detailed == "Black") |> 
  summarise(average_proportion = mean(percentage, na.rm = TRUE))

#and calculate average proportion of black people in DNA databases (use race_totals)
kable(average_black_database, caption = "Average Black Proportion In DNA Databases Across All States", format = "markdown")
```

On average, Black Americans make up 10.96% of their state, but 20.93% of their state's DNA database, displaying the racial disparities in these databases.

### Race Totals
The following bar chart displays the proportions of each race in a state's DNA database.

```{r}


# Set the desired order for race categories
desired_order <- c("Other or Unknown", "Native American", "Hispanic", "Asian", "Black", "White")

# Create the bar plot
ggplot(race_totals |> 
       mutate(variable_detailed = factor(variable_detailed, levels = desired_order)),
       aes(x = State, y = proportion_value, fill = variable_detailed)) +
  geom_bar(stat = "identity") + 
  labs(title = "Race Totals in DNA Database by State",
       x = "State",
       y = "Proportion",
       fill = "Race Category") +
  scale_fill_manual(values = c("#BAD1DC", "#489C92","#59B8B2", "#91C8D0", "#05686D", "#F38C79")) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

Despite White Americans usually making up the majority of the population in the database, there is still a significant disparity present in this data. Comparing this to a graph of the racial breakdown of the total population in each of these states uncovers that Black Americans and other people of color are over represented in their state's database.


```{r}

#bar plot of state totals

selected_states <- c("California", "Florida", "Indiana", "Maine", "Nevada", "South_Dakota", "Texas")

select_states <- data_state |> 
  filter(State %in% selected_states) |> 
  mutate(variable_detailed = case_when(
    variable_detailed == "white_state" ~ "White",
    variable_detailed == "black_state" ~ "Black",
    variable_detailed == "asian_state" ~ "Asian",
    variable_detailed == "hispanic_state" ~ "Hispanic",
    variable_detailed == "native_american_state" ~ "Native American",
    TRUE ~ variable_detailed
  ))

desired_order <- c("Native American", "Hispanic", "Asian", "Black", "White")

ggplot(select_states |>
       mutate(variable_detailed = factor(variable_detailed, levels = desired_order)),
       aes(x = State, y = value, fill = variable_detailed)) +
  geom_bar(stat = "identity") +
  labs(title = "Proportion of Races by State", x = "State", y = "Proportion", fill = "Race Category") +
  scale_fill_manual(values = c("#489C92","#59B8B2", "#91C8D0", "#05686D", "#F38C79")) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

#side by side plots
desired_order <- c("Other or Unknown", "Native American", "Hispanic", "Asian", "Black", "White")

selected_states <- c("Texas")

plot1 <- ggplot(race_totals |> 
        filter(State %in% selected_states) |> 
       mutate(variable_detailed = factor(variable_detailed, levels = desired_order)),
       aes(x = State, y = proportion_value, fill = variable_detailed)) +
  geom_bar(stat = "identity") + 
  labs(title = "DNA Database",
       x = "State",
       y = "Proportion") +
  scale_fill_manual(values = c("#BAD1DC", "#489C92","#59B8B2", "#91C8D0", "#05686D", "#F38C79")) +
  theme_minimal() +
  theme(legend.position = "none") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

data_state <- data_race |> 
  filter(str_detect(variable_detailed, "_state"))

selected_states <- c("Texas")

select_states <- data_state |> 
  filter(State %in% selected_states) |> 
  mutate(variable_detailed = case_when(
    variable_detailed == "white_state" ~ "White",
    variable_detailed == "black_state" ~ "Black",
    variable_detailed == "asian_state" ~ "Asian",
    variable_detailed == "hispanic_state" ~ "Hispanic",
    variable_detailed == "native_american_state" ~ "Native American",
    TRUE ~ variable_detailed
  ))

desired_order <- c("Native American", "Hispanic", "Asian", "Black", "White")

plot2 <- ggplot(select_states |>
       mutate(variable_detailed = factor(variable_detailed, levels = desired_order)),
       aes(x = State, y = value, fill = variable_detailed)) +
  geom_bar(stat = "identity") +
  labs(title = "State Population", x = "State", y = "Proportion", fill = "Race Category") +
  scale_fill_manual(values = c("#489C92","#59B8B2", "#91C8D0", "#05686D", "#F38C79")) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
combined_plot <- plot1 + plot2 +
  theme(plot.margin = margin(10, 10, 10, 10)) +
  plot_annotation(title = "Racial Breakdown of DNA Database and Total Population in Texas") +
  plot_spacer()
combined_plot

```

As you can see, the proportion of Black people in the population of Texas is much smaller than the proportion of Black people in the Texas DNA database. Conversely, the proportion of White people in the DNA database is smaller than the percent of the state's population that is White. This pattern is true for most other states shown, highlighting the over-representation of Black Americans in the DNA databases.

### Annual DNA Collection
These summary statistics describe the racial breakdown of DNA collected annually by state. On average, the percentage of DNA collected annually that is from Black Americans is 29.33% and the percentage of DNA collected annually that is from White Americans is 56.3%.

```{r}
#Racial Breakdown of Annual DNA Collection - extremes and averages (states with highest collection and average for black people and white people)
data_annual <- data_race |> 
  filter(value_type == "proportion" & str_detect(variable_detailed, "_annual"))

# Filter for Black annual data
black_annual <- data_annual |> 
  filter(str_detect(variable_detailed, "black_annual"))

avg_black_annual <- black_annual |> 
  summarise(percentage = round(mean(value, na.rm = TRUE) * 100, 2))

#and calculate average proportion of black people with DNA collected annually
kable(avg_black_annual, caption = "Average Proportion of DNA collected Annually that is from Black Americans (Across All States)", format = "markdown")

white_annual <- data_annual |> 
  filter(str_detect(variable_detailed, "white_annual"))

avg_white_annual <- white_annual |> 
  summarise(percentage = round(mean(value, na.rm = TRUE) * 100, 2))

#and calculate average proportion of black people with DNA collected annually
kable(avg_white_annual, caption = "Average Proportion of DNA collected Annually from White Americans (Across All States)", format = "markdown")
```

These pie charts show the comparison of the proportion of races with DNA collected and the racial breakdown of the population across all states. We can clearly see from these charts the overwhelming over representation of Black Americans with DNA collected annually compared to the percentage of the population that they make up. The same thing is true for other people of color, and the opposite is true for White Americans, whose DNA is less likely to be collected.

```{r}
data_annual_pie <- data_annual |> 
  mutate(race_category = case_when(
    str_detect(variable_detailed, "black_annual") ~ "Black",
    str_detect(variable_detailed, "white_annual") ~ "White",
    TRUE ~ "Other")) |> 
    group_by(State, race_category) |> 
  summarise(total_value = sum(value, na.rm = TRUE), .groups = "keep")

summary_data_annual_pie <- data_annual_pie |> 
  group_by(race_category) |> 
  summarise(avg_data_annual = mean(total_value, na.rm = TRUE))

summary_data_annual_pie <- summary_data_annual_pie |> 
  mutate(label = paste0(round(avg_data_annual * 100, 2), "%")) |> 
  mutate(race_category = factor(race_category, levels = c("Black", "White", "Other")))

# Create the pie chart
ggplot(summary_data_annual_pie, aes(x ="", y = avg_data_annual, fill = race_category)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar("y") +
   geom_text(aes(label = label), position = position_stack(vjust = 0.5), color = "white", size = 4) +
  theme_void() +
  theme(legend.position = "none") +
  labs(title = "Average Proportion of DNA Collected Annually (Across All States)",
       fill = "Race") +
  scale_fill_manual(values = c("#05686D", "#F38C79", "#BAD1DC"))

data_state <- data_state |> 
  mutate(race_category = case_when(
    str_detect(variable_detailed, "black_state") ~ "Black",
    str_detect(variable_detailed, "white_state") ~ "White",
    TRUE ~ "Other")) |> 
    group_by(State, race_category) |> 
  summarise(total_value = sum(value, na.rm = TRUE), .groups = "keep")

summary_data_state <- data_state |> 
  group_by(race_category) |> 
  summarise(avg_data_state = mean(total_value, na.rm = TRUE))

summary_data_state <- summary_data_state |> 
  mutate(label = paste0(round(avg_data_state * 100, 2), "%")) |> 
  mutate(race_category = factor(race_category, levels = c("Black", "White", "Other")))

# Create the pie chart
ggplot(summary_data_state, aes(x ="", y = avg_data_state, fill = race_category)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar("y") +
   geom_text(aes(label = label), position = position_stack(vjust = 0.5), color = "white", size = 4) +
  theme_void() +
  labs(title = "Average Racial Breakdown (Across All States)",
       fill = "Race") +
  scale_fill_manual(values = c("#05686D", "#F38C79", "#BAD1DC"))

#combined_plot <- plot1 + plot2 +
#  plot_annotation(title = "Side by Side Pie Charts of Proportion of Races with DNA #Collected and Racial Breakdown of the Population Across All States") +
#  plot_layout(ncol = 2) & theme(plot.margin = unit(c(1, 2, 1, 2), "cm")) +
#  plot_spacer()
#combined_plot
```

The state with the highest proportion of their DNA database containing Black Americans is Delaware (79.9%) and the state with the lowest proportion of their DNA database containing Black Americans is Montana (1.6%).

```{r}
# Find the state with the highest value of black_annual
state_highest_black_annual <- black_annual |> 
  filter(value == max(value, na.rm = TRUE)) |> 
  mutate(percentage = value * 100) |> 
  select(State, percentage)

# Find the state with the lowest value of black_annual
state_lowest_black_annual <- black_annual |> 
  filter(value == min(value, na.rm = TRUE)) |> 
  mutate(percentage = value * 100) |> 
  select(State, percentage)

# Print the results
kable(state_highest_black_annual, caption = "State with Highest Proportion of Black American DNA Collected Annually", format = "markdown")
kable(state_lowest_black_annual, caption = "State with Lowest Proportion of Black American DNA Collected Annually", format = "markdown")


# Find the state with the highest value of white_annual
state_highest_white_annual <- white_annual |> 
  filter(value == max(value, na.rm = TRUE)) |> 
  mutate(percentage = value * 100) |> 
  select(State, percentage)

# Find the state with the lowest value of white_annual
state_lowest_white_annual <- white_annual |> 
  filter(value == min(value, na.rm = TRUE)) |> 
  mutate(percentage = value * 100) |> 
  select(State, percentage)

# Print the results
kable(state_highest_white_annual, caption = "State with Highest Proportion of White American DNA Collected Annually", format = "markdown")
kable(state_lowest_white_annual, caption = "State with Lowest Proportion of White American DNA Collected Annually", format = "markdown")



```

### Percent of Population Collected by Race
We can look at this data further and compare it to the racial breakdown of the total population in each state to explore the percentage of that population whose DNA has been collected.

For most states, the percentage of non-White people whose DNA is collected annually is much greater compared to White people. This disparity is most apparent for the Black population, which consistently shows the highest proportion of DNA collected.

Printed below are the average proportions across all states of people whose DNA is collected annually, divided by race. For example, the average percentage of Black Americans whose DNA is collected annually is 3.25%, meaning on average, 3.25% of the Black population has DNA collected annually. Other people of color have an average of 2.7% collected annually, compared to the White population, whose average percent of population with DNA collected is only 0.82%.

```{r}
#Graph showing stats of percent of population 

data_pop_collected <- data_race |> 
  filter(str_detect(variable_detailed, "_pop_collected")) |> 
  mutate(race_category = case_when(
    str_detect(variable_detailed, "black") ~ "Black",
    str_detect(variable_detailed, "white") ~ "White",
    TRUE ~ "Other"
  )) |> 
    group_by(State, race_category) |> 
  summarise(total_value = sum(value, na.rm = TRUE), .groups = "keep")

summary_pop_collected <- data_pop_collected |> 
  group_by(race_category) |> 
  summarise(avg_pop_collected = mean(total_value, na.rm = TRUE))

summary_pop_collected <- summary_pop_collected |> 
  mutate(label = paste0(round(avg_pop_collected * 100, 2), "%")) |> 
  mutate(race_category = factor(race_category, levels = c("Black", "White", "Other")))

# Create the pie chart
ggplot(summary_pop_collected, aes(x ="", y = avg_pop_collected, fill = race_category)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar("y") +
   geom_text(aes(label = label), position = position_stack(vjust = 0.5), color = "white", size = 4) +
  theme_void() +
  labs(title = "Average Percentage of Each Race with DNA Collected Annually",
       fill = "Race") +
  scale_fill_manual(values = c("#05686D", "#F38C79", "#BAD1DC"))

```

```{r}


```


