---
title: "Estimating Differences in the Distribution of First-degree Relatives"
author: "Manqing Lin, Tina Lasisi"
date: "`r format(Sys.time(), '%Y-%m-%d %H:%M:%S')`"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

The relative genetic surveillance of a population is influenced by the number of genetically detectable relatives individuals have. First-degree relatives (parents, siblings, and children) are especially relevant in forensic analyses using short tandem repeat (STR) loci, where close familial searches are commonly employed. To explore potential disparities in genetic detectability between African American and European American populations, we examined U.S. Census data from four census years (1960, 1970, 1980, and 1990) focusing on the number of children born to women over the age of 40.

## Data Sources
We used publicly available data from the Integrated Public Use Microdata Series (IPUMS) for the U.S. Census years 1960, 1970, 1980, and 1990. The datasets include information on:

- AGE: Age of the respondent.
- RACE: Self-identified race of the respondent.
- chborn_num: Number of children ever born to the respondent.

Data citation: Steven Ruggles, Sarah Flood, Matthew Sobek, Daniel Backman, Annie Chen, Grace Cooper, Stephanie Richards, Renae Rogers, and Megan Schouweiler. IPUMS USA: Version 14.0 [dataset]. Minneapolis, MN: IPUMS, 2023. https://doi.org/10.18128/D010.V14.0


```{r setup, include=FALSE, echo=TRUE, eval=TRUE}
#knitr::opts_knit$set(root.dir = ".")
knitr::opts_chunk$set(eval = TRUE, echo = TRUE, warning = FALSE, fig.width = 9, fig.height = 7)


# load necessary libraries
library(dplyr)
library(tidyverse)
library(viridis)
library(knitr)
library(scales)
library(rempsyc)
library(patchwork)
library(ggnewscale)
library(gridExtra)
library(purrr)
library(MASS)
library(pscl)
library(ggplot2)
library(nnet)
library(car)
library(rstatix)
```

## Data Preparation
Filtering Criteria: We selected women aged 40 and above to ensure that most had completed childbearing.

Due to the terms of agreement for using this data, we cannot share the full dataset but our repo contains the subset that was used to calculate the mean number of offspring and variance.

Race Classification: We categorized individuals into two groups:

- African American: Those who identified as "Black" or "African American".
- European American: Those who identified as "White".

Calculating Number of Siblings: For each child of these women, the number of siblings (n_sib) is one less than the number of children born to the mother:

$$
n_{sib} = chborn_{num} - 1
$$

```{r}
path <- file.path(".", "data")
savepath <- file.path(".", "output")

prop_race_year <- file.path(path, "proportions_table_by_race_year.csv")
data_filter <- file.path(path, "data_filtered_recoded.csv")

children_data = read.csv(prop_race_year)
mother_data = read.csv(data_filter)
```

```{r}
df <- mother_data %>%
  # Filter for women aged 40 and above
  filter(AGE >= 40) %>%
  mutate(
    # Create new age ranges
    AGE_RANGE = case_when(
      AGE >= 70 ~ "70+",
      AGE >= 60 ~ "60-69",
      AGE >= 50 ~ "50-59",
      AGE >= 40 ~ "40-49",
      TRUE ~ as.character(AGE_RANGE)  # This shouldn't occur due to the filter, but included for completeness
    ),
    # Convert CHBORN to ordered factor
    CHBORN = factor(case_when(
      chborn_num == 0 ~ "No children",
      chborn_num == 1 ~ "1 child",
      chborn_num == 2 ~ "2 children",
      chborn_num == 3 ~ "3 children",
      chborn_num == 4 ~ "4 children",
      chborn_num == 5 ~ "5 children",
      chborn_num == 6 ~ "6 children",
      chborn_num == 7 ~ "7 children",
      chborn_num == 8 ~ "8 children",
      chborn_num == 9 ~ "9 children",
      chborn_num == 10 ~ "10 children",
      chborn_num == 11 ~ "11 children",
      chborn_num >= 12 ~ "12+ children"
    ), levels = c("No children", "1 child", "2 children", "3 children", 
                  "4 children", "5 children", "6 children", "7 children", 
                  "8 children", "9 children", "10 children", "11 children", 
                  "12+ children"), ordered = TRUE),
    
    # Ensure RACE variable is correctly formatted and filtered
    RACE = factor(RACE, levels = c("White", "Black/African American"))
  ) %>%
  # Filter for African American and European American women
  filter(RACE %in% c("Black/African American", "White")) %>%
  # Select and reorder columns
  dplyr::select(YEAR, SEX, AGE, BIRTHYR, RACE, CHBORN, AGE_RANGE, chborn_num)

# Display the first few rows of the processed data
head(df)

# Summary of the processed data
summary(df)

# Check the levels of the RACE factor
levels(df$RACE)

# Count of observations by RACE
table(df$RACE)

# Count of observations by AGE_RANGE
table(df$AGE_RANGE)
```


# Distribution of Number of Children Across Census Years
First we visualize the general trends in the frequency of the number of children for African American and European American mothers across the Census years by age group. 

```{r}
# Calculate proportions within each group, ensuring proper normalization
df_proportions <- df %>%
  group_by(YEAR, RACE, AGE_RANGE, chborn_num) %>%
  summarise(count = n(), .groups = "drop") %>%
  group_by(YEAR, RACE, AGE_RANGE) %>%
  mutate(proportion = count / sum(count)) %>%
  ungroup()

# Reshape data for the mirror plot
df_mirror <- df_proportions %>%
  mutate(proportion = if_else(RACE == "White", -proportion, proportion))

# Create color palette
my_colors <- colorRampPalette(c("#FFB000", "#F77A2E", "#DE3A8A", "#7253FF", "#5E8BFF"))(13)

# Create the plot
child_plot <- ggplot(df_mirror, aes(x = chborn_num, y = proportion, fill = as.factor(chborn_num))) +
  geom_col(aes(alpha = RACE)) +
  geom_hline(yintercept = 0, color = "black", size = 0.5) +
  facet_grid(AGE_RANGE ~ YEAR, scales = "free_y") +
  coord_flip() +
  scale_y_continuous(
    labels = function(x) abs(x),
    limits = function(x) c(-max(abs(x)), max(abs(x)))
  ) +
  scale_x_continuous(breaks = 0:12, labels = c(0:11, "12+")) +
  scale_fill_manual(values = my_colors) +
  scale_alpha_manual(values = c("White" = 0.7, "Black/African American" = 1), guide = "none") +
  labs(
    title = "Distribution of Number of Children by Census Year, Race, and Age Range",
    x = "Number of Children",
    y = "Proportion",
    fill = "Number of Children",
    caption = "White population shown on left (negative values), Black/African American on right (positive values)\nProportions normalized within each age range, race, and census year\nFootnote: The category '12+' includes families with 12 or more children."
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, hjust = 0.5),
    axis.text.y = element_text(size = 8),
    strip.text = element_text(size = 10),
    legend.position = "none",
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank()
  )

print(child_plot)

# Print summary to check age ranges and normalization
print(df_proportions %>%
  group_by(YEAR, RACE, AGE_RANGE) %>%
  summarise(total_proportion = sum(proportion), .groups = "drop") %>%
  arrange(YEAR, RACE, AGE_RANGE))
```

With this visualization of the distribution of the data, we can see that there are differences between White and Black Americans. 

```{r}
# Mean Number of Children By race and Year
df %>% group_by(RACE, YEAR) %>% summarise(Mean = mean(chborn_num))
```

We will now test for differences in 1) the mean and variance, and 2) zero-inflation.

## Model Fit Across Census Years 

For each combination of race, census year, and age range, we fit the following candidate models:

- **Poisson Model**
- **Negative Binomial (NB) Model**
- **Zero-Inflated Poisson (ZIP) Model**
- **Zero-Inflated Negative Binomial (ZINB) Model**

### Fit and Compare Models
```{r}
combinations <- df %>%
  group_by(YEAR, RACE, AGE_RANGE) %>%
  group_split()

# Initialize the data frame to store results
best_models <- data.frame(
  Race = character(),
  Census_Year = numeric(),
  Age_Range = character(),
  AIC_poisson = numeric(),
  AIC_nb = numeric(),
  AIC_zip = numeric(),
  AIC_zinb = numeric(),
  stringsAsFactors = FALSE
)

# Loop through each subset of data
for (i in seq_along(combinations)) {
  subset_data <- combinations[[i]]
  
  # Fit Poisson model
  poisson_model <- glm(chborn_num ~ 1, family = poisson, data = subset_data)
  # Fit Negative Binomial model
  nb_model <- glm.nb(chborn_num ~ 1, data = subset_data)

  # Fit Zero-Inflated Poisson model
  zip_model <- zeroinfl(chborn_num ~ 1 | 1, data = subset_data, dist = "poisson")

  # Fit Zero-Inflated Negative Binomial model
  zinb_model <- zeroinfl(chborn_num ~ 1 | 1, data = subset_data, dist = "negbin")
  
  # Append the result to the best_models data frame
  best_models <- rbind(
    best_models,
    data.frame(
      Race = unique(subset_data$RACE),
      Census_Year = unique(subset_data$YEAR),
      Age_Range = unique(subset_data$AGE_RANGE),
      AIC_poisson = AIC(poisson_model),
      AIC_nb = AIC(nb_model),
      AIC_zip = AIC(zip_model),
      AIC_zinb = AIC(zinb_model),
      stringsAsFactors = FALSE
    )
  )
}

```

### Record the Best Model for Each Subset
```{r}
# Add a Best_Model column to store the best model based on minimum AIC
best_models$Best_Model <- apply(best_models, 1, function(row) {
  # Get AIC values for Poisson, NB, ZIP, and ZINB models
  aic_values <- c(Poisson = as.numeric(row['AIC_poisson']),
                  Negative_Binomial = as.numeric(row['AIC_nb']),
                  Zero_Inflated_Poisson = as.numeric(row['AIC_zip']),
                  Zero_Inflated_NB= as.numeric(row['AIC_zinb']))
  
  # Find the name of the model with the minimum AIC value
  best_model <- names(which.min(aic_values))
  
  return(best_model)
})

best_models <- best_models %>% dplyr::select(Race, Census_Year, Age_Range, Best_Model)
# View the updated table with the Best_Model column
print(best_models)
```


### Analyze the Effect of Race, Census Year, and Age Range

```{r}
# Recode Best_Model to a binary variable (e.g., "Zero_Inflated_NB" vs. "Other")
best_models$Best_Model_Binary <- ifelse(best_models$Best_Model == "Zero_Inflated_NB", 1, 0)

# Fit binary logistic regression
model_logistic <- glm(Best_Model_Binary ~ Race + Census_Year + Age_Range, family = binomial(), data = best_models)

# Summary of the logistic regression model
summary(model_logistic)

```
**_RESULT:_**The p-value for each variable is larger than 0.05, indicating  non-significant effects on best model fitting. 

### Visualize the Results

```{r}
ggplot(best_models, aes(x = Census_Year, y = Age_Range, fill = Best_Model)) +
  geom_tile(color = "white", lwd = 0.5, linetype = 1) + 
  facet_wrap(~Race, nrow = 2) +  
  scale_fill_manual(values = c("Zero_Inflated_Poisson" = "#0072B2", "Zero_Inflated_NB" = "#F0E442")) + 
  labs(
    title = "Best Model by Race, Census Year, and Age Range",
    x = "Census Year",
    y = "Age Range",
    fill = "Best Model"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    legend.position = "bottom"
  )
```

### Conclusion
The logistics regression shows that there isn't a significant association between the predictors(races, age ranges, and census year) and the best-fitting model. According to the resulting visualization, the ZINB model is the best fit for the black population across census year and age group. However, the ZINB model perform the best among the white population only across year 1960 and 1970, and age group 60-69 and 70+. 

## Cohort Stability Analysis 
The goal is to determine if there is a significant change in **zero inflation**, **family size**, or **model fit** for the same cohort across different census years, given race. 

### Table with Birth-Year (Cohort)
```{r}
calculate_cohort <- function(census_year, age_range) {
  age_limits <- as.numeric(unlist(strsplit(age_range, "-|\\+")))
  
  if (length(age_limits) == 1) {
    lower_age <- age_limits
    upper_age <- Inf  
  } else {
    lower_age <- age_limits[1]
    upper_age <- age_limits[2]
  }
  
  lower_cohort <- census_year - upper_age
  upper_cohort <- census_year - lower_age
  
  
  if (upper_age == Inf) {
    upper_cohort <- census_year - 70
  }
  
  return(paste(lower_cohort, upper_cohort, sep = "-"))
}

best_models <- best_models %>%
  mutate(Cohort = mapply(calculate_cohort, Census_Year, Age_Range)) %>% 
  dplyr::select(Race, Census_Year, Age_Range, Best_Model, Cohort)

print(best_models)

```


### Summary Statistics for Each Subset

```{r}
df_merg <- df %>%
  rename(Census_Year = YEAR, Race = RACE, Age_Range = AGE_RANGE)

# Perform a left join to merge the data frames
merged_data <- left_join(df_merg, best_models, by = c("Race", "Census_Year", "Age_Range"))
```

```{r}
calculate_mode <- function(x) {
  ux <- unique(x)
  ux[which.max(tabulate(match(x, ux)))]
}

summary_table <- merged_data %>% group_by(Race, Cohort, Census_Year) %>%
  summarise(
    Mean = mean(chborn_num),
    Variance = var(chborn_num),
    Mode = calculate_mode(chborn_num),
    Prob_0 = mean(chborn_num == 0),
    Prob_1 = mean(chborn_num == 1),
    Prob_2 = mean(chborn_num == 2),
    Prob_3 = mean(chborn_num == 3),
    Prob_4 = mean(chborn_num == 4),
    Prob_5 = mean(chborn_num == 5),
    Prob_6 = mean(chborn_num == 6),
    Prob_7 = mean(chborn_num == 7),
    Prob_8 = mean(chborn_num == 8),
    Prob_9 = mean(chborn_num == 9),
    Prob_10 = mean(chborn_num == 10),
    Prob_11 = mean(chborn_num == 11),
    Prob_12plus = mean(chborn_num >= 12)
    #Total_Prob = Prob_0 + Prob_1 + Prob_2 + Prob_3 + Prob_4 + Prob_5 + Prob_6 + 
                # Prob_7 + Prob_8 + Prob_9 + Prob_10 + Prob_11 + Prob_12plus
  ) %>%
  ungroup() %>% as.data.frame()

# View the summary table
print(summary_table)
```


### Test for Significant Changes Over Time Within Each Cohort
The goal is to determine if there is a **significant change** in any of the following across census years for the same cohort:

1. **Zero Inflation**: Check if the proportion of women with zero children changes significantly over time for the same cohort.
2. **Family Size**: Test if the mean or variance in the number of children changes for the same cohort over time.
3. **Model Fit**: Analyze if the best-fitting model for the cohort changes over time.

#### a. **Zero-Inflation Analysis**

```{r}
## divided the corhort by race
merged_df_black <- merged_data %>% filter(Race == "Black/African American")
merged_df_white <- merged_data %>% filter(Race == "White")
```

```{r}
# Data frame to store proportions
proportions_df_b <- data.frame(Cohort = character(), Census_Year = character(), 
                             Proportion_Zero_Children = numeric(), Race = character(), stringsAsFactors = FALSE)

# Filter cohorts with more than one census year
cohort_counts <- merged_df_black %>%
  group_by(Cohort) %>%
  summarise(Unique_Census_Years = n_distinct(Census_Year)) %>%
  filter(Unique_Census_Years > 1)

# Get the list of cohorts for testing
cohorts_for_test <- cohort_counts$Cohort

# Initialize results data frame
results_black <- data.frame(RACE = character(), Cohort = character(), Chi_Square = numeric(), p_value = numeric(), stringsAsFactors = FALSE)

# Loop through each cohort and run the chi-square test
for (cohort in cohorts_for_test) {
  cohort_data <- merged_df_black %>% filter(Cohort == cohort)
  
  # Create contingency table
  contingency_table <- table(cohort_data$Census_Year, cohort_data$chborn_num == "0")
  
  # Perform the chi-square test only if more than one row is in the table
  if (nrow(contingency_table) > 1) {
    test_result <- chisq.test(contingency_table)
    
    # Append results to the results data frame
    results_black <- rbind(results_black, data.frame(
      RACE = "Black/African American",
      Cohort = cohort,
      Chi_Square = test_result$statistic,
      p_value = test_result$p.value
    ))
    
    # proportion across year
    proportions <- cohort_data %>%
      group_by(Census_Year) %>%
      summarise(Proportion_Zero_Children = mean(chborn_num == "0"))
    
    proportions$Cohort <- cohort
    proportions$Race <- "Black"
    
    # Combine into one data frame
    proportions_df_b <- rbind(proportions_df_b, proportions)
  
    # Print proportions for review
    print(paste("Proportions for cohort:", cohort))
    print(proportions)
  }
}

results_black <- results_black %>%
  mutate(Significance = ifelse(p_value > 0.05, "No", "Yes"))

print(results_black)

```

```{r}
# Function to determine the nature of change
determine_nature_of_change <- function(proportions) {
  if (nrow(proportions) > 1) {
    if (all(diff(proportions$Proportion_Zero_Children) > 0)) {
      return("Increase")
    } else if (all(diff(proportions$Proportion_Zero_Children) < 0)) {
      return("Decrease")
    } else {
      return("Mixed/No Change")
    }
  } else {
    return("Not Applicable")
  }
}

# Add the Nature_of_Change column
results_black$Nature_of_Change <- NA  # Initialize the column

# Loop through each cohort in results_black
for (i in 1:nrow(results_black)) {
  cohort <- results_black$Cohort[i]
  cohort_data <- merged_df_black %>% filter(Cohort == cohort)
  
  # Calculate proportions for the cohort
  proportions <- cohort_data %>%
    group_by(Census_Year) %>%
    summarise(Proportion_Zero_Children = mean(chborn_num == "0"))
    
  # Determine nature of change
  results_black$Nature_of_Change[i] <- determine_nature_of_change(proportions)
}

# View the updated results_black
print(results_black)

```

```{r}
# Data frame to store proportions
proportions_df_w <- data.frame(Cohort = character(), Census_Year = character(), 
                             Proportion_Zero_Children = numeric(), Race = character(), stringsAsFactors = FALSE)

# Filter cohorts with more than one census year
cohort_counts2 <- merged_df_white %>%
  group_by(Cohort) %>%
  summarise(Unique_Census_Years = n_distinct(Census_Year)) %>%
  filter(Unique_Census_Years > 1)

# Get the list of cohorts for testing
cohorts_for_test2 <- cohort_counts2$Cohort

# Initialize results data frame
results_white <- data.frame(RACE = character(), Cohort = character(), Chi_Square = numeric(), p_value = numeric(), stringsAsFactors = FALSE)

# Loop through each cohort and run the chi-square test
for (cohort in cohorts_for_test2) {
  cohort_data <- merged_df_white %>% filter(Cohort == cohort)
  
  # Create contingency table
  contingency_table <- table(cohort_data$Census_Year, cohort_data$chborn_num == "0")
  
  # Perform the chi-square test only if more than one row is in the table
  if (nrow(contingency_table) > 1) {
    test_result <- chisq.test(contingency_table)
    
    # Append results to the results data frame
    results_white <- rbind(results_white, data.frame(
      RACE = "White",
      Cohort = cohort,
      Chi_Square = test_result$statistic,
      p_value = test_result$p.value
    ))
    
    # proportion across year
    proportions <- cohort_data %>%
      group_by(Census_Year) %>%
      summarise(Proportion_Zero_Children = mean(chborn_num == "0"))
    
    proportions$Cohort <- cohort
    proportions$Race <- "White"
    
    # Combine into one data frame
    proportions_df_w <- rbind(proportions_df_w, proportions)
  
    # Print proportions for review
    print(paste("Proportions for cohort:", cohort))
    print(proportions)
  }
}
results_white <- results_white %>%
  mutate(Significance = ifelse(p_value > 0.05, "No", "Yes"))

print(results_white)
```

```{r}
# Add the Nature_of_Change column
results_white$Nature_of_Change <- NA  # Initialize the column

# Loop through each cohort in results_white
for (i in 1:nrow(results_white)) {
  cohort <- results_white$Cohort[i]
  cohort_data <- merged_df_white %>% filter(Cohort == cohort)
  
  # Calculate proportions for the cohort
  proportions <- cohort_data %>%
    group_by(Census_Year) %>%
    summarise(Proportion_Zero_Children = mean(chborn_num == "0"))
  
  # Determine nature of change
  results_white$Nature_of_Change[i] <- determine_nature_of_change(proportions)
}

# View the updated results_white
print(results_white)

```


```{r}
# Combine the two data frames
combined_result <- rbind(results_black, results_white)

# Create the plot
ggplot(combined_result, aes(x = Cohort, y = p_value, color = RACE, shape = Significance)) +
  geom_point(size = 2) +
  geom_line(aes(group = RACE)) +
    annotate("text", x = Inf, y = 0.05, label = "p = 0.05", hjust = 1.1, vjust = -0.5, color = "red", size = 3) +
  geom_hline(yintercept = 0.05, linetype = "dashed", color = "red", size = 0.5) + 
  scale_y_log10() +  # Use log scale for better visualization if p-values vary widely
  labs(
    title = "Change of p-values by Cohort",
    x = "Cohort",
    y = "p-value (log scale)",
    color = "Race",
    shape = "Significance"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```
**_RESULT:_** The graph shows that there is a discrepancy in the p-value within the cohorts like 1901-1910, 1911-1920 and 1921-1930 by race. However, in cohort 1931-1941, the p-value of each race is pretty close to each other. The overall trend of the p-value for black population is stable across cohort, while the trend for white population fluctuate a lot. 

#### b. **Family Size Analysis**

```{r}
merged_black = merged_data %>% filter(Race == "Black/African American")
merged_white = merged_data %>% filter(Race == "White")
```


```{r}
cohorts1 <- unique(merged_black$Cohort)

# Loop through each cohort and apply ANOVA for Mean and Variance
for (cohort in cohorts1) {
  
  # Filter data for the specific cohort
  cohort_data_b <- merged_black[merged_black$Cohort == cohort, ]
  
  # Check if there are at least two unique Census_Year values
  if (length(unique(cohort_data_b$Census_Year)) < 2) {
    cat("\nCohort:", cohort, "has only 1 Census Year. Skipping ANOVA.\n")
    next
  }
  else if (length(unique(cohort_data_b$Census_Year)) == 2){
    t_test_result <- t.test(chborn_num ~ Census_Year, data = cohort_data_b)
     # Output the results
  cat("\nCohort:", cohort)
  cat("\nT-test for Mean Family Size:\n")
  print(t_test_result)
  }
  else{
  # Perform ANOVA for Mean
  anova_mean_result <- aov(chborn_num ~ factor(Census_Year), data = cohort_data_b)
   # Output the results
  cat("\nCohort:", cohort)
  cat("\nANOVA for Mean Family Size:\n")
  print(summary(anova_mean_result))
  }
}

```

```{r}
cohorts2 <- unique(merged_white$Cohort)

# Loop through each cohort and apply ANOVA for Mean and Variance
for (cohort in cohorts2) {
  
  # Filter data for the specific cohort
  cohort_data_w <- merged_white[merged_white$Cohort == cohort, ]
  
  # Check if there are at least two unique Census_Year values
  if (length(unique(cohort_data_w$Census_Year)) < 2) {
    cat("\nCohort:", cohort, "has only 1 Census Year. Skipping ANOVA.\n")
    next
  }
  else if (length(unique(cohort_data_w$Census_Year)) == 2){
    t_test_result <- t.test(chborn_num ~ Census_Year, data = cohort_data_w)
     # Output the results
  cat("\nCohort:", cohort)
  cat("\nT-test for Mean Family Size:\n")
  print(t_test_result)
  }
  else{
  # Perform ANOVA for Mean
  anova_mean_result <- aov(chborn_num ~ factor(Census_Year), data = cohort_data_w)
   # Output the results
  cat("\nCohort:", cohort)
  cat("\nANOVA for Mean Family Size:\n")
  print(summary(anova_mean_result))
  }
}
```

**_RESULT:_** 
  1.In both racial group, the ANOVA and t-test is not applicable for the following cohort since there is only one census year available in the data for those cohorts:-Inf-1910, 1941-1950, -Inf-1920, -Inf-1890, -Inf-1900, and 1891-1900
  2.By looking at the ANOVA and t-test for mean family size in white population, the p-value of cohorts 1901-1910, 1911-1920 and 1921-1930 are smaller than 0.05. In the black population, the p-value of cohorts 1901-1910 and 1921-1930 are smaller than 0.05. These results indicate mean family size for these cohort has significantly changed over different census years in different racial population.


#### c. **Model Fit Analysis**

```{r}
# Due to the due lack of data variability, I use visualization instead of test

ggplot(best_models, aes(x = Census_Year, y = Cohort, fill = Best_Model)) +
  geom_tile(color = "white", lwd = 0.5, linetype = 1) + # Create the tiles
  facet_wrap(~Race, nrow = 2) + # Facet by Race
  scale_fill_manual(values = c("Zero_Inflated_Poisson" = "#0072B2", "Zero_Inflated_NB" = "#F0E442")) + 
  labs(
    title = "Best Model by Race, Census Year, and Cohort(Birth Range)",
    x = "Census Year",
    y = "Birth Range",
    fill = "Best Model"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    legend.position = "bottom"
  )
```

**_RESULT:_** Based on the heatmap created, we can see that the model fit for each cohort(cohort with 2+ corresponding census year) across year does not change.

### Conclusion
- **Zero-inflation**: We observe that the proportion of women with zero children change significantly across census year in the following cohort by race:
  -black population:1901-1910, 1921-1930
  -white population:1901-1910, 1911-1920, 1921-1930
- **Family size**: Cohorts 1901-1910, 1911-1920, 1921-1930, and 1931-1940 present significant change in mean family size across year for both races. 
- **Best-fitting Model**: Model fit for each cohort(cohort with 2+ corresponding census year) across year does not change in both racial group. 


## Analyzing and Visualizing Significant Fertility Shifts
The goal is to summarize the previous information and create visualization that illustrates significant fertility shifts in cohorts, compares fertility patterns of 40-49 year-olds to 50-59 year-olds in the 1990 census so we can pick the set of fertility distributions we want to use to visualize the sibling distribution and do the math on the genetic surveillance.


### Panel A: Fertility Distribution Shifts Across Cohorts

#### Cohorts with Significant Change in Zero Inflation
```{r}
# Significant Change in Zero Inflation
## Balck population
filter_prop_b <- proportions_df_b %>% 
  filter(Cohort %in% c("1901-1910", "1921-1930"))

## White population
filter_prop_w <- proportions_df_w %>% 
  filter(Cohort %in% c("1901-1910", "1911-1920", "1921-1930"))

combined_data <- bind_rows(filter_prop_b, filter_prop_w)

ggplot(combined_data, aes(x = Census_Year, y = Proportion_Zero_Children, color = Cohort)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  facet_wrap(~ Race, scales = "free_x") +
  labs(
    title = "Significant Change in Probability of Zero Children by Race and Cohort",
    x = "Census Year",
    y = "Proportion of Zero Children",
    color = "Cohort"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "bottom"
  )

```


#### Cohorts with Significant Change in Mean Family Size
```{r}
merged_data2 <- left_join(merged_data, summary_table, by = c("Race","Cohort","Census_Year")) %>% dplyr::select(Race, Cohort, Census_Year, Mean, Variance)

## Balck population
filter_mean_b <- merged_data2 %>% 
  filter(Cohort %in% c("1901-1910", "1921-1930") & Race == "Black/African American")

## White population
filter_mean_w <- merged_data2 %>% 
  filter(Cohort %in% c("1901-1910", "1911-1920", "1921-1930") & Race == "White")

combined_data <- bind_rows(filter_mean_b, filter_mean_w)

ggplot(combined_data, aes(x = Census_Year, y = Mean, color = Cohort)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  facet_wrap(~ Race, scales = "free_x") +
  labs(
    title = "Significant Changes in Mean Family Size Over Census Years",
    x = "Census Year",
    y = "Mean number of Children",
    color = "Cohort"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "bottom"
  )
```


### Panel B: Comparison of 40-49 and 50-59 Age Groups in 1990

#### Visulization of Distribution
```{r}
filter_df <- df_proportions %>% 
  filter(YEAR == "1990", AGE_RANGE %in% c("40-49", "50-59")) 
```


```{r}
ggplot(filter_df, aes(x = chborn_num, y = proportion, fill = AGE_RANGE)) +
  geom_col(position = "identity", alpha = 0.5) +  
  facet_wrap(~ RACE) +  
  labs(
    title = "Distribution of Number of Children for Women in 40-49 and 50-59 Age Groups in 1990",
    x = "Number of Children",
    y = "Proportion",
    fill = "Age Range"
  ) 
```

####  Summary Statistics Table
```{r}
summary_stats <- filter_df %>%
  group_by(AGE_RANGE, RACE) %>%
  summarise(
    mean_children = weighted.mean(chborn_num, count),  # Mean number of children
    variance_children = sum(count * (chborn_num - weighted.mean(chborn_num, count))^2) / sum(count),  # Variance
    zero_inflation = sum(count[chborn_num == 0]) / sum(count)  # Proportion of women with 0 children
  )

# Print summary statistics table
summary_stats
```

#### Test for Difference in Means

```{r}
## Check the normality assumption for black population
# Subset data for the 40-49 and 50-59 age groups
data_40_49_b <- df %>% filter(AGE_RANGE == "40-49", RACE == "Black/African American")
data_50_59_b <- df %>% filter(AGE_RANGE == "50-59", RACE == "Black/African American")

qqnorm(data_40_49_b$chborn_num, main = "QQ Plot: 40-49 Age Group (Black/African American)")
qqline(data_40_49_b$chborn_num)

# QQ plot for the 50-59 age group
qqnorm(data_50_59_b$chborn_num, main = "QQ Plot: 50-59 Age Group (Black/African American)")
qqline(data_50_59_b$chborn_num)
```


```{r}
## Check the normality assumption for white population
data_40_49_w <- df %>% filter(AGE_RANGE == "40-49", RACE == "White")
data_50_59_w <- df %>% filter(AGE_RANGE == "50-59", RACE == "White")

qqnorm(data_40_49_w$chborn_num, main = "QQ Plot: 40-49 Age Group (White)")
qqline(data_40_49_w$chborn_num)

# QQ plot for the 50-59 age group
qqnorm(data_50_59_w$chborn_num, main = "QQ Plot: 50-59 Age Group (White)")
qqline(data_50_59_w$chborn_num)
```

Since the data for black population and white population violate normality assumption, we perform a Mann-Whitney U test for both racial groups in the follwoing:
```{r}
wilcox_test_result <- wilcox.test(data_40_49_b$chborn_num, data_50_59_b$chborn_num)$p.value
wilcox_test_result2 <- wilcox.test(data_40_49_w$chborn_num, data_50_59_w$chborn_num)$p.value
print(c(wilcox_test_result, wilcox_test_result2))
```
**_RESULT:_** These p-values are both extremely small (close to zero), meaning there is a very strong statistical difference between the two age groups (40-49 and 50-59) in terms of the mean number of children within each racial group. 

#### Test for Difference in Variances

```{r}
## Apply Levene's test since non-normality
df_w <- df %>% filter(RACE == "White")
df_b <- df %>% filter(RACE == "Black/African American")

# Levene's test for Black group
levene_test_b <- leveneTest(chborn_num ~ as.factor(AGE_RANGE), data = df_b)

# Levene's test for White group
levene_test_w <- leveneTest(chborn_num ~ as.factor(AGE_RANGE), data = df_w)

# Print results
print(levene_test_b)
print(levene_test_w)
```
**_RESULT:_** Since the p values are smaller than 0.05 for both racial group, we have enough evidence to reject the null hypothesis, indicating that the variances of the number of children between the two age groups within each racial group are significantly different.

#### Chi-Square Test for Difference in Zero Inflation

```{r}
# contingency tables showing the counts of women with zero children and those with one or more children for each age group 
ZI_table <- df %>% filter(YEAR == "1990", AGE_RANGE %in% c("40-49", "50-59")) %>% 
  mutate(childlessness = ifelse(chborn_num == 0, "0 Children", "1+ Children")) %>%       group_by(RACE, AGE_RANGE, childlessness) %>%
  summarise(count = n())

ZI_table
```

```{r}
ZI_table_b <- ZI_table %>%
  filter(RACE == "Black/African American")

cont_table_b <- xtabs(count ~ AGE_RANGE + childlessness, data = ZI_table_b)
chi_test_b <- chisq.test(cont_table_b)
chi_test_b
```

```{r}
ZI_table_w <- ZI_table %>%
  filter(RACE == "White")

cont_table_w <- xtabs(count ~ AGE_RANGE + childlessness, data = ZI_table_w)
chi_test_w <- chisq.test(cont_table_w)
chi_test_w
```
**_RESULT:_** The p-values for both tests are extremely low, suggests that there is a significant difference in the proportion of women with 0 children across age ranges (40-49 vs. 50-59) for both the Black/African American and White racial groups. The results imply that childlessness is not uniformly distributed across age groups. 

### Step 4: Summarize Findings

Write a clear and concise summary addressing the following points.

#### 4.1 Significant Fertility Shifts Across Cohorts

- **Identify Timing of Shifts**:
  - Specify when significant fertility shifts occurred for each racial group based on your analysis in Question 2.
  
- **Describe Nature of Shifts**:
  - Detail the characteristics of the shifts, such as:
    - Decrease in mean number of children
    - Increase in childlessness (zero inflation)
    - Changes in variance or distribution shape
  
- **Highlight Differences Between Racial Groups**:
  - Compare the timing and nature of shifts between African American and European American women.
  - Discuss any patterns or discrepancies observed.

#### 4.2 Comparison of 40-49 and 50-59 Age Groups in 1990

- **Summarize Key Differences**:
  - Present the differences in fertility patterns between the two age groups for each racial group.
  - Include comparisons of:
    - Distribution shapes
    - Mean number of children
    - Variance
    - Zero inflation
  
- **Report Statistical Test Results**:
  - Provide the results of the t-tests, F-tests, and chi-square tests.
  - Interpret the significance of these results in the context of your analysis.
  
- **Discuss Implications**:
  - Consider what these differences suggest about fertility trends and behaviors.
  - Reflect on whether the older age group represents completed fertility patterns.

#### 4.3 Synthesis and Implications

- **Integrate Findings**:
  - Connect the insights from the cohort analysis with the age group comparison.
  - Discuss how the patterns observed in the 1990 census relate to the shifts identified across cohorts.
  
- **Consider Contributing Factors**:
  - Explore potential social, economic, or policy factors that may have contributed to the observed fertility shifts.
    - For example, changes in access to education, employment opportunities, or family planning resources.
  
- **Reflect on Broader Implications**:
  - Discuss how these fertility trends might impact your broader research topic, such as genetic surveillance disparities.
  - Consider the implications for future demographic research or policy development.

### Additional Considerations

- **Visual Clarity**:
  - Ensure all visualizations are easy to interpret.
  - Use clear labels, legends, and annotations.
  
- **Contextualize Statistical Significance**:
  - Explain not just whether results are statistically significant, but also what they mean in practical terms.
  
- **Acknowledge Data Limitations**:
  - Discuss any limitations or biases in the data that could affect your findings.
    - For instance, sample size constraints or missing data.
  
- **Ethical Considerations**:
  - Approach discussions of race and fertility sensitively and responsibly.
  - Avoid drawing causal conclusions without robust evidence.

### Integration with Previous Work

- **Leverage Previous Analyses**:
  - Use the visualizations and statistical summaries from Questions 1 and 2 as foundations for this analysis.
  
- **Create a Cohesive Narrative**:
  - Ensure that your findings from all questions are connected and build upon each other.
  - Tell a comprehensive story about fertility trends across cohorts and racial groups.

### Final Deliverables

- **Multi-Panel Visualization**:
  - Panel A: Fertility distribution shifts across cohorts with highlighted significant shifts.
  - Panel B: Comparative distribution plots for 40-49 and 50-59 age groups in 1990, including summary statistics.
  
- **Written Summary**:
  - A concise report that addresses the points outlined in Step 4.
  - Include interpretations of statistical analyses and discuss broader implications.
  
- **Statistical Analysis Documentation**:
  - Provide details of the statistical tests conducted, including test assumptions, results, and interpretations.
  
- **Annotated Code (if applicable)**:
  - While not the focus, include any new code used for this analysis with appropriate comments.


# Distribution of Number of Siblings Across Census Years
Having analyzed the distribution of the number of children, we now turn our attention to the distribution of the number of siblings. We will explore the trends in the frequency of the number of siblings for African American and European American mothers across the Census years by age group.

Frequency of siblings is calculated as follows.

$$
\text{freq}_{n_{\text{sib}}} = \text{freq}_{\text{mother}} \cdot \text{chborn}_{\text{num}}
$$

For example, suppose 10 mothers (generation 0) have 7 children, then there will be 70 children (generation 1) in total who each have 6 siblings.

We take our original data and calculate the frequency of siblings for each mother based on the number of children they have. We then aggregate this data to get the frequency of siblings for each generation along with details on the birth years of the relevant children to visualize the distribution of the number of siblings across generations.

## Data Preparation

### Calculate Sibling Frequencies
```{r}
df2 <- df %>% 
  dplyr::select(RACE, YEAR, AGE_RANGE, chborn_num)
```

```{r}
df2 <- df2 %>%
 mutate(n_siblings = chborn_num - 1)
```

```{r}
df2 <- df2 %>%
mutate(sibling_freq = ifelse(chborn_num != 1, n_siblings * 1, 1))  # Assuming each mother represents 1 in frequency
```

### Aggregate Sibling Data
```{r}
# Group the data and calculate sibling frequencies
df_siblings <- df2 %>%
 group_by(YEAR, RACE, AGE_RANGE, n_siblings) %>%
 summarise(
   sibling_count = sum(sibling_freq),
   .groups = "drop"
 )
```

```{r}
# Calculate proportions within each group
df_sibling_proportions <- df_siblings %>%
 group_by(YEAR, RACE, AGE_RANGE) %>%
 mutate(proportion = sibling_count / sum(sibling_count)) %>%
 ungroup()
```

### Check Normalization
```{r}
normalization_check <- df_sibling_proportions %>%
  group_by(YEAR, RACE, AGE_RANGE) %>%
  summarise(total_proportion = sum(proportion), .groups = "drop") %>%
  arrange(YEAR, RACE, AGE_RANGE)

print(normalization_check)
```


## Visualization of Distribution of Number of Siblings

```{r}
df_sibling_mirror <- df_sibling_proportions %>%
  mutate(proportion = if_else(RACE == "White", -proportion, proportion))
```


```{r}
my_colors <- colorRampPalette(c("#FFB000", "#F77A2E", "#DE3A8A", "#7253FF", "#5E8BFF"))(13)
```

```{r}
ggplot(data = df_sibling_mirror %>% filter(n_siblings != "-1"), aes(x = n_siblings, y = proportion, fill = as.factor(n_siblings))) +
 geom_col(aes(alpha = RACE)) +
 geom_hline(yintercept = 0, color = "black", size = 0.5) +
 facet_grid(AGE_RANGE ~ YEAR, scales = "free_y") +
 coord_flip() +
 scale_y_continuous(
   labels = function(x) abs(x),
   limits = function(x) c(-max(abs(x)), max(abs(x)))
 ) +
 scale_x_continuous(breaks = 0:11, labels = c(0:10, "11+")) +
 scale_fill_manual(values = my_colors) +
 scale_alpha_manual(values = c("White" = 0.7, "Black/African American" = 1)) +
 labs(
   title = "Distribution of Number of Siblings by Census Year, Race, and Age Range",
   x = "Number of Siblings",
   y = "Proportion",
   fill = "Number of Siblings",
   caption = "White population shown on left (negative values), Black/African American on right (positive values)\nProportions normalized within each age range, race, and census year\nFootnote: The category '11+' includes individuals with 11 or more siblings."
 ) +
 theme_minimal() +
 theme(
   plot.title = element_text(size = 14, hjust = 0.5),
   axis.text.y = element_text(size = 8),
   strip.text = element_text(size = 10),
   legend.position = "none",
   panel.grid.major.y = element_blank(),
   panel.grid.minor.y = element_blank()
 )
```


```{r}
# Mean Number of Sibling By race only
df2 %>% filter(RACE == "Black/African American" & n_siblings >= 0) %>% summarise(Mean = mean(n_siblings))

df2 %>% filter(RACE == "White" & n_siblings >= 0) %>% summarise(Mean = mean(n_siblings))
```

**_RESULT:_**

**_-By Census Year:_** 
In 1960 and 1970, individuals are more likely to have higher number of siblings,      especially in the 5-10 range. This trend diminishes over time.

By 1980 and 1990, the distribution shifts toward smaller family sizes, with a growing proportion of individuals having fewer siblings.

**_-By age range:_**

40-49 Age Group: For this group, the number of individuals with 0-2 siblings increases across census years, especially in 1980 and 1990, while the proportion of individuals with larger sibling counts decreases.

50-59 and 60-69 Age Groups: These groups show a similar shift toward smaller family sizes, but the trend is slightly more gradual compared to the younger age group.

70+ Age Group: The shift to fewer siblings is noticeable, although the trend is less pronounced. The distribution remains relatively stable across the census years, with a significant portion of individuals still coming from large families in 1960 and 1970.

2. Compare the distributions between White and Black/African American populations.

**_RESULT:_**
-Black/African American Populations (right side of each pair) consistently show a higher proportion of individuals with larger sibling counts (5-10 siblings) compared to White populations. However, similar to the White population, the number of individuals with fewer siblings increases over time.

3. Note any significant changes or trends in the number of siblings over time.

**_RESULT:_**
-White Populations: (left side of each pair) have a more marked shift toward smaller families by 1990, with a larger proportion of individuals having 0-2 siblings compared to the Black/African American population. The decline in larger family sizes (5+ siblings) is more pronounced among Whites, particularly by 1980 and 1990.

4. Consider how these sibling distributions might differ from the children distributions you analyzed earlier, and think about potential reasons for these differences.

**_RESULT:_** 
While both distributions show a trend toward smaller families, the sibling distribution is more spread out across different sibling counts, suggesting potential difference in the distribution. 



## Model Fit Across Census Years

We repeat the model fitting process we performed for the children distribution, this time using the sibling distribution data.

```{r}
combinations2 <- df2 %>%
  # treat number of sib = -1 as NA
  filter(n_siblings != -1) %>%
  group_by(YEAR, RACE, AGE_RANGE) %>%
  group_split()

# Initialize the data frame to store results
best_models_sib <- data.frame(
  Race = character(),
  Census_Year = numeric(),
  AGE_RANGE = character(),
  AIC_poisson = numeric(),
  AIC_nb = numeric(),
  AIC_zip = numeric(),
  AIC_zinb = numeric(),
  stringsAsFactors = FALSE
)

# Loop through each subset of data
for (i in seq_along(combinations2)) {
  subset_data <- combinations2[[i]]
  
  # Fit Poisson model
  poisson_model <- glm(n_siblings ~ 1, family = poisson, data = subset_data)
  # Fit Negative Binomial model
  nb_model <- glm.nb(n_siblings ~ 1, data = subset_data)

  # Fit Zero-Inflated Poisson model
  zip_model <- zeroinfl(n_siblings ~ 1 | 1, data = subset_data, dist = "poisson",control = zeroinfl.control(maxit = 100))
  
  # Fit Zero-Inflated Negative Binomial model with increased max iterations
  zinb_model <- zeroinfl(n_siblings ~ 1 | 1, data = subset_data, dist = "negbin",control = zeroinfl.control(maxit = 100))
  
  # Append the result to the best_models data frame
  best_models_sib <- rbind(
    best_models_sib,
    data.frame(
      Race = unique(subset_data$RACE),
      Census_Year = unique(subset_data$YEAR),
      AGE_RANGE = unique(subset_data$AGE_RANGE),
      AIC_poisson = AIC(poisson_model),
      AIC_nb = AIC(nb_model),
      AIC_zip = AIC(zip_model),
      AIC_zinb = AIC(zinb_model),
      stringsAsFactors = FALSE
    )
  )
}

```


### Identify Best-Fitting Models by comparing AIC
```{r}
# Add a Best_Model column to store the best model based on minimum AIC
best_models_sib$Best_Model <- apply(best_models_sib, 1, function(row) {
  # Get AIC values for Poisson, NB, ZIP, and ZINB models
  aic_values <- c(Poisson = as.numeric(row['AIC_poisson']),
                  Negative_Binomial = as.numeric(row['AIC_nb']),
                  Zero_Inflated_Poisson = as.numeric(row['AIC_zip']),
                  Zero_Inflated_NB= as.numeric(row['AIC_zinb']))
  
  # Find the name of the model with the minimum AIC value
  best_models_sib <- names(which.min(aic_values))
  
  return(best_models_sib)
})

best_models_sib <- best_models_sib %>% dplyr::select(Race, Census_Year, AGE_RANGE, Best_Model)
# View the updated table with the Best_Model column
print(best_models_sib)
```


### Visualize Results
```{r}
ggplot(best_models_sib, aes(x = Census_Year, y = AGE_RANGE, fill = Best_Model)) +
  geom_tile(color = "white", lwd = 0.5, linetype = 1) + # Create the tiles
  facet_wrap(~Race, nrow = 2) + # Facet by Race
  scale_fill_manual(values = c("Poisson" = "#0072B2", "Zero_Inflated_NB" = "#F0E442", "Negative_Binomial" = "grey")) + 
  labs(
    title = "Best Model(siblings) by Race, Census Year, and Birth Range",
    x = "Census Year",
    y = "Age Range",
    fill = "Best Model"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1), # Rotate x-axis labels for readability
    plot.title = element_text(size = 14, face = "bold"),
    legend.position = "bottom"
  )
```

**_RESULT:_** By comparing the pattern of best-fitting models between the sibling and children distributions, we observe that the best model for black population has the same best model(zero-inlfated NB) across year and age range except on one subset(age 40-49 in 1990) in siblings distribution. However, there is a large difference in best model for white population. A large portion of best model in children distribution for white population is zero-inlfated NB, while negative-binomial is the best model fitted for siblings distribution except for one subset(age 40-49 in 1990). 


## Cohort Stability Analysis Siblings

Analyze the stability of sibling distributions across cohorts, similar to the analysis performed for children.

1. **Zero Inflation**: Check if the proportion of individuals with zero siblings changes significantly over time for the same cohort
2. **Family Size**: Test if the mean or variance in the number of siblings changes for the same cohort over time.
3. **Model Fit**: Analyze if the best-fitting model for the cohort changes over time.

### a. **Zero-Inflation Analysis**
```{r}
best_models_sib <- best_models_sib %>%
  mutate(Cohort = mapply(calculate_cohort, Census_Year, AGE_RANGE)) %>% 
  dplyr::select(Race, Census_Year, AGE_RANGE, Best_Model, Cohort)

df_merg2 <- df2 %>%
  rename(Census_Year = YEAR, Race = RACE)

# Perform a left join to merge the data frames
merged_data_sib <- left_join(df_merg2, best_models_sib, by = c("Race", "Census_Year", "AGE_RANGE"))
```

```{r}
## divided the corhort by race
merged_df_black2 <- merged_data_sib %>% filter(Race == "Black/African American")
merged_df_white2 <- merged_data_sib %>% filter(Race == "White")
```

```{r}
# Filter cohorts with more than one census year
cohort_counts <- merged_df_black2 %>%
  group_by(Cohort) %>%
  summarise(Unique_Census_Years = n_distinct(Census_Year)) %>%
  filter(Unique_Census_Years > 1)

# Get the list of cohorts for testing
cohorts_for_test <- cohort_counts$Cohort

# Initialize results data frame
results_black2 <- data.frame(RACE = character(), Cohort = character(), Chi_Square = numeric(), p_value = numeric(), stringsAsFactors = FALSE)

# Loop through each cohort and run the chi-square test
for (cohort in cohorts_for_test) {
  cohort_data <- merged_df_black2 %>% filter(Cohort == cohort)
  
  # Create contingency table
  contingency_table <- table(cohort_data$Census_Year, cohort_data$n_siblings == "0")
  
  # Perform the chi-square test only if more than one row is in the table
  if (nrow(contingency_table) > 1) {
    test_result <- chisq.test(contingency_table)
    
    # Append results to the results data frame
    results_black2 <- rbind(results_black2, data.frame(
      RACE = "Black/African American",
      Cohort = cohort,
      Chi_Square = test_result$statistic,
      p_value = test_result$p.value
    ))
  }
}

results_black2 <- results_black2 %>%
  mutate(Significance = ifelse(p_value > 0.05, "No", "Yes"))

print(results_black2)
```
**_RESULT:_** The table shows that only the cohort 1921-1930 has significant change in probability of individuals with zero siblings in black population.

```{r}
# Filter cohorts with more than one census year
cohort_counts <- merged_df_white2 %>%
  group_by(Cohort) %>%
  summarise(Unique_Census_Years = n_distinct(Census_Year)) %>%
  filter(Unique_Census_Years > 1)

# Get the list of cohorts for testing
cohorts_for_test <- cohort_counts$Cohort

# Initialize results data frame
results_white2 <- data.frame(RACE = character(), Cohort = character(), Chi_Square = numeric(), p_value = numeric(), stringsAsFactors = FALSE)

# Loop through each cohort and run the chi-square test
for (cohort in cohorts_for_test) {
  cohort_data <- merged_df_white2 %>% filter(Cohort == cohort)
  
  # Create contingency table
  contingency_table <- table(cohort_data$Census_Year, cohort_data$n_siblings == "0")
  
  # Perform the chi-square test only if more than one row is in the table
  if (nrow(contingency_table) > 1) {
    test_result <- chisq.test(contingency_table)
    
    # Append results to the results data frame
    results_white2 <- rbind(results_white2, data.frame(
      RACE = "White",
      Cohort = cohort,
      Chi_Square = test_result$statistic,
      p_value = test_result$p.value
    ))
  }
}

results_white2 <- results_white2 %>%
  mutate(Significance = ifelse(p_value > 0.05, "No", "Yes"))

print(results_white2)
```
**_RESULT:_** The table shows that cohorts 1901-1910, 1911-1920, 1921-1930 have significant change in probability of individuals with zero siblings in white population.

### b. **Family Size Analysis**
```{r}
sum_table_black = merged_data_sib %>% filter(Race == "Black/African American" & n_siblings!=-1)
sum_table_white = merged_data_sib %>% filter(Race == "White"& n_siblings!=-1)
```

```{r}
##Test for Black population
cohorts1 <- unique(sum_table_black$Cohort)

# Loop through each cohort and apply ANOVA for Mean and Variance
for (cohort in cohorts1) {
  
  # Filter data for the specific cohort
  cohort_data_b <- sum_table_black[sum_table_black$Cohort == cohort, ]
  
  # Check if there are at least two unique Census_Year values
  if (length(unique(cohort_data_b$Census_Year)) < 2) {
    cat("\nCohort:", cohort, "has only 1 Census Year. Skipping ANOVA.\n")
    next
  }
  else if (length(unique(cohort_data_b$Census_Year)) == 2){
    t_test_result <- t.test(n_siblings ~ Census_Year, data = cohort_data_b)
     # Output the results
  cat("\nCohort:", cohort)
  cat("\nT-test for Mean Family Size:\n")
  print(t_test_result)
  }
  else{
  # Perform ANOVA for Mean
  anova_mean_result <- aov(n_siblings ~ factor(Census_Year), data = cohort_data_b)
   # Output the results
  cat("\nCohort:", cohort)
  cat("\nANOVA for Mean Family Size:\n")
  print(summary(anova_mean_result))
  }
}
```
**_RESULT:_** The mean number of siblings change significantly in cohorts 1901-1910 and 1921-1930 in black population.

```{r}
##Test for White population
cohorts2 <- unique(sum_table_white$Cohort)

# Loop through each cohort and apply ANOVA for Mean and Variance
for (cohort in cohorts2) {
  
  # Filter data for the specific cohort
  cohort_data_w <- sum_table_white[sum_table_white$Cohort == cohort, ]
  
 # Check if there are at least two unique Census_Year values
  if (length(unique(cohort_data_w$Census_Year)) < 2) {
    cat("\nCohort:", cohort, "has only 1 Census Year. Skipping ANOVA.\n")
    next
  }
  else if (length(unique(cohort_data_w$Census_Year)) == 2){
    t_test_result <- t.test(n_siblings ~ Census_Year, data = cohort_data_w)
     # Output the results
  cat("\nCohort:", cohort)
  cat("\nT-test for Mean Family Size:\n")
  print(t_test_result)
  }
  else{
  # Perform ANOVA for Mean
  anova_mean_result <- aov(n_siblings ~ factor(Census_Year), data = cohort_data_w)
   # Output the results
  cat("\nCohort:", cohort)
  cat("\nANOVA for Mean Family Size:\n")
  print(summary(anova_mean_result))
  }
}
```
**_RESULT:_** The mean number of siblings change significantly in cohorts 1901-1910, 1911-1920, 1921-1930 in white population.

### c. **Model Fit Analysis**
```{r}
ggplot(best_models_sib, aes(x = Census_Year, y = Cohort, fill = Best_Model)) +
  geom_tile(color = "white", lwd = 0.5, linetype = 1) + # Create the tiles
  facet_wrap(~Race, nrow = 2) + # Facet by Race
  scale_fill_manual(values = c("Poisson" = "#0072B2", "Zero_Inflated_NB" = "#F0E442", "Negative_Binomial" = "grey")) + 
  labs(
    title = "Best Model(siblings) by Race, Census Year, and Cohort(Birth Range)",
    x = "Census Year",
    y = "Birth Range",
    fill = "Best Model"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    legend.position = "bottom"
  )
```
**_RESULT:_** The best model for each cohort(those with 1+ corresponding census year) is stable over time.

### Addtional Analysis on Overall Distribution
We also perform additional analysis to see if the overall distribution is stable across census year for the same cohort

```{r}
 df_sibling_proportions <- df_sibling_proportions %>%
  mutate(Cohort = mapply(calculate_cohort, YEAR, AGE_RANGE))
```

```{r}
df_sibling_proportions %>%
  filter(n_siblings >= 0, sibling_count >= 0, proportion >= 0) %>% group_by(RACE, Cohort) %>%
  ggplot(aes(x = n_siblings, y = proportion, color = as.factor(YEAR))) +
  geom_line() +
  labs(
    title = "Sibling Distribution by Census Year",
    x = "Number of Siblings",
    y = "Proportion",
    color = "Census Year"
  ) +
  facet_grid(RACE ~ Cohort) +
  theme_minimal()
```

```{r}
df_sibling_b <- df_sibling_proportions %>% filter(RACE == "Black/African American", n_siblings >= 0, sibling_count >= 0, proportion >= 0)

results_sib_b <- list()

cohort_counts2 <- df_sibling_b %>%
  group_by(Cohort) %>%
  summarise(Unique_Census_Years = n_distinct(YEAR)) %>%
  filter(Unique_Census_Years > 1)

for (cohort in cohort_counts2$Cohort) {
  cohort_data <- df_sibling_b %>% filter(Cohort == cohort)
  
  test_result <- kruskal.test(sibling_count ~ YEAR, data = cohort_data)
  
  results_sib_b[[cohort]] <- test_result
}
results_sib_b
```

```{r}
df_sibling_w <- df_sibling_proportions %>% filter(RACE == "White", n_siblings >= 0, sibling_count >= 0, proportion >= 0)

results_sib_w <- list()

cohort_counts2 <- df_sibling_w %>%
  group_by(Cohort) %>%
  summarise(Unique_Census_Years = n_distinct(YEAR)) %>%
  filter(Unique_Census_Years > 1)

for (cohort in cohort_counts2$Cohort) {
  cohort_data <- df_sibling_w %>% filter(Cohort == cohort)
  
  test_result <- kruskal.test(sibling_count ~ YEAR, data = cohort_data)
  
  results_sib_w[[cohort]] <- test_result
}
results_sib_w

```


#### Summary table showing:
   - RACE
   - birth_range
   - Whether the distribution is stable across census years
   - Any significant changes observed
```{r}
results_summary <- data.frame(
  RACE = character(),
  Cohort = character(),
  Stable_Distribution = character(),
  Significant_Changes = character(),
  stringsAsFactors = FALSE
)

summarize_results <- function(test_results, race) {
  summary_list <- list()
  
  for (birth_range in names(test_results)) {
    test_result <- test_results[[birth_range]]
    
    stable <- ifelse(test_result$p.value >= 0.05, "Yes", "No")
    significant_changes <- ifelse(test_result$p.value < 0.05, "Yes", "No")
    
    summary_list[[birth_range]] <- data.frame(
      RACE = race,
      Cohort = birth_range,
      Stable_Distribution = stable,
      Significant_Changes = significant_changes
    )
  }
  
  do.call(rbind, summary_list)
}

results_summary_black <- summarize_results(results_sib_b, "Black/African American")

results_summary_white <- summarize_results(results_sib_w, "White")

results_summary <- rbind(results_summary, results_summary_black, results_summary_white)

rownames(results_summary) <- NULL 

results_summary

```

#### Visualize stability
```{r}
ggplot(results_summary, aes(x = Cohort, y = RACE, fill = Significant_Changes)) +
  geom_tile(color = "white") +
  scale_fill_manual(values = c("Yes" = "#e93e3a", "No" = "#fdc70c")) +
  labs(title = "Significant Changes in Sibling Count Distribution Across Years by Race and Birth Range",
       x = "Birth Range",
       y = "Race",
       fill = "Significant Changes") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

#### Result
From the plot above, we can see the distribution of sibling is stable in the following cohorts by race:
  -black population:1901-1919, 1911-1920, 1921-1930
  -white population:1901-1910
