---
title: "Short Range Familial Search"
author: "Tina Lasisi"
date: "`r format(Sys.time(), '%Y-%m-%d %H:%M:%S')`"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## CODIS marker allele frequencies

Frequencies and raw genotypes for different populations were found [here](https://www.promega.com/products/pm/genetic-identity/population-statistics/allele-frequencies/) and refer to Steffen, C.R., Coble, M.D., Gettings, K.B., Vallone, P.M. (2017) Corrigendum to 'U.S. Population Data for 29 Autosomal STR Loci' [Forensic Sci. Int. Genet. 7 (2013) e82-e83]. Forensic Sci. Int. Genet. 31, e36–e40. The US core CODIS markers are a subset of the 29 described here. 

```{r setup, include=FALSE}
# Load required libraries
library(readxl)
library(tidyverse)
```


```{r convertxl, eval=FALSE, include=FALSE}

# This chunk converts the Excel file that was downloaded into separate CSVs as there is a bunch of information that is spread across different sheets within the original Excel file.

# Read all sheets
file_path <- "data/1036_allelefreqs.xlsx"
sheet_names <- excel_sheets(file_path)

# Iterate over each sheet
for (sheet in sheet_names) {
  # Read the marker names and sample sizes
  markers <- read_excel(file_path, sheet = sheet, range = cell_rows(2:2), col_names = FALSE)
  sample_sizes <- read_excel(file_path, sheet = sheet, range = cell_rows(3:3), col_names = FALSE)

  # Read the main data (starting from the fourth row)
  data <- read_excel(file_path, sheet = sheet, skip = 3, col_names = FALSE)

  # Set the column names
  colnames(data) <- c("allele", as.character(unlist(markers)))

  # Transform the data to long format
  tidy_data <- data %>%
    pivot_longer(cols = -allele, names_to = "marker", values_to = "frequency") %>%
    rename_with(~tolower(.), everything()) # Convert all variable names to lowercase

  # Remove the rows that contain 'Hobs', 'PI', 'PPE' or 'HWE'
  tidy_data <- tidy_data[!tidy_data$allele %in% c("Hobs", "PI", "PPE", "HWE", "Allele"),] %>% 
    drop_na() %>% 
    select(marker, everything())

  # Extract population from sheet name
  tidy_data$population <- str_extract(sheet, "(?<=revised\\s)[^,]*") # Extract string between "revised " and the first comma
  
  # Clean the name for the CSV file
  filename <- gsub(" ", "", tidy_data$population[1]) # Remove spaces
  
  # Write to CSV
  write_csv(tidy_data, paste0("data/1036_", filename, ".csv")) # Write CSV to the "data" folder
}

```

## Load CODIS allele frequencies

CODIS allele frequencies were found through NIST STR base and specifically downloaded from the supplementary materials of Steffen et al 2017. These are 1036 unrelated individuals from the U.S. population. 

```{r allelefreqs}
# Define the file paths
file_paths <- list.files(path = "data", pattern = "1036_.*\\.csv", full.names = TRUE)

# Create a list of data frames
df_list <- lapply(file_paths, function(path) {
  read_csv(path, col_types = cols(
    marker = col_character(),
    allele = col_double(),
    frequency = col_double(),
    population = col_character()
  ))
})

# Bind all data frames into one
df <- bind_rows(df_list)
```


```{r}
# Find unique markers and split them into chunks of 5
unique_markers <- unique(df$marker)
marker_chunks <- split(unique_markers, ceiling(seq_along(unique_markers)/5))

# Loop through the chunks and create a plot for each chunk
for(i in seq_along(marker_chunks)) {
  df_chunk <- df %>% filter(marker %in% marker_chunks[[i]])
  
  p <- ggplot(df_chunk, aes(x = allele, y = frequency, fill = population)) +
    geom_col(position = "dodge", width = 0.7) +
    facet_grid(population ~ marker) +
    # scale_x_continuous(breaks = seq(2.2, 43.2, by = 1)) +
    labs(x = "Allele", y = "Frequency",
         title = paste("Allele Frequencies for Markers", i*5-4, "to", min(i*5, length(unique_markers))),
         caption = "Source: CSV data") +
    theme_bw()

  print(p)
}



```

```{r fig.width=8}
df %>%
  group_by(population) %>%
  filter(population != "all") %>%
  ggplot(aes(x = marker, y = allele, fill = population)) +
  geom_boxplot() +
  labs(x = "Marker", y = "Allele",
       title = "Range of Alleles Across Different Markers (Grouped by Population)") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


```{r}

# Filter out the "all" population
df_filtered <- df %>% filter(population != "all")

# Get the unique markers and split them into chunks of 5
unique_markers <- unique(df_filtered$marker)
marker_chunks <- split(unique_markers, ceiling(seq_along(unique_markers)/6))

# Loop through the chunks
for(i in seq_along(marker_chunks)) {
  
  # Subset the data for the current chunk of markers
  df_chunk <- df_filtered %>% filter(marker %in% marker_chunks[[i]])
  
  # Convert the allele variable to a factor
  df_chunk$allele <- as.factor(df_chunk$allele)
  
  # Create the plot
  p <- ggplot(df_chunk, aes(x = population, y = frequency, fill = allele)) +
    geom_bar(stat = "identity") +
    facet_wrap(~ marker) +
    labs(x = "Population", y = "Frequency", fill = "Allele",
         title = paste("Allele Frequencies for Markers", i*5-4, "to", min(i*5, length(unique_markers)))) +
    theme_classic()  # Apply theme_classic()
  
  # Print the plot
  print(p)
}
```

```{r}
summary_df <- df %>%
  group_by(marker, population) %>%
  summarise(num_alleles = n_distinct(allele))

summary_df
```

```{r}
summary_df <- df %>%
  group_by(marker, population) %>%
  summarise(num_alleles = n_distinct(allele))

summary_pivot <- summary_df %>%
  pivot_wider(names_from = population, values_from = num_alleles)

summary_pivot
```

## Simulating genotypes

### One marker at a time



```{r func-sim-geno-single}
# Function to simulate genotypes for a pair of individuals
simulate_genotypes <- function(allele_frequencies, relationship_type) {
  # Draw two alleles for the first individual from the population allele frequencies
  individual1 <- sample(names(allele_frequencies), size = 2, replace = TRUE, prob = allele_frequencies)
  
  # Define the probabilities of sharing 0, 1, or 2 IBD alleles for each relationship type
  relationship_probs <- list(
    'parent_child' = c(0, 1, 0), # In a parent-child relationship, always 1 allele is shared
    'full_siblings' = c(1/4, 1/2, 1/4), # For full siblings, the probabilities are 1/4 for sharing 0, 1/2 for sharing 1, and 1/4 for sharing 2 alleles
    'half_siblings' = c(1/2, 1/2, 0), # For half siblings, the probabilities are 1/2 for sharing 0 and 1/2 for sharing 1 allele
    'cousins' = c(7/8, 1/8, 0), # For cousins, the probabilities are 7/8 for sharing 0 and 1/8 for sharing 1 allele
    'second_cousins' = c(15/16, 1/16, 0), # For second cousins, the probabilities are 15/16 for sharing 0 and 1/16 for sharing 1 allele
    'unrelated' = c(1, 0, 0) # For unrelated individuals, they always share 0 alleles
  )
  
  # Get the probabilities of sharing alleles for the specific relationship type
  prob_shared_alleles <- relationship_probs[[relationship_type]]
  
  # Draw the number of shared alleles based on these probabilities
  num_shared_alleles <- sample(c(0, 1, 2), size = 1, prob = prob_shared_alleles)
  
  # Construct the genotype of the second individual by sampling the shared alleles from the first individual 
  # and the rest from the population allele frequencies
  individual2 <- c(sample(individual1, size = num_shared_alleles), sample(names(allele_frequencies), size = 2 - num_shared_alleles, replace = TRUE, prob = allele_frequencies))
  
  # Return the genotypes of the two individuals and the number of shared alleles
  return(list(individual1 = individual1, individual2 = individual2, num_shared_alleles = num_shared_alleles))
}

# Function to calculate the index of relatedness
calculate_relatedness <- function(simulated_genotypes, allele_frequencies) {
  # Calculate the number of alleles that the two individuals share
  num_shared_alleles <- sum(simulated_genotypes$individual1 %in% simulated_genotypes$individual2)
  
  # Calculate the index of relatedness as the number of shared alleles divided by the sum of the inverse of allele frequencies 
  # of the alleles in the genotypes of both individuals. This gives a higher weight to rare alleles.
  relatedness <- num_shared_alleles / (sum(1 / allele_frequencies[simulated_genotypes$individual1]) + sum(1 / allele_frequencies[simulated_genotypes$individual2]))
  
    # Return the index of relatedness and the number of shared alleles
  return(list(relatedness = relatedness, num_shared_alleles = simulated_genotypes$num_shared_alleles))
}

# Function to simulate genotypes and calculate relatedness for different relationships
simulate_relatedness <- function(df, marker, population, relationship_type) {
  # Filter the data frame to get the allele frequencies for the specific marker and population
  allele_frequencies <- df %>%
    filter(marker == marker, population == population) %>%
    pull(frequency) %>%
    setNames(df$allele)
  
  # Simulate the genotypes for the pair of individuals using these allele frequencies and the specific relationship type
  simulated_genotypes <- simulate_genotypes(allele_frequencies, relationship_type)
  
  # Calculate the relatedness index based on these simulated genotypes and the allele frequencies
  relatedness_data <- calculate_relatedness(simulated_genotypes, allele_frequencies)
  
  # Return the marker, population, relationship type, and the calculated relatedness data
  return(c(list(marker = marker, population = population, relationship_type = relationship_type), relatedness_data))
}

# Example usage
# simulate_relatedness(df, marker = "F13A01", population = "Asian", relationship_type = "full_siblings")


```


```{r func-iter-rel-single}
simulate_all_relationships <- function(df, num_simulations) {
  # Define the list of relationship types
  relationship_types <- c('parent_child', 'full_siblings', 'half_siblings', 'cousins', 'second_cousins', 'unrelated')
  
  # Initialize an empty list to store results
  results <- list()
  
  # Iterate over all combinations of markers, populations, and relationship types
  for (marker in unique(df$marker)) {
    for (population in unique(df$population)) {
      for (relationship_type in relationship_types) {
        for (i in 1:num_simulations) {
          # Simulate relatedness and add the result to the list
          results[[length(results) + 1]] <- simulate_relatedness(df, marker, population, relationship_type)
        }
      }
    }
  }
  
  # Convert the list of results to a dataframe
  results_df <- do.call(rbind, lapply(results, function(x) as.data.frame(t(unlist(x)))))
  
  return(results_df)
}
```


```{r df-sim-single}
# Usage:
# df <- # Your dataframe here
results_df <- simulate_all_relationships(df, num_simulations = 10)

```

#### Visualization

```{r func-plt-single}
# Function to capitalize the first letter of a string
ucfirst <- function(s) {
  paste(toupper(substring(s, 1,1)), substring(s, 2), sep = "")
}

create_plot <- function(df, variable_to_plot) {
  # Create the plot
  p <- ggplot(df, aes(x = relationship_type, y = .data[[variable_to_plot]], color = population, shape = population, fill = population)) +
    geom_boxplot(alpha = 0.4) + # Change the order of geom_boxplot() and geom_point() and adjust alpha
    geom_point(position = position_jitterdodge(jitter.width = 0.1, dodge.width = 0.75), size = 1, alpha = 0.6) +
    facet_grid(. ~ marker) +
    theme_classic() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    scale_x_discrete(limits = c('parent_child', 'full_siblings', 'half_siblings', 'cousins', 'second_cousins', 'unrelated')) +
    labs(title = paste(ucfirst(variable_to_plot), "by Relationship Type, Population, and STR Marker"),
         x = "Relationship Type", 
         y = ucfirst(variable_to_plot), 
         color = "Population",
         shape = "Population")
  
  return(p)
}
```

```{r plt-single, fig.width=8}
# Filter your data for the first 3 unique STR markers
factor_vars <- c("marker", "population", "relationship_type")
numeric_vars <- c("relatedness", "num_shared_alleles")

# Filter your data for unique STR markers and remove the "all" population
results_df_filtered <- results_df %>%
  filter(population != "all") %>%
  mutate(
    across(all_of(factor_vars), as.factor),
    across(all_of(numeric_vars), as.numeric)
  )

# Find the total number of unique markers
total_markers <- length(unique(results_df_filtered$marker))

# Iterate through the unique markers in groups of 3
for (i in seq(1, total_markers, by = 3)) {
  # Filter the data for the current set of 3 or fewer markers
  filtered_results_df <- results_df_filtered %>%
    filter(marker %in% unique(marker)[i:min(i+2, total_markers)])
  
  # Create and display the plots for the current set of markers
  for (variable_to_plot in c("relatedness", "num_shared_alleles")) {
    plot <- create_plot(df = filtered_results_df, variable_to_plot = variable_to_plot)
    print(plot) # Print the plot to display it in the RMarkdown document
  }
}
```

### All Markers together

```{r sim-multi}
# Function to simulate genotypes for a pair of individuals
multi_simulate_genotypes <- function(allele_frequencies, relationship_type) {
  # Initialize an empty list to store results
  genotypes <- list()
  
  # Iterate over all markers
  for (marker in names(allele_frequencies)) {
    individual1 <- sample(names(allele_frequencies[[marker]]), size = 2, replace = TRUE, prob = allele_frequencies[[marker]])
    
    # Define the probabilities for sharing 0, 1, or 2 IBD alleles for different types of relationships
    relationship_probs <- list(
      'parent_child' = c(0, 1, 0),
      'full_siblings' = c(1/4, 1/2, 1/4),
      'half_siblings' = c(1/2, 1/2, 0),
      'cousins' = c(7/8, 1/8, 0),
      'second_cousins' = c(15/16, 1/16, 0),
      'unrelated' = c(1, 0, 0)
    )

    # Get the probabilities for sharing 0, 1, or 2 alleles for the specified relationship type    
    prob_shared_alleles <- relationship_probs[[relationship_type]]
    # Sample the number of shared alleles based on these probabilities
    num_shared_alleles <- sample(c(0, 1, 2), size = 1, prob = prob_shared_alleles)
    # Simulate the second individual's alleles based on the number of shared alleles and the marker's population frequency
    individual2 <- c(sample(individual1, size = num_shared_alleles), sample(names(allele_frequencies[[marker]]), size = 2 - num_shared_alleles, replace = TRUE, prob = allele_frequencies[[marker]]))
    
    # Print to trouble shoot
    # print(paste("The value of individual1 is:", individual1))
    # print(paste("The value of individual2 is:", individual2))


    # Store the simulated genotypes for the current marker in the list
    genotypes[[marker]] <- list(individual1 = individual1, individual2 = individual2, num_shared_alleles = num_shared_alleles)
  }
  
  # Return the list of simulated genotypes for all markers
  return(genotypes)
}

# Function to simulate genotypes and calculate relatedness for different relationships across multiple genetic markers
multi_simulate_relatedness <- function(df, population, relationship_type) {
  # Filter the allele frequencies for the marker and population
  # allele_frequencies <- split(df[df$population == population, ], df$marker)
  allele_frequencies <- split(df[df$population == population, ]$frequency, df$marker)
  print(allele_frequencies)


  # Simulate the genotypes for the pair of individuals using these allele frequencies and the specific relationship type
  simulated_genotypes <- multi_simulate_genotypes(allele_frequencies, relationship_type)
  
  # Calculate the relatedness index for each marker based on the simulated genotypes and the allele frequencies
  relatedness_data <- lapply(names(simulated_genotypes), function(marker) {
    calculate_relatedness(simulated_genotypes[[marker]], allele_frequencies[[marker]])
  })
  
  # Convert the list of relatedness data to a dataframe
  relatedness_df <- do.call(rbind, lapply(relatedness_data, function(x) as.data.frame(t(unlist(x)))))
  
  # Add the marker, population, and relationship type to the dataframe
  relatedness_df$marker <- names(relatedness_data)
  relatedness_df$population <- population
  relatedness_df$relationship_type <- relationship_type
  
  return(relatedness_df)
}


```



```{r}
# Function to simulate genotypes for a pair of individuals for all markers
simulate_genotypes_all_markers <- function(df, relationship_type, population) {
  markers <- unique(df$marker)
  
  # Simulate the first individual's alleles by drawing from the population frequency for each marker
  individual1 <- setNames(lapply(markers, function(marker) {
    allele_frequencies <- df %>%
      filter(marker == marker, population == population) %>%
      pull(frequency) %>%
      setNames(df$allele)
      
    sample(names(allele_frequencies), size = 2, replace = TRUE, prob = allele_frequencies)
  }), markers)
  
  # Relationship probabilities
  relationship_probs <- list(
    'parent_child' = c(0, 1, 0),
    'full_siblings' = c(1/4, 1/2, 1/4),
    'half_siblings' = c(1/2, 1/2, 0),
    'cousins' = c(7/8, 1/8, 0),
    'second_cousins' = c(15/16, 1/16, 0),
    'unrelated' = c(1, 0, 0)
  )
  
  prob_shared_alleles <- relationship_probs[[relationship_type]]
  num_shared_alleles <- sample(c(0, 1, 2), size = 1, prob = prob_shared_alleles)
  
  individual2 <- setNames(lapply(markers, function(marker) {
  allele_frequencies <- df %>%
    filter(marker == marker, population == population) %>%
    arrange(marker, allele) %>%
    pull(frequency) %>%
    setNames(df$allele)
  
  prob_shared_alleles <- relationship_probs[[relationship_type]]
  non_zero_indices <- which(prob_shared_alleles != 0)
  num_shared_alleles <- sample(non_zero_indices - 1, size = 1, prob = prob_shared_alleles[non_zero_indices])

  alleles_from_individual1 <- sample(individual1[[marker]], size = num_shared_alleles)
  alleles_from_population <- sample(names(allele_frequencies), size = 2 - num_shared_alleles, replace = TRUE, prob = allele_frequencies)
  
  return(c(alleles_from_individual1, alleles_from_population))
}), markers)

  # Return the simulated genotypes
  return(list(individual1 = individual1, individual2 = individual2))
}



# Function to calculate the index of relatedness for all markers
calculate_relatedness_all_markers <- function(simulated_genotypes, df, population) {
  markers <- names(simulated_genotypes$individual1)
  
  # Calculate the number of shared alleles for each marker
  num_shared_alleles <- sapply(markers, function(marker) {
    sum(simulated_genotypes$individual1[[marker]] %in% simulated_genotypes$individual2[[marker]])
  })
  
  # Calculate the index of relatedness as the number of shared alleles weighted inversely to their frequencies
  # Now considering both individuals' alleles for the inverse frequency weighting
  relatedness <- sapply(markers, function(marker) {
    allele_frequencies <- df %>%
      filter(marker == marker, population == population) %>%
      pull(frequency) %>%
      setNames(df$allele)
      
    num_shared_alleles[marker] / (sum(1 / allele_frequencies[simulated_genotypes$individual1[[marker]]]) + sum(1 / allele_frequencies[simulated_genotypes$individual2[[marker]]]))
  })
  
  # Return the index of relatedness
  return(list(relatedness = relatedness, num_shared_alleles = num_shared_alleles))
  }

# Function to simulate genotypes and calculate relatedness for different relationships for all markers
simulate_relatedness_all_markers <- function(df, relationship_type, population) {
  # Simulate genotypes for all markers
  simulated_genotypes <- simulate_genotypes_all_markers(df, relationship_type, population)
  
  # Calculate and return the relatedness for all markers
  relatedness_data <- calculate_relatedness_all_markers(simulated_genotypes, df, population)
  
   # Before returning the results_df, add marker and population information
  results_df$marker <- rownames(results_df)
  results_df$population <- population
  
  return(relatedness_data)
}


```

```{r eval=FALSE, include=FALSE}
df <- df %>%
  arrange(marker, allele)

results <- simulate_relatedness_all_markers(df, 'parent_child', 'Hispanic')

# print the results
results
```


#### Visualizations
```{r}
# Define the list of relationship types
relationship_types <- c('parent_child', 'full_siblings', 'half_siblings', 'cousins', 'second_cousins', 'unrelated')

# Create a dataframe of all combinations of populations and relationship types
combinations <- expand.grid(population = unique(df$population), relationship_type = relationship_types)

# Apply the function to each combination
results <- combinations %>%
  split(seq(nrow(.))) %>%
  map_dfr(function(combination) {
    population <- combination$population
    relationship_type <- combination$relationship_type
    # cat("Processing:", "population =", population, "; relationship_type =", relationship_type, "\n")
    sim_results <- simulate_relatedness_all_markers(df, relationship_type[[1]], population[[1]])
    
    # Bind resulting data frames
    tibble(
      population = population,
      relationship_type = relationship_type,
      sim_results = list(sim_results)
    )
  })

multi_results <- results %>% 
  mutate(
    sum_relatedness = map_dbl(sim_results, function(x) {
      sum(x[["relatedness"]], na.rm = TRUE)
    }),
    sum_alleles = map_dbl(sim_results, function(x) {
      sum(x[["num_shared_alleles"]], na.rm = TRUE)
    })
  ) %>% 
  select(-sim_results)

multi_results
```


```{r v2-multi}
# Define the number of simulations
n_sims <- 10

# Define the list of relationship types
relationship_types <- c('parent_child', 'full_siblings', 'half_siblings', 'cousins', 'second_cousins', 'unrelated')

# Create a dataframe of all combinations of populations, relationship types, and simulations
combinations <- expand.grid(population = unique(df$population), relationship_type = relationship_types, simulation = 1:n_sims)

# Apply the function to each combination
results <- combinations %>%
  split(seq(nrow(.))) %>%
  map_dfr(function(combination) {
    population <- combination$population
    relationship_type <- combination$relationship_type
    sim <- combination$simulation
    # cat("Processing:", "population =", population, "; relationship_type =", relationship_type, "; simulation =", sim, "\n")
    sim_results <- simulate_relatedness_all_markers(df, relationship_type[[1]], population[[1]])
    
    # Bind resulting data frames
    tibble(
      population = population,
      relationship_type = relationship_type,
      simulation = sim,
      sim_results = list(sim_results)
    )
  })

multi_results <- results %>% 
  mutate(
    sum_relatedness = map_dbl(sim_results, function(x) {
      sum(x[["relatedness"]], na.rm = TRUE)
    }),
    sum_alleles = map_dbl(sim_results, function(x) {
      sum(x[["num_shared_alleles"]], na.rm = TRUE)
    })
  ) %>% 
  select(-sim_results)

multi_results


```

```{r}
create_plot <- function(df, variable_to_plot) {
  # Create the plot
  p <- ggplot(df, aes(x = relationship_type, y = .data[[variable_to_plot]], color = population, shape = population, fill = population)) +
    geom_boxplot(alpha = 0.4, position = position_dodge(width = 0.75)) +
    geom_point(position = position_jitterdodge(jitter.width = 0.1, dodge.width = 0.75), size = 1, alpha = 0.6) +
    theme_classic() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    scale_x_discrete(limits = c('parent_child', 'full_siblings', 'half_siblings', 'cousins', 'second_cousins', 'unrelated')) +
    labs(title = paste(ucfirst(variable_to_plot), "by Relationship Type and Population"),
         x = "Relationship Type", 
         y = ucfirst(variable_to_plot), 
         color = "Population",
         shape = "Population")
  
  return(p)
}

```

```{r fig.width=8}

# Filter your data for unique STR markers and remove the "all" population
df_plt_multi_results <- multi_results %>%
  filter(population != "all") %>%
  select(-simulation)

p <- create_plot(df_plt_multi_results, "sum_relatedness")
p

plt_allele <- create_plot(df_plt_multi_results, "sum_alleles")

plt_allele
```



`

## Probabilities

```{r eval=FALSE, include=FALSE}
# Define relationship probabilities
relationship_probs <- list(
  'parent_child' = c(0, 1, 0),
  'full_siblings' = c(1/4, 1/2, 1/4),
  'half_siblings' = c(1/2, 1/2, 0),
  'cousins' = c(7/8, 1/8, 0),
  'second_cousins' = c(15/16, 1/16, 0),
  'unrelated' = c(1, 0, 0)
)

# Function to calculate the expected number of shared alleles for each relationship
calculate_shared_alleles <- function(df, relationship_type) {
  probs <- relationship_probs[[relationship_type]]
  df %>%
    group_by(marker, population) %>%
    do({
      alleles <- .$allele
      frequencies <- .$frequency
      prob_0 <- prob_1 <- prob_2 <- 0
      for (i in seq_along(alleles)) {
        for (j in seq_along(alleles)) {
          num_shared_alleles <- length(intersect(alleles[i], alleles[j]))
          prob_ij <- frequencies[i] * frequencies[j]
          if (num_shared_alleles == 0) {
            prob_0 <- prob_0 + prob_ij
          } else if (num_shared_alleles == 1) {
            prob_1 <- prob_1 + prob_ij
          } else if (num_shared_alleles == 2) {
            prob_2 <- prob_2 + prob_ij
          }
        }
      }
      data.frame(
        marker = unique(.$marker),
        population = unique(.$population),
        p0 = prob_0 * probs[1],
        p1 = prob_1 * probs[2],
        p2 = prob_2 * probs[3]
      )
    })
}


# Calculate the expected number of shared alleles for each relationship
relationship_types <- c('parent_child', 'full_siblings', 'half_siblings', 'cousins', 'second_cousins', 'unrelated')
results <- lapply(relationship_types, function(rel) calculate_shared_alleles(df, rel))
names(results) <- relationship_types

# Print the results
print(results)

```

```{r eval=FALSE, include=FALSE}
calculate_relationship_probabilities <- function(allele_frequencies, relationship) {
  # Get unique alleles for each marker
  alleles <- unique(allele_frequencies$allele)
  
  # Initialize a list to store the probabilities for each number of IBD alleles
  genotype_probs <- list()
  
  # Calculate the probabilities for each number of IBD alleles
  for (m in 0:2) {
    # Initialize a matrix to store the genotype probabilities
    genotype_probs[[m + 1]] <- matrix(0, nrow = length(alleles), ncol = length(alleles))
    
    for (i in 1:length(alleles)) {
      for (j in i:length(alleles)) {
        if (m == 0) {
          # If m = 0, the alleles are independent, so the probability is the product of the allele frequencies
          genotype_probs[[m + 1]][i, j] <- allele_frequencies[allele_frequencies$allele == alleles[i], "frequency"] * allele_frequencies[allele_frequencies$allele == alleles[j], "frequency"]
        } else if (m == 1) {
          # If m = 1, one allele is shared IBD, so the probability is twice the product of the allele frequencies
          genotype_probs[[m + 1]][i, j] <- 2 * allele_frequencies[allele_frequencies$allele == alleles[i], "frequency"] * allele_frequencies[allele_frequencies$allele == alleles[j], "frequency"]
        } else if (m == 2) {
          # If m = 2, both alleles are shared IBD, so the probability is the square of the allele frequency
          genotype_probs[[m + 1]][i, j] <- allele_frequencies[allele_frequencies$allele == alleles[i], "frequency"]^2
        }
        
        if (i != j) {
          # If the alleles are different, they can be received in either order, so we add the probability to the symmetric position in the matrix
          genotype_probs[[m + 1]][j, i] <- genotype_probs[[m + 1]][i, j]
        }
      }
    }
  }
  
  # Calculate the total probability of observing the pair of genotypes given the relationship as a weighted sum of the probabilities for each number of IBD alleles
  total_probability <- 0
  for (m in 0:2) {
    total_probability <- total_probability + relationship[m + 1] * sum(genotype_probs[[m + 1]])
  }
  
  return(total_probability)
}
```


```{r eval=FALSE, include=FALSE}

calculate_relationship_probabilities(df_test, relationship_probs)
```


```{r calc-v1,  eval=FALSE, include=FALSE}
calculate_conditional_probability <- function(df, relationship_type) {
  
  total_probability <- 0
  for (allele1 in df$allele) {
    for (allele2 in df$allele) {
      
      # The probability that the first individual has the genotype (allele1, allele2)
      prob_genotype_1 <- 2 * df$frequency[df$allele == allele1] * df$frequency[df$allele == allele2] 
      
      if (allele1 != allele2) {
        # If the alleles are different, we have double the possibilities (allele1, allele2) and (allele2, allele1)
        prob_genotype_1 <- 2 * prob_genotype_1
      }
      
      # Now we consider the possible genotypes for the second individual, taking into account the relationship type
      for (shared_allele in c(allele1, allele2)) {
        for (other_allele in df$allele) {
          
          # The chance that the first individual passes on the shared_allele
          prob_pass_on_shared_allele <- 0.5
          
          # The chance that the second individual gets the other allele from the other parent
          prob_get_other_allele <- df$frequency[df$allele == other_allele]
          
          if (relationship_type == 'parent_child' || relationship_type == 'full_siblings') {
            total_probability <- total_probability + prob_genotype_1 * prob_pass_on_shared_allele * prob_get_other_allele
          }
          else if (relationship_type == 'half_siblings') {
            if (allele1 == shared_allele && allele2 == shared_allele) {
              total_probability <- total_probability + prob_genotype_1 * prob_pass_on_shared_allele * prob_get_other_allele
            }
          }
          else if (relationship_type == 'cousins' || relationship_type == 'second_cousins') {
            if (allele1 == shared_allele && allele2 == shared_allele && allele1 == other_allele) {
              total_probability <- total_probability + prob_genotype_1 * prob_pass_on_shared_allele * prob_get_other_allele
            }
          }
          
        }
      }
      
    }
  }
  
  return(total_probability)
  
}

```

```{r calc-v2,  eval=FALSE, include=FALSE}
calculate_conditional_probability <- function(df, relationship_type) {
  
  # Create a lookup table to quickly find allele frequencies
  frequency_lookup <- df$frequency
  names(frequency_lookup) <- df$allele
  
  # Get the allele pairs
  allele_pairs <- expand.grid(df$allele, df$allele)
  
  # Calculate the genotype probabilities for the first individual
  prob_genotype_1 <- frequency_lookup[allele_pairs$Var1] * frequency_lookup[allele_pairs$Var2]
  
  # Double the probabilities for pairs with different alleles
  prob_genotype_1[allele_pairs$Var1 != allele_pairs$Var2] <- 2 * prob_genotype_1[allele_pairs$Var1 != allele_pairs$Var2]
  
  total_probability <- 0
  
  # Iterate over each possible shared allele
  for (shared_allele in unique(c(allele_pairs$Var1, allele_pairs$Var2))) {
    
    # Calculate the probability of passing on the shared allele
    prob_pass_on_shared_allele <- 0.5
    
    # Calculate the probability of the second individual getting each other allele
    prob_get_other_allele <- frequency_lookup
    
    # Determine which allele pairs share the current allele
    shares_allele <- allele_pairs$Var1 == shared_allele | allele_pairs$Var2 == shared_allele
    
    # Calculate the conditional probability for each relationship type
    if (relationship_type == 'parent_child' || relationship_type == 'full_siblings') {
      total_probability <- total_probability + sum(prob_genotype_1[shares_allele] * prob_pass_on_shared_allele * prob_get_other_allele)
    }
    else if (relationship_type == 'half_siblings') {
      total_probability <- total_probability + sum(prob_genotype_1[shares_allele & allele_pairs$Var1 == allele_pairs$Var2] * prob_pass_on_shared_allele * prob_get_other_allele)
    }
    else if (relationship_type == 'cousins' || relationship_type == 'second_cousins') {
      total_probability <- total_probability + sum(prob_genotype_1[shares_allele & allele_pairs$Var1 == shared_allele & allele_pairs$Var2 == shared_allele] * prob_pass_on_shared_allele * prob_get_other_allele)
    }
  }
  
  return(total_probability)
  
}



```



```{r eval=FALSE, include=FALSE}
# List of relationship types
relationship_types <- c('parent_child', 'full_siblings', 'half_siblings', 'cousins', 'second_cousins', 'unrelated')


# Function to calculate conditional probabilities for all combinations
# generate_conditional_probabilities <- function(df, relationship_types) {
#   # Replace with your logic to calculate the conditional probabilities
#   prob <- calculate_conditional_probability(df, relationship_types)
#   
#   # Return the result as a tibble
#   tibble(prob = prob)
# }

generate_conditional_probabilities <- function(df, relationship_types) {
  # Loop over relationship types and calculate the probability for each
  probs <- purrr::map_dfr(relationship_types, ~calculate_conditional_probability(df, .x) %>% 
                            tibble(relationship_type = .x, prob = .))
  return(probs)
}

```


```{r test,  eval=FALSE, include=FALSE}

# test set
df_test <- df %>% 
  filter(marker == "Penta_D")

# Generate the data
# conditional_probabilities <- df_test %>%
#   group_by(marker, population) %>%
#   nest(data = c(allele, frequency)) %>%
#   mutate(probabilities = map(data, ~ generate_conditional_probabilities(.x, relationship_types))) %>%
#   unnest

conditional_probabilities <- df_test %>%
  group_by(marker, population) %>%
  nest(data = c(allele, frequency)) %>%
  mutate(probabilities = map(data, ~ generate_conditional_probabilities(.x, relationship_types))) %>%
  select(-data) %>% 
  unnest(cols = probabilities)


conditional_probabilities
```


```{r eval=FALSE, include=FALSE}
# Plot the data
ggplot(conditional_probabilities, aes(x = relationship, y = prob, color = population)) +
  geom_line() +
  facet_wrap(~marker) +
  labs(x = "Relationship type", y = "Probability of sharing at least one allele",
       title = "Conditional probability of sharing at least one allele for different relationships and populations",
       color = "Population") +
  theme_minimal()

```

