---
title: "Disparities in Short-Range Familial Searches"
author: "Junhui He"
date: "`r Sys.Date()`"
output: workflowr::wflow_html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

# load package
library(ggplot2)
library(ggpubr)
```

## 1 Objective

We aims to examines disparities in the probability of detecting an individual through short-range familial searches using forensic DNA databases for Black and White Americans. This analysis integrates forensic database sizes, forensic database representation (from Result 1), family size distributions (from Result 2), and considers different levels of true positive rates in determining a familial relationship of a particular kind.

## 2 Model assumptions

1.  The population size is $N$, and the racial proportion of the population is $\alpha$.

2.  The database has $K$ individuals, and the racial proportion of the database is $\beta$. For a given race, the database is randomly sampled from the current population. Both the population proportion $\alpha$ and the database proportion $\beta$ are considered fixed constants.

3.  The family trees of different races are strictly separated. For instance, the parents and siblings of black Americans are black Americans, while the parents and siblings of white Americans are white Americans.

4.  The distribution of the children number is estimated based on the distribution of the number of children born to women aged 40-49 in 1990.

5.  The distribution of the sibling number is estimated based on the distribution of the number of children born to women aged 50-59 in 1990. In particular, the frequency of siblings is calculated as $freq_{n_{sib}} = freq_{mother} * chborn_{num}$.

6.  All states in US follow the same distribution of children and sibling numbers.

7.  In the short-range familial search, we only consider the parent-kids and full siblings for relationship detection.

8.  All relatives involved are included in the same population as the individual. For example, if a person lives in Michigan, their parents and siblings also live in Michigan.

## 3 Hypothesis testing in determining a familial relationship with forensic STR loci

The short-range familial search utilizes a hypothesis testing to determine a familial relationship of a particular kind. Specifically, for a pair of individuals, consider the following null and alternative hypothesis: $$H_0:\text{unrelated pairs}\leftrightarrow H_1:  \text{exist a certain familial relationship}.$$ The kinds of familial relationship involved in short-range familial search contain the individual, parents and full siblings.

When the alternative hypothesis is accepted, there is supposed to exist a certain kind of relationship between the pair of individuals, and we call it a positive test. There are two possibilities for a positive test:

1.  False positive: The two individuals are unrelated in fact, but the test claims they are related mistakenly.
2.  True positive: The two individuals are related and the test detects their relationship correctly.

In this analysis, we denote the True Positive Rate as TPR, and denote the False Positive Rate as FPR. The short-range familial search conducts a likelihood ratio test (LRT) to detect the relationship. The performance of LRT will depend on many factors, including the familial relationshiop considered in the test, the number of STR loci, population allele frequency differences and cutoff thresholds. See Result 4 for further discussions.

## 4 The detection probability

### 4.1 Definition

Before our formal derivation, we have to define the concept of a detection. A detection, or a match is the decision that we accept the alternative hypothesis in the likelihood ratio test. Consequently, a detection is equivalent to a positive test. A detection contains two possible cases: (1) correct detection, corresponding to true positive; (2) incorrect detection, corresponding to false positive.

### 4.2 Derivation

We give a step-by-step derivation for the detection probability in a database.

1.  For any random person, the probability of the event {A person is present in a database} is $P(present)=\frac{\beta K}{\alpha N}$, where $K$ is the database size, $N$ is the population size, $\beta$ and $\alpha$ and racial proportions in the database and population.

2.  Suppose the individual has $n$ relatives. The probability of the event {$r$ relatives present in the database \| $n$ relatives} is $Bin(r;n,P(present))={n \choose r} P(present)^r(1-P(present))^{n-r}$.

3.  Suppose $r$ relatives are present in the database, then $K-r$ unrelated people are present in the database. The number of correct detection and incorrect detection in the database follows $Bin(r, TPR)$ and $Bin(K-r, FPR)$, respectively. Thus, the probability of the event {at least one detection \| $r$ relatives present in the database, $K-r$ unrelated people present in the database} is $1 - (1-TPR)^r(1-FPR)^{K-r}$.

4.  For a given short-range familial relationship, the probability of the event {at least one detection of a certain relationship} is $$P(at~least~one~detection~of~a~certain~relationship)=\\\sum_{n=1}^{n_\max}P(n~relatives)\sum_{r=0}^nBin(r;n,P(present))\{1 - (1-TPR)^r(1-FPR)^{K-r}\}.$$

5.  The probability of the event {at least one detection} is $$1-\prod_{i=1}^{i_\max}(1-P(at~least~one~detection~of~relationship~i)).$$

## 5 Experimental Setting

We focus on short-range familial search in the United States, and will calculate the detection probabilities for Black and White Americans state-by-state.

```{r data}
data_path = "./data"

if(FALSE) {
# Read in dataframe for Murphy and Tong profiles
murphy.tong = read.csv(file.path(data_path, "/murphy_foia_cleaned.csv"))

# Read CODIS data by race
murphy.tong.codis <- murphy.tong[which(grepl('_annual$', murphy.tong$variable_detailed) & murphy.tong$value_type == "number"),
                                  c("State","variable_detailed", "value")]

# Make sure column names and values are consistent between data frames
colnames(murphy.tong.codis) = c("State","Race","CODIS_value")
murphy.tong.codis$Race = gsub('.{7}$', '', murphy.tong.codis$Race)

census.population = read.csv(file.path(data_path, 'regression_data', 'CODIS_regression_predict_data.csv'))

# Combine CODIS and Census data
codis_census = data.frame("State" = census.population$State, "census.total" = census.population$census.total, "census.black" = census.population$census.total*census.population$census.percent.black, "census.white" = census.population$census.total*census.population$census.percent.white, "codis.total" = c(by(murphy.tong.codis$CODIS_value, murphy.tong.codis$State, sum)), "codis.black" = murphy.tong.codis[murphy.tong.codis$Race=="black", "CODIS_value"], "codis.white" = murphy.tong.codis[murphy.tong.codis$Race=="white", "CODIS_value"])
}

# Read data
predictor_data = read.csv(file.path(data_path, 'regression_data', 'CODIS_regression_predict_data.csv'))
response_data = read.csv(file.path(data_path, 'regression_data', 'CODIS_regression_data.csv'))

codis_proportion_data = read.csv(file = file.path(data_path, 'regression_data', 'CODIS_predicted_racial_proportions.csv'))

codis_size_data = read.csv(file.path(data_path, "regression_data", "CODIS_predicted_total_number.csv"))
codis.total = codis_size_data$N_total.SDIS
codis.total[is.na(codis.total)] = codis_size_data$N_total.SDIS.prediction[is.na(codis.total)]
idx = c(5, 9, 14, 19, 28, 41, 43)
codis.total[idx] = response_data$mt.total

# Combine CODIS and Census data
codis_census = data.frame("State" = predictor_data$State, "census.total" = predictor_data$census.total, "census.black" = predictor_data$census.total*predictor_data$census.percent.black, "census.white" = predictor_data$census.total*predictor_data$census.percent.white, "codis.total" = codis.total, "codis.black" = codis.total*codis_proportion_data$mt.percent.black, "codis.white" = codis.total*codis_proportion_data$mt.percent.white)

# black_sibling_distribution = c(0.03567456, 3.306873, 3.069263) # zero-inflated negative binomial
# white_sibling_distribution = c(2.227449, 13.35249) # negative binomial
```

```{r short-range-search}
detect_prob <- function(N, K, r, TPR, FPR) {
  p_present = K / N
  tot_p = 0
  
  if(r>=1) {
    for(k in 0:r) {
      p = dbinom(k, r, p_present) * (1 - (1 - TPR)^k*(1 - FPR)^(K-k))
      tot_p = tot_p + p
    }
  }
  
  return(tot_p)
}

detect_prob_dist <- function(N, K, dist_vect, TPR, FPR) {
  
  tot_p = 0
  
  r = length(dist_vect)
  for (i in 1:r) {
    p_size = dist_vect[i]
    tot_p = tot_p + p_size * detect_prob(N, K, i-1, TPR, FPR)
  }
  
  return(tot_p)
}

# N: the population size of a certain race
# K: the forensic database size of a certain race
# dist_vect: the distribution vector
# TPR: true positive rate
# FPR: false positive rate
```

Our analysis examines how family size distributions, database representation and true/false positive rate influence the probability of detecting an individual in a database.

### 5.1  Family size distributions

The distributions of number of children and siblings are given in the following figures, where White American shown on left (negative values), Black American on right (positive values).

```{r sibling-number, fig.width=15, fig.height=7}
# read family size
family_size = read.csv(file.path(data_path, "family_size.csv"))

black_family_size = family_size$proportion[which(family_size$RACE=="Black/African American" & family_size$YEAR=="1990" & family_size$AGE_RANGE=="50-59")]

white_family_size = family_size$proportion[which(family_size$RACE=="White" & family_size$YEAR=="1990" & family_size$AGE_RANGE=="50-59")]

black_child_size = family_size$proportion[which(family_size$RACE=="Black/African American" & family_size$YEAR=="1990" & family_size$AGE_RANGE=="40-49")]

white_child_size = family_size$proportion[which(family_size$RACE=="White" & family_size$YEAR=="1990" & family_size$AGE_RANGE=="40-49")]

df_child = data.frame(n_children = rep(0:12, 2), proportions = c(black_child_size, -white_child_size), race = as.factor(rep(c("Black American", "White American"), each=13)))

# calculate sibling size
black_sibling_size = black_family_size[-1] * c(1:12) / sum(black_family_size[-1] * c(1:12))
white_sibling_size = white_family_size[-1] * c(1:12) / sum(white_family_size[-1] * c(1:12))

#black_sibling_size = black_family_size[-1]/sum(black_family_size[-1])
#white_sibling_size = white_family_size[-1]/sum(white_family_size[-1])

df_sibling = data.frame(n_siblings = rep(0:11, 2), proportions = c(black_sibling_size, -white_sibling_size), race = as.factor(rep(c("Black American", "White American"), each=12)))

my_colors <- colorRampPalette(c("#FFB000", "#F77A2E", "#DE3A8A", "#7253FF", "#5E8BFF"))(13)

p1 = ggplot(data = df_child, mapping = aes(x=n_children, y=proportions, fill=as.factor(n_children))) + geom_col(aes(alpha = race)) + geom_hline(yintercept = 0, color = "black", linewidth = 0.5) + scale_y_continuous(
   labels = function(x) abs(x),
   limits = function(x) c(-max(abs(x)), max(abs(x)))
 ) + coord_flip() +
 scale_x_continuous(breaks = 0:12, labels = c(0:11, "12+")) +
 scale_fill_manual(values = my_colors) +
 scale_alpha_manual(values = c("White American" = 0.7, "Black American" = 1)) +
 labs(
   title = "Distribution of Number of Children",
   x = "Number of Children",
   y = "Proportion",
   fill = "Number of Children"
 ) +
 theme_minimal() +
 theme(
   plot.title = element_text(size = 14, hjust = 0.5),
   axis.text.y = element_text(size = 8),
   strip.text = element_text(size = 10),
   legend.position = "none"
 )

p2 = ggplot(data = df_sibling, mapping = aes(x=n_siblings, y=proportions, fill=as.factor(n_siblings))) + geom_col(aes(alpha = race)) + geom_hline(yintercept = 0, color = "black", linewidth = 0.5) + scale_y_continuous(
   labels = function(x) abs(x),
   limits = function(x) c(-max(abs(x)), max(abs(x)))
 ) + coord_flip() +
 scale_x_continuous(breaks = 0:11, labels = c(0:10, "11+")) +
 scale_fill_manual(values = my_colors) +
 scale_alpha_manual(values = c("White American" = 0.7, "Black American" = 1)) +
 labs(
   title = "Distribution of Number of Siblings",
   x = "Number of Siblings",
   y = "Proportion",
   fill = "Number of Siblings"
 ) +
 theme_minimal() +
 theme(
   plot.title = element_text(size = 14, hjust = 0.5),
   axis.text.y = element_text(size = 8),
   strip.text = element_text(size = 10),
   legend.position = "none"
 )

ggarrange(p1, p2)
```

```{r comments1}
# Following Result 2, we use a negative binomial distribution to model the number of siblings for White Americans, and use a zero-inflated negative binomial distribution to model the number of siblings for Black Americans. In calculating the detection probability, we truncate the number of siblings at $20$.
```

### 5.2  Database representation

We use Census data as the population, and use CODIS data as the forensic database. In particular, we run regression to predict the number of people in CODIS by state and race. The state-level population and forensic database size by race is given as follows:

```{r datasize, include=TRUE}
rownames(codis_census) = NULL
knitr::kable(codis_census)
```

Furthermore, we show the present probability for Black and White Americans in the following figure.

```{r plot-proportion, include=TRUE}
ggplot(data = codis_census) + geom_col(mapping = aes(x=State, y=codis.black/census.black), position = position_dodge()) + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1), axis.title.x = element_blank(), plot.title = element_text(hjust = 0.5, size = 10), legend.position = 'bottom') + ylab('P(present)') + ylim(0,1) + ggtitle('State-level database size/population size for Black Americans', subtitle = "(The present probability for Black Americans)")

ggplot(data = codis_census) + geom_col(mapping = aes(x=State, y=codis.white/census.white), position = position_dodge()) + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1), axis.title.x = element_blank(), plot.title = element_text(hjust = 0.5, size = 10), legend.position = 'bottom') + ylab('P(present)') + ylim(0,1) + ggtitle('State-level database size/population size for White Americans', subtitle = '(The present probability for White Americans)')
```

### 5.3  True/False positive rate

The relationship between true and false positive rate is super complex in short-range familial search. It could depend on the hypothesis type, the likelihood cutoff, the number of STR loci, the race and so on. See [Likelihood Ratio Distribution](https://lasisilab.github.io/PODFRIDGE/lr_dist.html) for an overview.

In this analysis, we consider a special case of true/false positive rates. For parent-child hypothesis, TPR is set to 1 and FPR is set to 0.01%. For full-sibling hypothesis, TPR is set to 75% and FPR is set to 1%.

## 6 Results

In this section, we present State-level estimated detection probabilities.

```{r parameter}
TPR = c(1, 1, 0.75)
FPR = c(0.0001, 0.0001, 0.01)/1e5
P_pres = seq(0, 0.1, length.out = 50)
```

### 6.1 Detection probability versus present probability

```{r detect-present}
pr_black_pres = array(NA, c(4, length(P_pres)))
pr_white_pres = array(NA, c(4, length(P_pres)))

for (i in 1:length(P_pres)) {
  pr_black_pres[1,i] = detect_prob(1, P_pres[i], 2, TPR[1], FPR[1])
  pr_black_pres[2,i] = detect_prob_dist(1, P_pres[i], black_child_size, TPR[2], FPR[2])
  pr_black_pres[3,i] = detect_prob_dist(1, P_pres[i], black_sibling_size, TPR[3], FPR[3])
  
  pr_white_pres[1,i] = detect_prob(1, P_pres[i], 2, TPR[1], FPR[1])
  pr_white_pres[2,i] = detect_prob_dist(1, P_pres[i], white_child_size, TPR[2], FPR[2])
  pr_white_pres[3,i] = detect_prob_dist(1, P_pres[i], white_sibling_size, TPR[3], FPR[3])
}

pr_black_pres[4,] = apply(pr_black_pres[1:3,], 2, function(x) 1-prod(1-x))
pr_white_pres[4,] = apply(pr_white_pres[1:3,], 2, function(x) 1-prod(1-x))

black_df_pres = data.frame(prob=c(pr_black_pres), prop=rep(P_pres, each=4), cousin=factor(rep(c("parent", "child", "full sibling", "total"), length(P_pres))), race=factor("Black American"))
white_df_pres = data.frame(prob=c(pr_white_pres), prop=rep(P_pres, each=4), cousin=factor(rep(c("parent", "child", "full sibling", "total"), length(P_pres))), race=factor("White American"))
prob_df_pres = rbind(black_df_pres, white_df_pres)

ggplot(prob_df_pres, aes(x=prop, y=prob)) + geom_line(aes(colour=cousin, linetype=race)) + xlab("P(present)") + ylab("detection probability") + ggtitle("The relationship of detection probability versus present probability") + ylim(0, 0.7)
```


```{r detect-probability}
pr_black = array(NA, dim = c(nrow(codis_census), 4))
pr_white = array(NA, dim = c(nrow(codis_census), 4))

for(i in 1:nrow(codis_census)) {
  pr_black[i,1] = detect_prob(codis_census[i, "census.black"], codis_census[i, "codis.black"], 2, TPR[1], FPR[1]) # parent
  pr_black[i,2] = detect_prob_dist(codis_census[i, "census.black"], codis_census[i, "codis.black"], black_child_size, TPR[2], FPR[2]) # child
  pr_black[i,3] = detect_prob_dist(codis_census[i, "census.black"], codis_census[i, "codis.black"], black_sibling_size, TPR[3], FPR[3]) # full sibling
  
  pr_white[i,1] = detect_prob(codis_census[i, "census.white"], codis_census[i, "codis.white"], 2, TPR[1], FPR[1]) # parent
  pr_white[i,2] = detect_prob_dist(codis_census[i, "census.white"], codis_census[i, "codis.white"], white_child_size, TPR[2], FPR[2]) # child
  pr_white[i,3] = detect_prob_dist(codis_census[i, "census.white"], codis_census[i, "codis.white"], white_sibling_size, TPR[3], FPR[3]) # full sibling
}

pr_black[,4] = apply(pr_black[,1:3], 1, function(x) 1-prod(1-x))
pr_white[,4] = apply(pr_white[,1:3], 1, function(x) 1-prod(1-x))

pr_black_df = data.frame(pr_black)
colnames(pr_black_df) = c("parent", "child", "full_sibling", "total")
pr_white_df = data.frame(pr_white)
colnames(pr_white_df) = c("parent", "child", "full_sibling", "total")
```

### 6.2 Total detection probability

```{r plot-total, include=FALSE}

ggplot() + geom_col(mapping = aes(x=codis_census$State, y=pr_black_df$total), position = position_dodge()) + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1), axis.title.x = element_blank(), plot.title = element_text(hjust = 0.5, size = 10), legend.position = 'bottom') + ylab('Detection probability') + ylim(0,1) + ggtitle('State-level detection probability for Black Americans')

ggplot() + geom_col(mapping = aes(x=codis_census$State, y=pr_white_df$total), position = position_dodge()) + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1), axis.title.x = element_blank(), plot.title = element_text(hjust = 0.5, size = 10), legend.position = 'bottom') + ylab('Detection probability') + ylim(0,1) + ggtitle('State-level detection probability for White Americans')

ggplot() + geom_col(mapping = aes(x=codis_census$State, y=pr_black_df$total-pr_white_df$total), position = position_dodge()) + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1), axis.title.x = element_blank(), plot.title = element_text(hjust = 0.5, size = 10), legend.position = 'bottom') + ylab('Detection probability')  + ggtitle('Difference of state-level detection probabilities between Black and White Americans')
```

### 6.3 Parent, child, full sibling detection probability

```{r plot-separate}
# parent
ggplot() + geom_col(mapping = aes(x=codis_census$State, y=pr_black_df$parent), position = position_dodge()) + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1), axis.title.x = element_blank(), plot.title = element_text(hjust = 0.5, size = 10), legend.position = 'bottom') + ylab('Detection probability') + ylim(0,1) + ggtitle('State-level parent detection probability for Black Americans')

ggplot() + geom_col(mapping = aes(x=codis_census$State, y=pr_white_df$parent), position = position_dodge()) + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1), axis.title.x = element_blank(), plot.title = element_text(hjust = 0.5, size = 10), legend.position = 'bottom') + ylab('Detection probability') + ylim(0,1) + ggtitle('State-level parent detection probability for White Americans')

ggplot() + geom_col(mapping = aes(x=codis_census$State, y=pr_black_df$parent-pr_white_df$parent), position = position_dodge()) + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1), axis.title.x = element_blank(), plot.title = element_text(hjust = 0.5, size = 10), legend.position = 'bottom') + ylab('Detection probability')  + ggtitle('Difference of state-level parent detection probabilities between Black and White Americans')

# child
ggplot() + geom_col(mapping = aes(x=codis_census$State, y=pr_black_df$child), position = position_dodge()) + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1), axis.title.x = element_blank(), plot.title = element_text(hjust = 0.5, size = 10), legend.position = 'bottom') + ylab('Detection probability') + ylim(0,1) + ggtitle('State-level child detection probability for Black Americans')

ggplot() + geom_col(mapping = aes(x=codis_census$State, y=pr_white_df$child), position = position_dodge()) + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1), axis.title.x = element_blank(), plot.title = element_text(hjust = 0.5, size = 10), legend.position = 'bottom') + ylab('Detection probability') + ylim(0,1) + ggtitle('State-level child detection probability for White Americans')

ggplot() + geom_col(mapping = aes(x=codis_census$State, y=pr_black_df$child-pr_white_df$child), position = position_dodge()) + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1), axis.title.x = element_blank(), plot.title = element_text(hjust = 0.5, size = 10), legend.position = 'bottom') + ylab('Detection probability')  + ggtitle('Difference of state-level child detection probabilities between Black and White Americans')

# full sibling
ggplot() + geom_col(mapping = aes(x=codis_census$State, y=pr_black_df$full_sibling), position = position_dodge()) + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1), axis.title.x = element_blank(), plot.title = element_text(hjust = 0.5, size = 10), legend.position = 'bottom') + ylab('Detection probability') + ylim(0,1) + ggtitle('State-level full sibling detection probability for Black Americans')

ggplot() + geom_col(mapping = aes(x=codis_census$State, y=pr_white_df$full_sibling), position = position_dodge()) + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1), axis.title.x = element_blank(), plot.title = element_text(hjust = 0.5, size = 10), legend.position = 'bottom') + ylab('Detection probability') + ylim(0,1) + ggtitle('State-level full sibling detection probability for White Americans')

ggplot() + geom_col(mapping = aes(x=codis_census$State, y=pr_black_df$full_sibling-pr_white_df$full_sibling), position = position_dodge()) + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1), axis.title.x = element_blank(), plot.title = element_text(hjust = 0.5, size = 10), legend.position = 'bottom') + ylab('Detection probability')  + ggtitle('Difference of state-level full sibling detection probabilities between Black and White Americans')
```

## 7 Conclusions

1.  The detection probability for Black Americans is higher than that for White Americans in most states. This is because (1) Black Americans have mildly more children and siblings than White Americans, and (2) Black Americans are overrepresented while White Americans are underrepresented in the forensic database.

```{r comments2}
# 2.  The situation is opposite in Florida, where Black Americans are underrepresented while White Americans are overrepresented in the forensic database.
```
