---
title: "Siblings analysis"
author: "Manqing Lin"
date: "`r format(Sys.time(), '%Y-%m-%d %H:%M:%S')`"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

# Introduction

The sibling analysis aims to provide an overview of pattern about family size. By examining the number of children a mother has and its relationship with the number of sibling each individual has, we know the number of first degree relatives each individual has. Therefore, it is helpful to predict the probability of individual to be detected in the forensic database. 

# Data Exploration
```{r setup, include=FALSE, echo=FALSE}
#knitr::opts_knit$set(root.dir = ".")
knitr::opts_chunk$set(eval = TRUE, echo = FALSE, warning = FALSE, fig.width = 7, fig.height = 6)


# load necessary libraries
library(readr)
library(tidyverse)
library(tidyr)
library(ggplot2)
library(viridis)
library(dplyr)
library(knitr)
library(scales)
library(rempsyc)
```


```{r data, results = FALSE,warning=FALSE,message=FALSE}
children_data = read.csv("./data/proportions_table_by_race_year.csv")
mother_data = read.csv("./data/data_filtered_recoded.csv")
```


## Frequency of Number of Children & Sibling of Black/African American
Here, we are trying to create a frequency table for each race to show the number of children born, the number of mother who has "chborn_num" children, the number of sibling, and the number of individuals who has "n_sib" siblings. 
$$
n_{sib} = chborn_{num} - 1
$$

$$
\text{freq}_{n_{\text{sib}}} = \text{freq}_{\text{mother}} \cdot \text{chborn}_{\text{num}}
$$

For example: Suppose 10 mothers (generation 0) has 7 children, then there will be 70 children (generation 1) in total who has 6 siblings.

```{r}
fert_df <- mother_data %>%
  count(RACE, YEAR, chborn_num) %>%
  arrange(RACE, YEAR, chborn_num) %>% 
  mutate(chborn_num = chborn_num, 
    n_sib = as.integer(if_else(chborn_num == 0, 0, chborn_num - 1)),
    freq_n_sib = n * chborn_num) %>%
  rename(freq_mother = n)

nice_table(
  fert_df[fert_df$RACE == "Black/African American", ],
  title = c("Table 1", "Frequency of each Number of Children & Sibling of Black/African American")
)
```

## Frequency of Number of Children & Sibling of White
```{r}
nice_table(
  fert_df[fert_df$RACE == "White", ],
  title = c("Table 2", "Frequency of each Number of Children & Sibling of White")
)
```


## Summary Statistics of Number of Sibling of Black/African American
```{r}
fert_df_sum <- fert_df%>%group_by(RACE, YEAR) %>%
  summarise(
    Mean = sum(freq_n_sib) / sum(freq_mother),
    Variance = sum(freq_mother * (n_sib - Mean)^2) / (sum(freq_mother) - 1), 
    Std_Dev = sqrt(Variance)
    )

nice_table(
  fert_df_sum[fert_df_sum$RACE == "Black/African American", ],
  title = c("Table 3", "Summary Statistics of each Number of Sibling of Black/African American")
)
```


## Summary Statistics of Number of Sibling of White

```{r}
nice_table(
  fert_df_sum[fert_df_sum$RACE == "White", ],
  title = c("Table 4", "Summary Statistics of each Number of Sibling of White")
)
```


```{r chidren, results = FALSE,warning=FALSE,message=FALSE}
ggplot(data = children_data,aes(x=chborn_num,y=proportion)) +
  geom_col(position = position_dodge()) +
  facet_grid(RACE~YEAR) + theme_classic()
```

# Data Visualization
```{r,warning=FALSE,out.width="100%"}
mother_data <- mother_data %>% mutate(id = row_number())
child_df = as.data.frame(rep(mother_data$id,mother_data$chborn_num))
colnames(child_df) = "id"
child_df = merge(child_df,mother_data,by="id")
child_df$nsib = child_df$chborn_num - 1

child_df_gb = child_df %>%
  group_by(RACE,nsib) %>%
  summarise(count = n()) 

child_df_gb$prob = ifelse(child_df_gb$RACE == "White",child_df_gb$count/4380009, child_df_gb$count/547692)

ggplot(data = child_df_gb, aes(x = nsib, y = prob)) +
  geom_col(position = position_dodge()) +
  facet_wrap(~RACE) + theme_classic()
```

 
```{r}
colors <- c(
  "#8A9A5B",  # Sage
  "#355E3B",  # Hunter Green
  "#5F8575",  # Eucalyptus
  "#66C2A5",  # Teal
  "#088F8F",  # Blue Green
  "#b2d8d8",  # Robin Egg
  "#93C572",  # Pistachio
  "#454B1B",  # Army Green
  "#5F9EA0",  # Cadet Blue
  "#00A36C",  # Jade
  "#9FE2BF",  # Seafoam
  "#2E8B57",  # Sea Green
  "#40B5AD"   # Verdigris

)
unique_chborn_num <- sort(unique(mother_data$chborn_num))
color_mapping <- setNames(colors, unique_chborn_num)

fert_df <- mother_data %>%
  count(RACE, YEAR, chborn_num) %>%
  arrange(RACE, YEAR, chborn_num) %>% 
  mutate(n_sib = chborn_num -1, freq_n_sib = n * chborn_num, 
         color = color_mapping[as.character(chborn_num)]) %>%
  rename(freq_mother = n)

print(fert_df)
```

```{r}
fert_df_black <- fert_df %>% filter(RACE == "Black/African American")
fert_df_white <- fert_df %>% filter(RACE == "White")
  
df_long_black <- fert_df_black %>%
  pivot_longer(cols = c(freq_mother, freq_n_sib),
               names_to = "frequency_type",
               values_to = "frequency") %>%
  mutate(x_axis = ifelse(frequency_type == "freq_mother", chborn_num, n_sib), 
         color.label = paste(chborn_num, "_kids"," ", n_sib, "_sibs", sep = ""), 
         color_order = factor(chborn_num, levels = unique_chborn_num)
  )

```


```{r}
ggplot(df_long_black, aes(x = x_axis, y = frequency, fill = color_order)) +
  geom_col() +
  facet_grid(frequency_type ~ YEAR, 
             labeller = labeller(frequency_type = c("freq_mother" = "Number of Children Born", "freq_n_sib" = "Number of Siblings"), 
                        YEAR = function(x) paste("Year", x))) +
  scale_fill_manual(name = "Number of Children/Siblings",
                    values = color_mapping,
                    labels = unique(df_long_black$color.label),
                    guide = "legend") +
  scale_y_continuous(labels = comma_format(scale = 1, big.mark = ",")) +
  labs(title = "Distribution of Number of Children/Siblings by Census Year (Black/African American)",
       x = "Number of Children/Siblings",
       y = "Frequency") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "right",
        plot.title = element_text(size = 10, hjust = 0.5))
```


```{r}
df_long_white <- fert_df_white %>%
   pivot_longer(cols = c(freq_mother, freq_n_sib),
               names_to = "frequency_type",
               values_to = "frequency") %>%
  mutate(x_axis = ifelse(frequency_type == "freq_mother", chborn_num, n_sib), 
         color.label = paste(chborn_num, "_kids"," ", n_sib, "_sibs", sep = ""), 
         color_order = factor(chborn_num, levels = unique_chborn_num)
  )
```


```{r}
ggplot(df_long_white, aes(x = x_axis, y = frequency, fill = color_order)) +
  geom_col() +
  facet_grid(frequency_type ~ YEAR, 
             labeller = labeller(frequency_type = c("freq_mother" = "Number of Children Born", "freq_n_sib" = "Number of Siblings"), 
                        YEAR = function(x) paste("Year", x))) +
  scale_fill_manual(name = "Number of Children/Siblings",
                    values = color_mapping,
                    labels = unique(df_long_white$color.label),
                    guide = "legend") +
  scale_y_continuous(labels = comma_format(scale = 1, big.mark = ",")) +
  labs(title = "Distribution of Number of Children/Siblings by Census Year (White)",
       x = "Number of Children/Siblings",
       y = "Frequency") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "right",
        plot.title = element_text(size = 12, hjust = 0.5))
```



```{r}
fert_df2 <- mother_data %>% mutate(Age_bin = case_when(
    AGE >= 20 & AGE < 30 ~ "20-30",
    AGE >= 30 & AGE < 40 ~ "30-40",
    AGE >= 40 & AGE < 50 ~ "40-50",
    AGE >= 50 & AGE < 60 ~ "50-60",
    AGE >= 60 & AGE < 70 ~ "60-70",
    AGE >= 70 & AGE < 80 ~ "70-80", 
    AGE >= 80 ~ "80"
  ))

fert_df2 <- fert_df2 %>%count(RACE, chborn_num, Age_bin) %>%
  arrange(RACE, chborn_num, Age_bin) %>%
  mutate(n_sib = chborn_num -1, freq_n_sib = n * chborn_num, 
         color = color_mapping[as.character(chborn_num)], 
         ) %>% rename(freq_mother = n)

print(fert_df2)
```


```{r}
fert_df2_black <- fert_df2 %>% filter(RACE == "Black/African American")
fert_df2_white <- fert_df2 %>% filter(RACE == "White")

df_long_black2 <- fert_df2_black %>%
  pivot_longer(cols = c(freq_mother, freq_n_sib),
               names_to = "frequency_type",
               values_to = "frequency") %>%
  mutate(x_axis = ifelse(frequency_type == "freq_mother", chborn_num, n_sib), 
         color.label = paste(chborn_num, "_kids"," ", n_sib, "_sibs", sep = ""), 
         color_order = factor(chborn_num, levels = unique_chborn_num)
  )

df_long_white2 <- fert_df2_white %>%
  pivot_longer(cols = c(freq_mother, freq_n_sib),
               names_to = "frequency_type",
               values_to = "frequency") %>%
  mutate(x_axis = ifelse(frequency_type == "freq_mother", chborn_num, n_sib), 
         color.label = paste(chborn_num, "_kids"," ", n_sib, "_sibs", sep = ""), 
         color_order = factor(chborn_num, levels = unique_chborn_num)
  )
```

```{r}
ggplot(df_long_black2, aes(x = x_axis, y = frequency, fill = color_order)) +
  geom_col() +
  facet_grid(frequency_type ~ Age_bin, 
             labeller = labeller(frequency_type = c("freq_mother" = "Number of Children Born", "freq_n_sib" = "Number of Siblings"), 
                        Age_bin = function(x) paste("Age_bin", x))) +
  scale_fill_manual(name = "Number of Children/Siblings",
                    values = color_mapping,
                    labels = unique(df_long_black2$color.label),
                    guide = "legend") +
  scale_y_continuous(labels = comma_format(scale = 1, big.mark = ",")) +
  labs(title = "Distribution of Number of Children/Siblings by Age (Balck/African American)",
       x = "Number of Children/Siblings",
       y = "Frequency") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "right",
        plot.title = element_text(size = 12, hjust = 0.5))
```

```{r}
ggplot(df_long_white2, aes(x = x_axis, y = frequency, fill = color_order)) +
  geom_col() +
  facet_grid(frequency_type ~ Age_bin, 
             labeller = labeller(frequency_type = c("freq_mother" = "Number of Children Born", "freq_n_sib" = "Number of Siblings"), 
                        Age_bin = function(x) paste("Age_bin", x))) +
  scale_fill_manual(name = "Number of Children/Siblings",
                    values = color_mapping,
                    labels = unique(df_long_white2$color.label),
                    guide = "legend") +
  scale_y_continuous(labels = comma_format(scale = 1, big.mark = ",")) +
  labs(title = "Distribution of Number of Children/Siblings by Age (White)",
       x = "Number of Children/Siblings",
       y = "Frequency") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "right",
        plot.title = element_text(size = 12, hjust = 0.5))
```

