---
title: "Short Range Familial Search"
author: "Tina Lasisi"
date: "`r format(Sys.time(), '%Y-%m-%d %H:%M:%S')`"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## CODIS marker allele frequencies

Frequencies and raw genotypes for different populations were found [here](https://www.promega.com/products/pm/genetic-identity/population-statistics/allele-frequencies/) and refer to Steffen, C.R., Coble, M.D., Gettings, K.B., Vallone, P.M. (2017) Corrigendum to 'U.S. Population Data for 29 Autosomal STR Loci' [Forensic Sci. Int. Genet. 7 (2013) e82-e83]. Forensic Sci. Int. Genet. 31, e36–e40. The US core CODIS markers are a subset of the 29 described here. 

```{r setup, include=FALSE}
# Load required libraries
library(readxl)
library(tidyverse)
# Load required libraries
library(furrr)
# Set up the plan for parallel processing
library(future)

ncores = 10
```

```{r convertxl, eval=FALSE, include=FALSE}

# This chunk converts the Excel file that was downloaded into separate CSVs as there is a bunch of information that is spread across different sheets within the original Excel file.

# Read all sheets
file_path <- "data/1036_allelefreqs.xlsx"
sheet_names <- excel_sheets(file_path)

# Iterate over each sheet
for (sheet in sheet_names) {
  # Read the marker names and sample sizes
  markers <- read_excel(file_path, sheet = sheet, range = cell_rows(2:2), col_names = FALSE)
  sample_sizes <- read_excel(file_path, sheet = sheet, range = cell_rows(3:3), col_names = FALSE)

  # Read the main data (starting from the fourth row)
  data <- read_excel(file_path, sheet = sheet, skip = 3, col_names = FALSE)

  # Set the column names
  colnames(data) <- c("allele", as.character(unlist(markers)))

  # Transform the data to long format
  tidy_data <- data %>%
    pivot_longer(cols = -allele, names_to = "marker", values_to = "frequency") %>%
    rename_with(~tolower(.), everything()) # Convert all variable names to lowercase

  # Remove the rows that contain 'Hobs', 'PI', 'PPE' or 'HWE'
  tidy_data <- tidy_data[!tidy_data$allele %in% c("Hobs", "PI", "PPE", "HWE", "Allele"),] %>% 
    drop_na() %>% 
    select(marker, everything())

  # Extract population from sheet name
  tidy_data$population <- str_extract(sheet, "(?<=revised\\s)[^,]*") # Extract string between "revised " and the first comma
  
  # Clean the name for the CSV file
  filename <- gsub(" ", "", tidy_data$population[1]) # Remove spaces
  
  # Write to CSV
  write_csv(tidy_data, paste0("data/1036_", filename, ".csv")) # Write CSV to the "data" folder
}

```

## Load CODIS allele frequencies

CODIS allele frequencies were found through NIST STR base and specifically downloaded from the supplementary materials of Steffen et al 2017. These are 1036 unrelated individuals from the U.S. population. 

```{r allelefreqs}
# Define the file paths
file_paths <- list.files(path = "data", pattern = "1036_.*\\.csv", full.names = TRUE)

# Create a list of data frames
df_list <- lapply(file_paths, function(path) {
  read_csv(path, col_types = cols(
    marker = col_character(),
    allele = col_double(),
    frequency = col_double(),
    population = col_character()
  ))
})

# Bind all data frames into one
df <- bind_rows(df_list)

df_freq <- df
```



## Simulating genotypes

```{r df-ibd-probs}

df_ibdprobs <- tibble(
  relationship = factor(
    c("parent_child", "full_siblings", "half_siblings", "cousins", "second_cousins", "unrelated"),
    ordered = TRUE,
    levels = c("parent_child", "full_siblings", "half_siblings", "cousins", "second_cousins", "unrelated")
  ),
  k0 = c(0, 1/4, 1/2, 7/8, 15/16, 1),
  k1 = c(1, 1/2, 1/2, 1/8, 1/16, 0),
  k2 = c(0, 1/4, 0, 0, 0, 0)
)

df_ibdprobs

```


```{r func-get-allelefreq}

get_allelefreq <- function(allele, marker, population, df=df) {
    frequency <- df$`frequency`[which(df$`allele` == allele & df$`marker` == marker & df$`population` == population)]
    return(frequency)
  }

```


```{r func-sim-geno-parallel}

plan(multicore, workers = ncores) 


# Create the parallelized function
sim_geno_parallel <- function(relationship_type, population, num_samples = 1, marker='all', df, df_ibdprobs) {
  
  # Check if the 'marker' input is 'all'
  # If yes, assign the unique markers from 'df' data frame to 'markers'
  if (marker == 'all') {
    markers <- unique(df$marker)
  } else {
    # If the 'marker' input is not 'all', use the provided marker vector
    markers <- marker
  }
  
  # Use future_map instead of lapply to iterate through samples in parallel and create a list of tibbles
  sim_genotypes <- future_map(1:num_samples, function(i) {
    
    # Simulate the first individual's alleles for all markers using the population frequencies
    individual1 <- setNames( # Assign names to the list based on the marker names
      lapply(markers, function(current_marker) { # Loop through each marker using lapply with the variable name 'current_marker'
        
        # Filter the data frame 'df' based on the 'current_marker' and population 'population',
        # then extract the 'frequency' column as a vector and set the names of the frequency
        # vector using the filtered alleles for the current marker
        allele_frequencies <- df %>%
          filter(marker == current_marker, population == population) %>%
          pull(frequency) %>%
          setNames(filter(df, marker == current_marker, population == population)$allele)
        
        # Return a sample of two alleles from the allele names, with replacement and probabilities
        # based on the 'allele_frequencies'
        return(sample(names(allele_frequencies), size = 2, replace = TRUE, prob = allele_frequencies))
      }), markers) # End of individual1 simulation

    # Simulate the second individual's alleles based on the first individual's alleles and population frequencies
    individual2 <- setNames( # Assign names to the list based on the marker names
      lapply(markers, function(current_marker) { # Loop through each marker using lapply with the variable name 'current_marker'
        
        # Filter the data frame 'df' based on the 'current_marker' and population 'population',
        # then extract the 'frequency' column as a vector and set the names of the frequency
        # vector using the filtered alleles for the current marker
        allele_frequencies <- df %>%
          filter(marker == current_marker, population == population) %>%
          pull(frequency) %>%
          setNames(filter(df, marker == current_marker, population == population)$allele)
        
        # Filter the 'df_ibdprobs' data frame based on the provided relationship type 'relationship_type',
        # select the 'k0', 'k1', and 'k2' columns, unlist the selection, and convert the result to a numeric vector
        prob_shared_alleles <- df_ibdprobs %>%
          filter(relationship == relationship_type) %>%
          select(k0, k1, k2) %>%
          unlist() %>%
          as.numeric()

        # Calculate the indices of non-zero probabilities in 'prob_shared_alleles'
        non_zero_indices <- which(prob_shared_alleles != 0)
        # Sample the number of shared alleles based on the non-zero probabilities from 'prob_shared_alleles'
        num_shared_alleles <- sample(non_zero_indices - 1, size = 1, prob = prob_shared_alleles[non_zero_indices])
        
        # Sample alleles from 'individual1' based on 'num_shared_alleles'
        alleles_from_individual1 <- sample(individual1[[current_marker]], size = num_shared_alleles)
        # Sample alleles from the population based on the remaining alleles needed (2 - 'num_shared_alleles')
        alleles_from_population <- sample(names(allele_frequencies), size = 2 - num_shared_alleles, replace = TRUE, prob = allele_frequencies)

        # Combine and return the sampled alleles from 'individual1' and the population
        return(c(alleles_from_individual1, alleles_from_population))
      }), markers) # End of individual2 simulation

    # Create a tibble for individual 1 by combining the simulated alleles and metadata
    ind1_tibble <- tibble(
      # relationship_type = relationship_type, # Set the 'relationship_type' value to the provided 'relationship_type' argument
      # population = population, # Set the 'population' value to the provided 'population' argument
      pair_id = i, # Set the 'pair_id' value to the current iteration value (sample number)
      individual_id = 1, # Set the 'individual_id' value to 1 (first individual)
      marker = names(individual1), # Set the 'marker' value to the names of 'individual1' list (marker names)
      allele_1 = unlist(lapply(individual1, "[[", 1)), # Extract the first allele of each marker from 'individual1'
      allele_2 = unlist(lapply(individual1, "[[", 2)) # Extract the second allele of each marker from 'individual1'
    )

    # Concatenate individual 2's alleles and metadata with individual 1's tibble
    ind2_tibble <- tibble(
        # relationship_type = relationship_type, # Set the 'relationship_type' value to the provided 'relationship_type' argument
        # population = population, # Set the 'population' value to the provided 'population' argument
        pair_id = i, # Set the 'pair_id' value to the current iteration value (sample number)
        individual_id = 2, # Set the 'individual_id' value to 2 (second individual)
        marker = names(individual2), # Set the 'marker' value to the names of 'individual2' list (marker names)
        allele_1 = unlist(lapply(individual2, "[[", 1)), # Extract the first allele of each marker from 'individual2'
        allele_2 = unlist(lapply(individual2, "[[", 2)) # Extract the second allele of each marker from 'individual2'
      )
    
    # Combine individual1 and individual2 tibbles using rbind
    combined_tibble <- rbind(ind1_tibble, ind2_tibble)

    # Apply the transformation to the combined_tibble and return the result
    return(combined_tibble %>%
             pivot_longer(
               cols = c(allele_1, allele_2),
               names_to = "allele"
             ) %>%
             mutate(individual_id = paste0("ind_", individual_id)) %>%
             pivot_wider(
               names_from = c(allele, individual_id),
               values_from = value,
               names_glue = "{individual_id}_{allele}"
             ))
  }, .options = furrr_options(seed = TRUE)) # End of future_map over num_samples

  # Combine all simulated and transformed genotypes from the list 'sim_genotypes' into a single tibble and return it
  return(bind_rows(sim_genotypes))
}

```



```{r test}

# plan(multicore, workers = ncores) 

# result_parallel <- sim_geno_parallel("full_siblings", "AfAm", num_samples=10, marker="all", df=df, df_ibdprobs=df_ibdprobs)

# df_simgeno <- result_parallel


```

```{r df-parallel-combos}

# Define the number of simulations
n_sims <- 10000

# Define the list of relationship types
relationship_types <- c('parent_child', 'full_siblings', 'half_siblings', 'cousins', 'second_cousins', 'unrelated')

# Create a dataframe of all combinations of populations, relationship types, and simulations
unique_populations <- unique(df$population)
filtered_populations <- unique_populations[unique_populations != "all"]

combinations <- expand_grid(population = filtered_populations, relationship_type = relationship_types, simulation = 1:n_sims)


# Run the simulation function for each combination
df_simgeno <- combinations %>%
  mutate(sim_results = future_map2(population, relationship_type, ~sim_geno_parallel(.y, .x, num_samples = 1, marker = "all", df = df, df_ibdprobs = df_ibdprobs), .options = furrr_options(seed = TRUE))) %>%
  unnest(cols = c(sim_results))

df_simgeno


```



## Calculating LR

```{r}
get_allele_frequencies <- function(population, marker, num_shared_alleles, A, B, df_allelefreq) {
  # Initialize the output values for p_A and p_B
  p_A <- NA
  p_B <- NA

  # Check whether there are shared alleles
  if (num_shared_alleles > 0) {
    # Get the frequency of allele A in the df_allelefreq dataset
    p_A <- df_allelefreq %>%
      filter(population == population, marker == marker, allele == A) %>%
      group_by(population, marker, allele) %>%
      pull(frequency) %>%
      first()

    # If two alleles are shared and B is not NA, get the frequency of allele B
    if (num_shared_alleles == 2 && !is.na(B)) {
      p_B <- df_allelefreq %>%
        filter(population == population, marker == marker, allele == B) %>%
        group_by(population, marker, allele) %>%
        pull(frequency) %>%
        first()
    }
  }

  # Return named list of p_A and p_B
  return(list(p_A = p_A, p_B = p_B))
}


```

```{r}
# Define a function to calculate R_Xp values based on ind1_geno (c), ind2_geno (Q), p_A, and p_B values
get_R_Xp <- function(c, Q, p_A, p_B) {
  if (c == "AA") {
    if (Q == "AA") {
      return(p_A)
    } else if (Q == "AB") {
      return(2 * p_A)
    }
  } else if (c == "AB") {
    if (Q == "AA") {
      return(2 * p_A)
    } else if (Q == "AB") {
      return(4 * p_A * p_B / (p_A + p_B))
    } else if (Q == "AC") {
      return(4 * p_A)
    }
  }
  return(0)
}

# Define a function to calculate R_Xu values based on ind1_geno (c), ind2_geno (Q), p_A, and p_B values
get_R_Xu <- function(c, Q, p_A, p_B) {
  if (c == Q) {
    if (c == "AA") {
      return(p_A^2)
    } else if (c == "AB") {
      return(2 * p_A * p_B)
    }
  }
  return(0)
}
```

```{r}
# Define a function to calculate R values using the relationship, R_Xp, and R_Xu values, and the global df_ibdprobs dataframe
calc_R <- function(relationship_type, R_Xp, R_Xu, df_ibdprobs) {
  
  # Get the corresponding kappa values based on the given relationship type
    df_ibdprobs$relationship <- as.character(df_ibdprobs$relationship)
    
  k_values <- df_ibdprobs %>%
    filter(relationship == relationship_type) %>%
    select(k0, k1, k2) %>%
    as.list()

  k0 <- k_values$k0
  k1 <- k_values$k1
  k2 <- k_values$k2

  # Calculate the R value using the provided formula with correct parenthesis and handling of R_Xp and R_Xu equal to zero
  R <- k0
  if (R_Xp != 0) {
    R <- R + (k1 / R_Xp)
  }
  if (R_Xu != 0) {
    R <- R + (k2 / R_Xu)
  }

  # Calculate the log(R) value
  log_R <- log(R)

  # Return R and log(R) as a list
  return(list(R = R, log_R = log_R))
}

```



```{r func-shared-allele}

shared_alleles <- function(ind_1_allele_1, ind_1_allele_2, ind_2_allele_1, ind_2_allele_2, population, marker, simulation, relationship_type, df_allelefreq, df_ibdprobs) {

  # Create allele vectors for individuals 1 and 2
  alleles_ind1 <- c(ind_1_allele_1, ind_1_allele_2)
  alleles_ind2 <- c(ind_2_allele_1, ind_2_allele_2)

  # Initialize variables to keep track of shared alleles and used alleles
  shared <- character()
  used_ind1 <- rep(FALSE, 2)
  used_ind2 <- rep(FALSE, 2)

  # Iterate through alleles of individual 1
  for (i in 1:2) {
    allele_ind1 <- alleles_ind1[i]
    
    # Iterate through alleles of individual 2
    for (j in 1:2) {
      # Check if alleles are not used and are the same
      if (!used_ind2[j] && allele_ind1 == alleles_ind2[j]) {
        # Add the shared allele to the shared vector
        shared <- c(shared, allele_ind1)
        # Mark the alleles as used
        used_ind1[i] <- TRUE
        used_ind2[j] <- TRUE
        break
      }
    }
  }

  # Calculate the number of shared alleles and find unique shared alleles
  num_shared_alleles <- length(shared)
  unique_shared <- unique(shared)

  # Create a letter map for shared and non-shared alleles using capitalized letters
  letter_map <- setNames(LETTERS[seq_along(unique_shared)], unique_shared)
  remaining_alleles <- union(alleles_ind1, alleles_ind2)
  non_shared <- setdiff(remaining_alleles, unique_shared)
  letter_map <- c(letter_map, setNames(LETTERS[length(unique_shared) + seq_along(non_shared)], non_shared))

  # Assign capitalized letters to alleles based on the letter map and create sorted vectors
  ind1_geno <- sort(letter_map[alleles_ind1])
  ind2_geno <- sort(letter_map[alleles_ind2])

  # Initialize A and B
  A <- NA_character_
  B <- NA_character_
  
  if (num_shared_alleles > 0) {
    A <- names(letter_map)[1]
    
    if (length(unique_shared) > 1) {
      B <- names(letter_map)[2]
    }
  }

  # Sort ind1_geno and ind2_geno alphabetically and paste them together as a string vector
  ind1_geno <- paste0(sort(ind1_geno), collapse = "")
  ind2_geno <- paste0(sort(ind2_geno), collapse = "")
  
  # Return a single-row tibble directly
  result <- tibble::tibble(
    num_shared_alleles = num_shared_alleles,
    ind1_geno = ind1_geno,
    ind2_geno = ind2_geno,
    A = A,
    B = B
  )
  
    # Call the get_allele_frequencies function with required arguments
  allele_freqs <- get_allele_frequencies(population, marker, num_shared_alleles, A, B, df_allelefreq)

  # Add p_A and p_B values to the result tibble
  result$p_A <- allele_freqs$p_A
  result$p_B <- allele_freqs$p_B

  # Call the new functions to calculate R_Xp and R_Xu values and add them to the result tibble
  result$R_Xp <- get_R_Xp(result$ind1_geno, result$ind2_geno, result$p_A, result$p_B)
  result$R_Xu <- get_R_Xu(result$ind1_geno, result$ind2_geno, result$p_A, result$p_B)

  # Call the new calc_R function to calculate R and log(R) values and add them to the result tibble
  R_values <- calc_R(relationship_type, result$R_Xp, result$R_Xu, df_ibdprobs)
  result$R <- R_values$R
  result$log_R <- R_values$log_R

  return(result)
}

```




```{r}
# Wrapper function for applying shared_alleles in parallel
apply_shared_alleles_parallel <- function(input_df, df_allelefreq, df_ibdprobs = df_ibdprobs, n_workers = ncores) {
  # Select only the relevant allele columns from the dataframe, and add populationand marker columns
  all_columns <- dplyr::select(input_df, ind_1_allele_1, ind_1_allele_2, ind_2_allele_1, ind_2_allele_2, population, marker, simulation, relationship_type)

  # Set the number of workers for parallel processing
  plan(multicore, workers = n_workers)

  # Apply the shared_alleles function to the dataframe using future_map_dfr
  df_shared_alleles <- furrr::future_map_dfr(seq_len(nrow(input_df)), function(idx) {
    row_data <- all_columns[idx, ]
    shared_alleles(row_data$ind_1_allele_1, row_data$ind_1_allele_2,
                   row_data$ind_2_allele_1, row_data$ind_2_allele_2,
                   row_data$population, row_data$marker, row_data$simulation, row_data$relationship_type, df_allelefreq, df_ibdprobs)
  })

  # Combine the calculated values with the original dataframe
  df_combined <- dplyr::bind_cols(input_df, df_shared_alleles)

  return(df_combined)
}

```


```{r}

# Apply the wrapper function on the input dataframe
df_shared_alleles <- apply_shared_alleles_parallel(df_simgeno, df_freq, df_ibdprobs)

```

```{r}
df_final <- df_shared_alleles %>% 
    select(c(relationship_type, population, marker, num_shared_alleles, log_R), everything()) %>% 
  group_by(relationship_type, population, pair_id, simulation) %>%
  summarise(num_shared_alleles_sum = sum(num_shared_alleles),
            log_R_sum = sum(log_R),
            .groups = 'drop')

```

```{r}

# Function to capitalize the first letter of a string
ucfirst <- function(s) {
  paste(toupper(substring(s, 1,1)), substring(s, 2), sep = "")
}

create_plot <- function(df, variable_to_plot, relationship_col, population_col) {
  # Set the population_shape variable as factor
  df$population_shape <- factor(df[[population_col]])

  # Create the plot
  p <- ggplot(df, aes(x = .data[[relationship_col]], y = .data[[variable_to_plot]], 
                      color = .data[[population_col]], shape = .data[[population_col]], fill = .data[[population_col]])) +
    geom_boxplot(alpha = 0.4, position = position_dodge(width = 0.75)) +
    geom_point(position = position_jitterdodge(jitter.width = 0.1, dodge.width = 0.75), size = 1, alpha = 0.6) +
    theme_classic() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    scale_x_discrete(limits = c('parent_child', 'full_siblings', 'half_siblings', 'cousins', 'second_cousins', 'unrelated')) +
    scale_shape_manual(values = c(16, 17, 15, 18)) + # Change these values to desired shapes
    labs(title = paste(ucfirst(variable_to_plot), "by Relationship Type and Population"),
         x = "Relationship Type", 
         y = ucfirst(variable_to_plot), 
         color = "Population",
         shape = "Population",
         fill = "Population")
  
  # Save the plot to the /output folder with a custom file name
  save_plot <- function(plot, plot_name) {
    ggsave(filename = paste0("output/", plot_name, ".png"), plot = plot, height = 6, width = 8, units = "in")
  }
  
  # Call the save_plot function to save the plot
  save_plot(p, paste("plot_", variable_to_plot, sep = ""))
  
  return(p)
}

```

```{r fig.width=8}

# Filter your data for unique STR markers and remove the "all" population
df_plt_final <- df_final %>%
  select(-c(simulation, pair_id))

p <- create_plot(df_plt_final, "log_R_sum", "relationship_type", "population")
p

plt_allele <- create_plot(df_plt_final, "num_shared_alleles_sum", "relationship_type", "population")
plt_allele

```


