---
title: "Short Range Familial Search"
author: "Tina Lasisi"
date: "`r format(Sys.time(), '%Y-%m-%d %H:%M:%S')`"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## CODIS marker allele frequencies

Frequencies and raw genotypes for different populations were found [here](https://www.promega.com/products/pm/genetic-identity/population-statistics/allele-frequencies/) and refer to Steffen, C.R., Coble, M.D., Gettings, K.B., Vallone, P.M. (2017) Corrigendum to 'U.S. Population Data for 29 Autosomal STR Loci' [Forensic Sci. Int. Genet. 7 (2013) e82-e83]. Forensic Sci. Int. Genet. 31, e36–e40. The US core CODIS markers are a subset of the 29 described here. 

```{r setup, include=FALSE}
# Load required libraries
library(readxl)
library(tidyverse)
# Load required libraries
library(furrr)
# Set up the plan for parallel processing
library(future)
library(progressr)

ncores = 4
```

```{r convertxl, eval=FALSE, include=FALSE}

# This chunk converts the Excel file that was downloaded into separate CSVs as there is a bunch of information that is spread across different sheets within the original Excel file.

# Read all sheets
file_path <- "data/1036_allelefreqs.xlsx"
sheet_names <- excel_sheets(file_path)

# Iterate over each sheet
for (sheet in sheet_names) {
  # Read the marker names and sample sizes
  markers <- read_excel(file_path, sheet = sheet, range = cell_rows(2:2), col_names = FALSE)
  sample_sizes <- read_excel(file_path, sheet = sheet, range = cell_rows(3:3), col_names = FALSE)

  # Read the main data (starting from the fourth row)
  data <- read_excel(file_path, sheet = sheet, skip = 3, col_names = FALSE)

  # Set the column names
  colnames(data) <- c("allele", as.character(unlist(markers)))

  # Transform the data to long format
  tidy_data <- data %>%
    pivot_longer(cols = -allele, names_to = "marker", values_to = "frequency") %>%
    rename_with(~tolower(.), everything()) # Convert all variable names to lowercase

  # Remove the rows that contain 'Hobs', 'PI', 'PPE' or 'HWE'
  tidy_data <- tidy_data[!tidy_data$allele %in% c("Hobs", "PI", "PPE", "HWE", "Allele"),] %>% 
    drop_na() %>% 
    select(marker, everything())

  # Extract population from sheet name
  tidy_data$population <- str_extract(sheet, "(?<=revised\\s)[^,]*") # Extract string between "revised " and the first comma
  
  # Clean the name for the CSV file
  filename <- gsub(" ", "", tidy_data$population[1]) # Remove spaces
  
  # Write to CSV
  write_csv(tidy_data, paste0("data/1036_", filename, ".csv")) # Write CSV to the "data" folder
}

```

## Load CODIS allele frequencies

CODIS allele frequencies were found through NIST STR base and specifically downloaded from the supplementary materials of Steffen et al 2017. These are 1036 unrelated individuals from the U.S. population. 

```{r allelefreqs}
# Define the file paths
file_paths <- list.files(path = "data", pattern = "1036_.*\\.csv", full.names = TRUE)

# Create a list of data frames
df_list <- lapply(file_paths, function(path) {
  read_csv(path, col_types = cols(
    marker = col_character(),
    allele = col_double(),
    frequency = col_double(),
    population = col_character()
  ))
})

# Bind all data frames into one
df <- bind_rows(df_list)

df_freq <- df
```



## Simulating genotypes

```{r df-ibd-probs}

df_ibdprobs <- tibble(
  relationship = 
    c("parent_child", "full_siblings", "half_siblings", "cousins", "second_cousins", "unrelated"),
  k0 = c(0, 1/4, 1/2, 7/8, 15/16, 1),
  k1 = c(1, 1/2, 1/2, 1/8, 1/16, 0),
  k2 = c(0, 1/4, 0, 0, 0, 0)
)

df_ibdprobs

```



```{r v2}

simulate_STRpairs <- function(population, relationship_type, df_allelefreq, df_ibdprobs, n_sims=1) {
  
  markers <- unique(df_allelefreq$marker)
  
  allele_frequencies_by_marker <- df_allelefreq %>%
    filter(population == population) %>%
    split(.$marker)

  allele_frequencies_by_marker <- map(allele_frequencies_by_marker, ~.x %>% pull(frequency) %>% setNames(.x$allele))

  prob_shared_alleles <- df_ibdprobs %>%
    filter(relationship == relationship) %>%
    select(k0, k1, k2) %>%
    unlist() %>%
    as.numeric()
  non_zero_indices <- which(prob_shared_alleles != 0)

  k0 <- prob_shared_alleles[1]
  k1 <- prob_shared_alleles[2]
  k2 <- prob_shared_alleles[3]

  results <- lapply(markers, function(current_marker) {
    marker_results <- lapply(seq_len(n_sims), function(replicate_id) {
      individual1 <- setNames(
        lapply(markers, function(current_marker) {
          allele_frequencies <- allele_frequencies_by_marker[[current_marker]]
          return(sample(names(allele_frequencies), size = 2, replace = TRUE, prob = allele_frequencies))
        }), markers)
      
      individual2 <- setNames(
        lapply(markers, function(current_marker) {
          allele_frequencies <- allele_frequencies_by_marker[[current_marker]]
          num_shared_alleles <- sample(non_zero_indices - 1, size = 1, prob = prob_shared_alleles[non_zero_indices])
          alleles_from_individual1 <- sample(individual1[[current_marker]], size = num_shared_alleles)
          alleles_from_population <- sample(names(allele_frequencies), size = 2 - num_shared_alleles, replace = TRUE, prob = allele_frequencies)
          return(c(alleles_from_individual1, alleles_from_population))
        }), markers)

      ind1_alleles <- individual1[[current_marker]]
      ind2_alleles <- individual2[[current_marker]]
      shared_alleles <- intersect(ind1_alleles, ind2_alleles)
      num_shared_alleles <- length(shared_alleles)
      
      R_Xp <- sum(purrr::map_dbl(shared_alleles, function(x) unlist(allele_frequencies_by_marker[[current_marker]][x])))
R_Xu <- sum(purrr::map_dbl(ind1_alleles, function(x) unlist(allele_frequencies_by_marker[[current_marker]][x])) * purrr::map_dbl(ind2_alleles, function(x) unlist(allele_frequencies_by_marker[[current_marker]][x])))


      R <- k0
      if (R_Xp != 0) { R <- R + (k1 / R_Xp) }
      if (R_Xu != 0) { R <- R + (k2 / R_Xu) }
      
      log_R <- log(R)
      
       # Add the replicate_id column to the output tibble
      return(tibble(population = population,
                relationship_type = relationship_type,
                marker = current_marker,
                num_shared_alleles = num_shared_alleles,
                log_R = log_R,
                replicate_id = replicate_id)) # Add this line
})
    
    marker_results <- bind_rows(marker_results)
    return(marker_results)
  })

  result <- bind_rows(results)

  # Aggregate results for each replicate, summing num_shared_alleles and log_R values.
  result_by_replicate <- result %>%
    group_by(population, relationship_type, replicate_id) %>%
    summarise(num_shared_alleles_sum = sum(num_shared_alleles),
              log_R_sum = sum(log_R),
              .groups = "drop")
  
  return(result_by_replicate)

}


```


```{r}
# 
result <- simulate_STRpairs(population = "AfAm", relationship_type = "full_siblings", df_allelefreq = df_freq, df_ibdprobs = df_ibdprobs, n_sims=4)

result
```


## Combinations 
```{r}
simulation_combinations <- function(df, n_sims_unrelated, n_sims_related) {
  # Define the list of relationship types
  relationship_types <- c('parent_child', 'full_siblings', 'half_siblings', 'cousins', 'second_cousins', 'unrelated')
  
  # Get unique populations from the input dataframe
  unique_populations <- unique(df$population)
  filtered_populations <- unique_populations[unique_populations != "all"]
  
  # Create a dataframe of all combinations of populations and relationship types
  combinations <- expand_grid(population = filtered_populations, relationship_type = relationship_types)

  # Add the number of simulations for unrelated or related relationships
  combinations$n_sims <- ifelse(combinations$relationship_type == "unrelated", n_sims_unrelated, n_sims_related)
  
  return(combinations)
}
```


```{r}
# Example usage
result_combinations <- simulation_combinations(df, n_sims_unrelated = 10, n_sims_related = 5)

result_combinations

```


## Parallel

```{r}
# # Apply the simulate_STRpairs function to each row in result_combinations
# results <- result_combinations %>%
#   pmap_dfr(function(population, relationship_type, n_sims) {
#     simulate_STRpairs(population, relationship_type, df_allelefreq = df_freq, df_ibdprobs = df_ibdprobs, n_sims = n_sims)
#   })

```

```{r}
# Load the furrr, progressr, and future packages
library(furrr)
library(progressr)

# Set the desired plan (multisession or multicore)
plan(multisession)

# Enable the progress bar and configure its appearance
handlers(global = TRUE)
handlers("progress")

# Apply the simulate_STRpairs function to each row in result_combinations
with_progress({
  results_parallel <- result_combinations %>%
    future_pmap_dfr(function(population, relationship_type, n_sims) {
      simulate_STRpairs(population, relationship_type, df_allelefreq = df_freq, df_ibdprobs = df_ibdprobs, n_sims = n_sims)
    }, .progress = TRUE)
})

# Check the results
print(results_parallel)

```



## Figures
```{r}

# Function to capitalize the first letter of a string
ucfirst <- function(s) {
  paste(toupper(substring(s, 1,1)), substring(s, 2), sep = "")
}

create_plot <- function(df, variable_to_plot, relationship_col, population_col) {
  # Set the population_shape variable as factor
  df$population_shape <- factor(df[[population_col]])

  # Create the plot
  p <- ggplot(df, aes(x = .data[[relationship_col]], y = .data[[variable_to_plot]], 
                      color = .data[[population_col]], shape = .data[[population_col]], fill = .data[[population_col]])) +
    geom_boxplot(alpha = 0.4, position = position_dodge(width = 0.75)) +
    geom_point(position = position_jitterdodge(jitter.width = 0.1, dodge.width = 0.75), size = 1, alpha = 0.6) +
    theme_classic() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    scale_x_discrete(limits = c('parent_child', 'full_siblings', 'half_siblings', 'cousins', 'second_cousins', 'unrelated')) +
    scale_shape_manual(values = c(16, 17, 15, 18)) + # Change these values to desired shapes
    labs(title = paste(ucfirst(variable_to_plot), "by Relationship Type and Population"),
         x = "Relationship Type", 
         y = ucfirst(variable_to_plot), 
         color = "Population",
         shape = "Population",
         fill = "Population")
  
  # Save the plot to the /output folder with a custom file name
  save_plot <- function(plot, plot_name) {
    ggsave(filename = paste0("output/", plot_name, ".png"), plot = plot, height = 6, width = 8, units = "in")
  }
  
  # Call the save_plot function to save the plot
  save_plot(p, paste("plot_", variable_to_plot, sep = ""))
  
  return(p)
}

```

```{r}
simulation_results <- read.csv("data/simulation_results.csv")


```


```{r fig.width=8}


# Filter your data for unique STR markers and remove the "all" population
# df_plt_final <- results_parallel %>%
df_plt_final <- simulation_results %>%
  select(-replicate_id)

p <- create_plot(df_plt_final, "log_R_sum", "relationship_type", "population")
p

plt_allele <- create_plot(df_plt_final, "num_shared_alleles_sum", "relationship_type", "population")
plt_allele

```


